# 🚧 Under Construction:  Derivative Rules

Calculus is a powerful tool for analyzing change—but calculating derivatives from first principles using the limit definition can be time-consuming. Thankfully, mathematicians have developed **rules** that make the process faster, more intuitive, and easier to apply to real-world problems.

In this chapter, you’ll learn how to use these rules to quickly find derivatives of functions you’re likely to encounter in environmental science—whether you're modeling the growth of a forest, the spread of an invasive species, or the rate at which pollutants enter a river.

We’ll focus on developing your comfort with:

- **Recognizing patterns** in how functions are built
- **Reading mathematical expressions** like a set of instructions
- **Applying derivative rules** like shortcuts to understand change more efficiently

By the end of this chapter, you’ll be able to:

- Use **basic derivative rules** to handle powers, constants, sums, and exponential functions  
- Apply the **product**, **quotient**, and **chain rule** to more complex functions  
- Interpret what each rule tells you about **how and why a system is changing**

> *Remember: Derivatives aren't just algebra tricks—they're windows into the behavior of systems over time.*

---

## Reading Mathematical Expressions Like a Set of Instructions

Many students struggle with math not because they don’t understand the underlying concepts, but because they see mathematical expressions as mysterious symbols on a page rather than what they truly are: a sequence of instructions. Each symbol and operator is like a verb telling you what to *do* next. For example, the “+” sign isn’t just a mark—it literally means “add these numbers together.” A derivative symbol, rather than being abstract, is simply an instruction to “find how fast this quantity is changing.”  

When you begin to think of math as a set of steps or commands, it becomes less about memorization and more about following a recipe or a set of directions. This shift in perspective can transform math from something intimidating into something procedural and logical, where each step flows naturally from the last. Just as reading a musical score tells a musician which notes to play and when, reading a mathematical expression tells us which operations to perform and in what order.  



### Think of Math as a Recipe

Just like a recipe tells you what to do and in what order, a mathematical expression tells you **how to process the input** (usually a variable like \( x \)) to get an output.

Let’s take a simple example:

$$
f(x) = 3x^2 - 5x + 2
$$

This is not just a formula—it’s a **set of steps**:

1. Take the input \( x \)
2. Square it: \( x^2 \)
3. Multiply by 3: \( 3x^2 \)
4. Multiply the original \( x \) by 5: \( 5x \)
5. Subtract: \( 3x^2 - 5x \)
6. Add 2

You don’t need to memorize the formula—you just need to **follow the steps**.

---

### Operators Are Instructions

The symbols you see are **operators**:  
They **tell you what to do** with each piece of the function.

| Symbol | Instruction |
|--------|-------------|
| \( + \) | Add two quantities |
| \( - \) | Subtract one quantity from another |
| \( \times \) | Multiply |
| \( \div \) or fraction bar | Divide |
| \( ^ \) | Raise to a power |
| \( \sqrt{\cdot} \) | Take the square root |

Each operator builds on the last. When reading from left to right, **use parentheses to help clarify what happens first**.

---

### Why This Matters for Derivatives

To take a derivative, you must examine **how a function is constructed**. That’s because:

- Each term contributes to the overall rate of change
- Derivative rules apply to specific structures (e.g. products, chains, powers)

The better you can “read” an expression, the easier it becomes to:

- **Pick the right rule**
- **Avoid mistakes**
- **Explain what the math means in context**

---

::: activity
**Activity: Reading the Math — What Is the Function Telling You?**

In environmental science, mathematical models describe how systems change over time. To understand those models, we need to learn how to *read equations like a story*.

Let’s consider a function that models fish population \( P(t) \) (in thousands) over time \( t \) (in years):

$$
P(t) = 100e^{0.05t} - 20t
$$

This expression might describe fish population growth that is limited by fishing, pollution, or resource depletion.

---

**Questions to Guide Your Thinking**

Try to interpret each part of this function like a set of instructions. Answer the questions below to help unpack the meaning of the expression.

1. **Overall Structure**
   - How many terms are in this function?
   - What is being added or subtracted?

2. **First Term: \( 100e^{0.05t} \)**
   - What kind of growth does this term represent?
   - What is the role of the 100?
   - What does the exponent \( 0.05t \) suggest about the growth rate?

3. **Second Term: \( -20t \)**
   - What kind of relationship is this?
   - Why might the model include a linear decrease?
   - What could be happening in the environment to cause a consistent population loss over time?

4. **Big Picture**
   - What do you think happens to the population as time goes on?
   - Is there a time when the population might start declining instead of growing?
   - What real-world processes could this equation be modeling?

---

**Write Your Own Interpretation**

In your own words, describe what each part of the function is doing. Try completing this sentence:

> "The term \( 100e^{0.05t} \) tells me that the population is..., while the term \( -20t \) tells me that..."

---

:::

### Why This Matters

Being able to **read a function** is just as important as being able to solve it. By understanding how the structure reflects environmental processes, you'll be better prepared to:

- Apply the correct derivative rules
- Interpret changes in real systems
- Communicate what the math means to others

---

### Environmental Example

::: activity
Suppose the temperature of a lake over time is modeled by:

$$
T(t) = 18 + 2 \sin\left(\frac{\pi t}{12}\right)
$$
And we are also given the derivative function:

$$
T'(t) = \frac{\pi}{6} \cos\left(\frac{\pi t}{12}\right)
$$

Then:

- $T(t)$ tells the relationship between temperature an time
- \( T'(t) \) tells you the rate of change in temperature at any moment.
- \( \frac{dT}{dt} \) is the same thing, just written in a form that highlights that we're differentiating temperature with respect to time.

```{r lake-temp-2panel, echo=FALSE, fig.width=7, fig.height=6}
# Define functions
T <- function(t) 18 + 2 * sin(pi * t / 12)
Tprime <- function(t) (pi / 6) * cos(pi * t / 12)

# Time range (0 to 48 hours = 2 days)
t_vals <- seq(0, 48, length.out = 400)

# Compute values
temp_vals <- T(t_vals)
deriv_vals <- Tprime(t_vals)

# Set two-panel layout
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

# Plot temperature
plot(t_vals, temp_vals, type = "l", lwd = 2, col = "blue",
     ylab = "Temperature (°C)", xlab = "Time (hours)",
     main = "Lake Temperature T(t)")

# Plot derivative
plot(t_vals, deriv_vals, type = "l", lwd = 2, col = "red",
     ylab = "Rate of Change (°C/hour)", xlab = "Time (hours)",
     main = "Derivative T'(t)")
abline(h = 0, col = "gray", lty = 2)
```


Reading the math:

- Try to explain in words what T(t) looks like, break the equation up and tell the story of temperature of this lake. 
- Now try and tell the story of the derivative function.
- When is the temperature the highest? What is the rate of change at this point? Does this make sense?
- When is the rate of temperature change the highest? What is the temperature at this point? Does this make sense?

:::

---

### Reflect and Connect

::: activity
**Graphical relationship between functions and their derivatives**

Examine the plots below. 

- Can you explain the relationship between the two lines?
- What do they tell us about the function?
- Where is the derivative equal to zero? What feature do you see on the function at these points?
- The function changes concavity at some point, what do you see in the derivative at this point?

```{r, echo=FALSE}
# Example of using the operator
f <- function(x) x^3 - 5*x
f_prime <- function(x) 3*x^2 - 5
curve(f(x), from = -3, to = 3, lwd = 2,
      ylab = "f(x)", xlab = "x", main = "f(x) and its Derivative")
curve(f_prime(x), from = -3, to = 3, col = "red", lwd = 2, add = TRUE)
legend("bottomright", legend = c("f(x)", "f'(x)"), col = c("black", "red"), lwd = 2)
```

:::

---



::: activity
**Activity: Reading and Writing Derivatives**

Understanding notation is like learning to read a map: it helps you navigate and communicate mathematical ideas clearly. With practice, it becomes second nature!

Understanding derivative notation helps you interpret, communicate, and solve real-world problems. In this activity, you'll practice converting between different forms and interpreting real-world scenarios.

---

**Step 1: Interpreting a Scenario**

**Scenario:**  
The population \( P \) of a coastal bird species is changing over time. A biologist says:

> "The rate of change in population is proportional to the square of the time since a conservation effort began."

**Questions:**

1. Define the input and output variables in this problem.
2. What does “rate of change in population” suggest mathematically?
3. Write a formula that expresses this relationship using derivative notation.

*Hint: If $y$ was proportional to $x$ it would mean the $y = k.x$ where $k$ is some constant \( k \).*

---

**Discussion Prompts**

- Why might a conservationist want to know the *rate* at which a population is changing, not just its size?


By the end of this activity, you should feel more confident identifying, interpreting, and converting derivative notation—skills that will help throughout this chapter.

:::

## Introduction to Basic Derivative Rules

Now that you understand what a derivative is and how it connects to rates of change, we can start building your toolkit for taking derivatives quickly and confidently. 

The following table summarizes the most commonly used **basic derivative rules**. These rules allow us to differentiate many types of functions without going back to the limit definition each time.

Keep this table handy—it’s your go-to reference for tackling many of the problems you’ll encounter in calculus and environmental modeling.

| Rule                    | Formula                                        | Example                              |
|:-------------------------:|:-----------------------------------------------:|:--------------------------------------:|
| **Constant Rule**       | \( \frac{d}{dx}(c) = 0 \)                     | \( \frac{d}{dx}(7) = 0 \)            |
| **Power Rule**          | \( \frac{d}{dx}(x^n) = n x^{n - 1} \)         | \( \frac{d}{dx}(x^3) = 3x^2 \)       |
| **Constant Multiple**   | \( \frac{d}{dx}[c \cdot f(x)] = c f'(x) \)    | \( \frac{d}{dx}(5x^4) = 20x^3 \)     |
| **Sum Rule**            | \( \frac{d}{dx}[f(x) + g(x)] = f'(x) + g'(x) \) | \( \frac{d}{dx}(x^2 + 3x) = 2x + 3 \) |
| **Difference Rule**     | \( \frac{d}{dx}[f(x) - g(x)] = f'(x) - g'(x) \) | \( \frac{d}{dx}(x^3 - 4x^2) = 3x^2 - 8x \) |
| **Exponential Rule**    | \( \frac{d}{dx}(e^x) = e^x \)                 | \( \frac{d}{dx}(3e^x) = 3e^x \)      |
| **Logarithmic Rule**    | \( \frac{d}{dx}(\ln x) = \frac{1}{x} \)       | \( \frac{d}{dx}(\ln(2x)) = \frac{1}{x} \) |
| **Trig: Sine**          | \( \frac{d}{dx}(\sin x) = \cos x \)           | —                                    |
| **Trig: Cosine**        | \( \frac{d}{dx}(\cos x) = -\sin x \)          | —                                    |

### The Constant Rule

The **Constant Rule** tells us how to take the derivative of a constant. A constant doesn’t change, so its rate of change is zero. This makes sense because the derivative measures how quickly a function’s output changes as the input changes. If the function is always the same number, then no matter what value of \(x\) you plug in, the result doesn’t move—it’s perfectly flat.  

**Rule:**  

If  
\[
f(x) = c
\]  
where \( c \) is a constant (like 3, 5, or 100), then  
\[
f'(x) = 0
\]  

You can visualize this by graphing a constant function, such as \(f(x) = 4\). The graph is a horizontal line. Since the slope of a horizontal line is zero everywhere, the derivative is also zero everywhere.  

### Environmental Interpretation 
This rule may seem simple, but it matters in real-world modeling. For example, suppose you measure the background level of a pollutant in a lake that remains steady over time (say, \(2\) parts per million). Because this concentration is constant, its rate of change with respect to time is zero. Similarly, a flat baseline temperature in a climate dataset or a fixed conservation quota can be represented as a constant function whose derivative is always zero.  

The Constant Rule reminds us that when nothing changes, the mathematics reflects that stillness directly: no change means no slope, and thus a derivative of zero.Graphically, we can see this by the fact than a tangent line anywhere on teh funtion would be 'flat' with a slope of zero.  

```{r constant-rule-plot-fixed, echo=FALSE, fig.width=6, fig.height=4}
# Define the constant function and its derivative to return vectors
f <- function(x) rep(4, length(x))
f_prime <- function(x) rep(0, length(x))

# Create a sequence of x-values
x_vals <- seq(-5, 5, length.out = 200)

# Evaluate function and derivative
y_vals <- f(x_vals)
y_prime_vals <- f_prime(x_vals)

# Plot the constant function
plot(x_vals, y_vals, type = "l", lwd = 2,
     ylim = c(-1, 5),
     xlab = "x", ylab = "y",
     main = "Constant Function and Its Derivative")

# Add the derivative as a horizontal line at 0
lines(x_vals, y_prime_vals, lwd = 2, lty = 2)

legend("topright",
       legend = c("f(x) = 4", "f'(x) = 0"),
       lwd = 2, lty = c(1, 2))

```



The **Constant Rule** tells us how to take the derivative of a constant. A constant doesn't change, so its rate of change is zero.

**Rule:**

If  
$$
f(x) = c
$$  
where \( c \) is a constant (like 3, 5, or 100), then  
$$
f'(x) = 0
$$

#### Environmental Example

Imagine you're measuring the temperature in a perfectly climate-controlled greenhouse. If the temperature stays at 25°C all day, the rate of temperature change at any point is zero—it's constant. So the derivative is:

$$
\frac{d}{dt}(25) = 0
$$

#### Example 2

Let  
$$
f(x) = 7
$$  
Then  
$$
f'(x) = 0
$$

#### Example 3

Let  
$$
T(t) = 12
$$  
$T(t)$ represents a temperature function that never changes. The rate of change of temperature over time is:  
$$
T'(t) = 0
$$

#### Takeaway

If a function doesn’t change—like a flat line on a graph—its slope and rate of change are both zero. So anytime you're taking the derivative of a constant value, the result is always:

$$
\boxed{0}
$$

### The Power Rule

The **Power Rule** is one of the most important and widely used rules in calculus. It tells us how to differentiate expressions involving powers of \(x\). Because so many functions in environmental science (and beyond) can be expressed as polynomials, this rule becomes one of the most frequently applied tools in your calculus toolkit.  

**Rule:**  

If  
\[
f(x) = x^n
\]  
where \( n \) is any real number, then  
\[
f'(x) = n x^{n - 1}
\]  

The idea is straightforward: bring the exponent down in front as a multiplier, and then reduce the exponent by one. This simple pattern makes it easy to quickly compute derivatives of terms like \(x^2\), \(x^3\), or even fractional and negative powers such as \(x^{1/2}\) or \(x^{-3}\).  

**Examples:**  

- \( \frac{d}{dx}(x^2) = 2x \)  
- \( \frac{d}{dx}(x^5) = 5x^4 \)  
- \( \frac{d}{dx}(x^{-2}) = -2x^{-3} \)  
- \( \frac{d}{dx}(\sqrt{x}) = \frac{1}{2}x^{-1/2} \)  

**Why it matters:**  
Most polynomials, which are just sums of power functions (like \(3x^4 - 2x^2 + 7\)), can be differentiated term by term using the Power Rule. This makes them some of the easiest functions to work with in calculus.  

**Environmental interpretation:**  
The Power Rule shows up when modeling physical quantities that scale with area or volume. For example:

- The cross-sectional **area of a circular tree trunk** is proportional to \(r^2\). If the radius of the trunk is changing over time, the rate of change of area can be found using the Power Rule.  
- The **volume of a sphere** (like a balloon or water droplet) depends on \(r^3\). Differentiating with respect to time gives the rate at which volume changes as the radius grows or shrinks.  
- In **allometric scaling** (relationships between body size and metabolism), power functions often describe how one ecological trait depends on another, and their rates of change can be quickly found using this rule.  

By mastering the Power Rule, you gain a quick way to move from expressions of size or shape to expressions of *change*—a central theme in applying calculus to environmental science.  


---

### Polynomials and Their Derivatives

Here's how different powers of \( x \) behave and how their derivatives affect the shape of their graphs:

| **Power (n)** | **Function** \( f(x) = x^n \) | **Graph Type**        | **Derivative** \( f'(x) \) | **Graph of Derivative** |
|:---------------:|:-------------------------------:|:------------------------:|:-----------------------------:|:--------------------------:|
| 0             | \( x^0 = 1 \)                 | Constant               | \( 0 \)                     | Flat line at 0          |
| 1             | \( x^1 = x \)                 | Linear (straight line) | \( 1 \)                     | Flat line at 1          |
| 2             | \( x^2 \)                     | Quadratic (parabola)   | \( 2x \)                    | Linear                   |
| 3             | \( x^3 \)                     | Cubic curve            | \( 3x^2 \)                  | Parabola                 |
| 4             | \( x^4 \)                     | Quartic (W-shape)      | \( 4x^3 \)                  | Cubic curve              |

> **Note**: The derivative of a polynomial is always another polynomial of one degree lower.

---

### Environmental Interpretation

Suppose you're modeling tree growth height over time using a power function:

$$
H(t) = t^2
$$

The rate of growth at any time \( t \) is:

$$
H'(t) = 2t
$$

This means the tree grows faster as time increases—something we often observe in early growth stages of many plants.

---

### Example 1

Let  
$$
f(x) = x^5
$$  


It helps to identify the different parts of the function. So in this function $n=5$. So we could now replace 5 into the power rule derivative expression 

Then  
$$
f'(x) = n x^{n - 1} \\
f'(x) = (5) x^{5-1} \\
f'(x) = 5x^4
$$

---

### Example 2

Let  
$$
V(t) = t^3
$$  
Where V(t) represents the volume of glacial melt water over time. To find the rate of melt is we would use the power law with $n=3$  
$$
V'(t) = nt^{n-1} \\
V'(t) = n(3)t^{3-1} \\
V'(t) = 3t^2
$$

---

### Takeaway

Any time you see a variable raised to a power, apply the power rule:

1. Multiply by the original exponent  
2. Subtract 1 from the exponent

This rule works for whole numbers, fractions, and even negative exponents!

::: activity

**Derivatives of fractional powers**

Find the derivative of

$$
y = \sqrt {x}
$$
*Hint: can you write $\sqrt {x}$ as a power?*

:::

### The Constant Multiple Rule

The **Constant Multiple Rule** tells us that when a function is multiplied by a constant, we can find the derivative by simply multiplying the derivative of the function by that constant.  

This rule may feel obvious once you see it, but it’s surprisingly powerful. It tells us that constants don’t change the *shape* of the function’s derivative—they only *scale* it up or down. The slope patterns all stay the same, just stretched or compressed vertically.  

---

#### Constant Multiple Rule (Function Notation)

If  
\[
f(x) = c \cdot g(x)
\]  
where \( c \) is a constant, then  
\[
f'(x) = c \cdot g'(x)
\]  

---

#### Constant Multiple Rule (Leibniz / Operator Notation)

We can also write this using the derivative operator \( \frac{d}{dx} \):

\[
\frac{d}{dx} \left[ c \cdot g(x) \right] = c \cdot \frac{d}{dx} \left[ g(x) \right]
\]

This shows that the operator \( \frac{d}{dx} \) "passes through" the constant—it only acts on the function part \( g(x) \), not on the constant.  

> This works because constants don’t change as \(x\) changes, so they don’t contribute to the rate of change. Instead, they simply scale the rate of change of the underlying function.

---

### Environmental Interpretation

The Constant Multiple Rule is especially important in environmental modeling, where constants often represent scaling factors such as efficiency, conversion rates, or unit adjustments. These constants don’t alter *how* a system changes, only the *magnitude* of that change.  

Suppose the **CO₂ uptake** by a certain plant species is modeled by:

\[
U(t) = 0.3t^2
\]

Here we can map the function to the rule:  

- \(c = 0.3\) (a constant that might represent efficiency or an uptake rate coefficient)  
- \(g(t) = t^2\) (the underlying growth process in time)  

Applying the Constant Multiple Rule:  

\[
U'(t) = c \cdot g'(t) = (0.3)(2t) = 0.6t
\]

This result tells us that the CO₂ uptake rate increases linearly with time. The coefficient \(0.6\) is simply the constant scaling the underlying process.  

---

### Example 1

Let  
\[
f(x) = 7x^3
\]

Here, \(c = 7\) and \(g(x) = x^3\).  

Differentiate \(g(x)\) using the Power Rule:

\[
g'(x) = 3x^2
\]

Now apply the Constant Multiple Rule:

\[
f'(x) = c \cdot g'(x) = 7 \cdot (3x^2) = 21x^2
\]


```{r constant-multiple-rule-plot, echo=FALSE, fig.width=6, fig.height=4}
# Define the function and its derivative
f <- function(x) 7 * x^3
f_prime <- function(x) 21 * x^2

# Create a sequence of x-values
x_vals <- seq(-3, 3, length.out = 200)

# Evaluate function and derivative
y_vals <- f(x_vals)
y_prime_vals <- f_prime(x_vals)

# Plot the original function
plot(x_vals, y_vals, type = "l", col = "blue", lwd = 2,
     ylim = c(-60, 60),
     xlab = "x", ylab = "y",
     main = "Constant Multiple Rule Example: f(x) = 7x^3")

# Add the derivative
lines(x_vals, y_prime_vals, col = "red", lwd = 2, lty = 2)

legend("topleft",
       legend = c("f(x) = 7x^3", "f'(x) = 21x^2"),
       col = c("blue", "red"), lwd = 2, lty = c(1, 2))
```


### Key Idea

The constant just **comes along for the ride**. Differentiate the function as usual, then multiply the result by the constant.

> **Tip**: This rule is especially helpful when dealing with models where scaling factors are built into the equation. 

---


### The Sum and Difference Rules

The **Sum and Difference Rules** tell us that the derivative of a sum or difference is simply the sum or difference of the derivatives. In other words, differentiation distributes over addition and subtraction. This is powerful because it means we can take a complicated expression, break it into smaller, more manageable parts, and then handle each piece on its own before recombining the results.  

For example, suppose we want to differentiate  
\[
f(x) = 3x^2 + \sin(x) - e^x.
\]  
Instead of treating the entire function as a single, tangled object, we can separate it into three parts: \(3x^2\), \(\sin(x)\), and \(-e^x\). We find the derivative of each term individually (\(6x\), \(\cos(x)\), and \(-e^x\), respectively) and then add them back together. The process mirrors how scientists often analyze complex environmental systems—by breaking them into components that are easier to study before piecing them back together into a whole.  

In environmental contexts, this rule shows up naturally. Imagine a model of a river’s flow rate that combines multiple factors: snowmelt, rainfall, and groundwater inputs. Each of these factors may follow its own mathematical relationship with time. By using the Sum Rule, we can calculate the overall rate of change in river flow by adding the individual rates of change of snowmelt, rainfall, and groundwater. This makes the mathematics not only simpler but also more meaningful, because it aligns with how we understand real-world systems as combinations of interacting parts.  


---

#### Rule (Function Notation)

If  
$$
f(x) = g(x) \pm h(x)
$$  
then  
$$
f'(x) = g'(x) \pm h'(x)
$$

---

#### Rule (Leibniz / Operator Notation)

Using operator notation:

$$
\frac{d}{dx} \left[ g(x) \pm h(x) \right] = \frac{d}{dx} \left[ g(x) \right] \pm \frac{d}{dx} \left[ h(x) \right]
$$

> The derivative distributes over addition and subtraction. This works because differentiation is a **linear operation**.

---

### Environmental Interpretation

Suppose a wetland ecosystem has **total carbon uptake** modeled as the sum of uptake from two plant species:

$$
U_1(t) = (0.2t^2 + 0.5t) \\
U_2(t) = (t^3) \\
U_T= U_1(t)+U_2(t)=0.2t^2 + 0.5t +t^3
$$
It makes sense then that the total uptake, $U_T$, would be the uptake from both the species added together.

The **rate of total uptake** is:

$$
U_T'(t) = \frac{d}{dt}(0.2t^2 + 0.5t +t^3) 
$$

Using the sum derivative rule we could re-write the about expression as the sum of three derivatives:

$$
U_T'(t) = \frac{d}{dt}(0.2t^2) + \frac{d}{dt}(0.5t) + \frac{d}{dt}(t^3) 
$$

One it is in this form we apply the rules we let he math tell us what to do. The above expression is saying "the rate of uptake is the sum of three derivatives". Remember $\frac{d}{dt}$ is an operatator telling us to 'take the derivative of' what ever comes next.

\[
\begin{aligned}
\text{1st Term:} \quad & \frac{d}{dt}(0.2t^2) 
   = 0.2 \cdot \frac{d}{dt}(t^2) 
   = 0.2 \cdot (2t) 
   = 0.4t \\[6pt]

\text{2nd Term:} \quad & \frac{d}{dt}(0.5t) 
   = 0.5 \cdot \frac{d}{dt}(t) 
   = 0.5 \cdot (1) 
   = 0.5 \\[6pt]

\text{3rd Term:} \quad & \frac{d}{dt}(t^3) 
   = 3t^2 \\[10pt]
\end{aligned}
\]

So the derivative is:
\[
U'_T(t) = 0.4t + 0.5 + 3t^2.
\]


We applied the derivative to each term, then combined the results. For the first and second term we used the constant multiplier rule and power rules. For the third term all we needed was the power rule.

---

### Example 1

Let  
$$
f(x) = x^3 + 5x - 4
$$  
Then  
$$
f'(x) = 3x^2 + 5
$$  
(The derivative of a constant like \(-4\) is 0.)

---

### Example 2

Let  
$$
P(t) = 6t^2 - 4t + 9
$$  
Then  
$$
P'(t) = 12t - 4
$$

---

### Key Idea

Break the function into pieces using **+** and **–** signs.  
Apply known derivative rules to each term, then combine them.

> **Tip**: Combine the **Constant Multiple Rule** with the **Sum and Difference Rules** to quickly differentiate polynomials and real-world models.

---

### The Exponential Rule

One of the most remarkable functions in all of calculus is the **natural exponential function (e)**:

\[
f(x) = e^x
\]

This function is unique because its **rate of change is equal to its value** at every point:

\[
\frac{d}{dx} (e^x) = e^x
\]

This means that the tangent line at any point on the graph has a slope equal to the height of the curve at that point. No other elementary function has this “self-replicating” property.  

- If the function’s value is small, its slope is small.  
- If the function’s value is large, its slope is large.  
- Growth and steepness always keep pace with each other.  

This is why \(e^x\) is so central in calculus: it perfectly embodies the idea of *proportional change*.  

```{r exp-clean-2point2s, echo=FALSE, fig.width=7, fig.height=5}
f       <- function(x) exp(x)
fprime  <- function(x) exp(x)
tangent <- function(x, x0) f(x0) + fprime(x0) * (x - x0)

# Plot range
xlim <- c(-2, 2)
ylim <- c(0, exp(2))
x    <- seq(xlim[1], xlim[2], length.out = 600)

# Base plot
par(mar = c(4, 4.5, 3, 1))
plot(x, f(x), type = "l", lwd = 3, col = "dodgerblue3",
     xlim = xlim, ylim = ylim, xaxs = "i", yaxs = "i",
     xlab = "x", ylab = "y",
     main = "y = e^x with tangents at x = 0 and x = 1")
grid(nx = NA, ny = NA, col = "gray90", lty = 1)

# Two points & styling
x0s   <- c(0, 1)
cols  <- c("#009E73", "#7B61FF")
span  <- 0.7
labels_xy <- list(c(0.35-1, 2.0),  # label position for x0=0
                  c(1.35-1, 4.0))  # label position for x0=1

for (i in seq_along(x0s)) {
  x0 <- x0s[i]; y0 <- f(x0)
  xt <- seq(x0 - span, x0 + span, length.out = 200)

  # Tangent segment
  lines(xt, tangent(xt, x0), lwd = 2.2, col = adjustcolor(cols[i], 0.85), lty = 2)

  # Point
  points(x0, y0, pch = 21, bg = cols[i], col = "white", lwd = 2, cex = 1.3)

  # Plain-text label: f(x0)=number = f'(x0)
  lab <- sprintf("f(%g)=%.2f = f'(%g)", x0, y0, x0)

  # Connector
  segments(labels_xy[[i]][1], labels_xy[[i]][2], x0, y0,
           col = adjustcolor(cols[i], 0.85), lwd = 1)

  # Label with soft background
  rect(labels_xy[[i]][1] - strwidth("M")*0.25,
       labels_xy[[i]][2] - strheight("M")*0.7,
       labels_xy[[i]][1] + strwidth(lab) + strwidth("M")*0.25,
       labels_xy[[i]][2] + strheight("M")*0.25,
       col = rgb(1,1,1,0.9), border = NA)
  text(labels_xy[[i]][1], labels_xy[[i]][2], lab,
       adj = c(0,0.9), col = cols[i], cex = 0.95)
}

legend("topleft",
       legend = c("y = e^x", "tangent segments"),
       col = c("dodgerblue3", "black"),
       lty = c(1, 2), lwd = c(3, 2), bty = "n")
```







### Environmental Interpretation

Exponential functions appear frequently in environmental science because many natural processes change *at a rate proportional to their current size or amount*. This “proportional change” is what gives rise to exponential growth or decay.  

Some common examples include:  

- **Population growth in ideal conditions**  
  When resources are abundant and limiting factors are minimal, populations can grow exponentially. For example, bacteria in a nutrient-rich environment may double at a fixed time interval. While no population can grow exponentially forever, this model is useful for understanding the early phases of population expansion or for comparing how fast different species can reproduce.  

- **Radioactive decay**  
  Radioactive isotopes, such as carbon-14, decay at a rate proportional to the remaining quantity. This predictable exponential decay makes them powerful tools for dating environmental samples, such as using radiocarbon dating to estimate the age of fossils, soils, or archaeological artifacts.  

- **Spread of invasive species or diseases**  
  Invasive plants or pathogens can initially spread exponentially when introduced into a new habitat, before natural controls (predators, competition, or resource limits) slow their growth.  

- **Atmospheric processes**  
  Exponential decay is also used to model how pollutants or greenhouse gases diminish in the atmosphere over time through processes like chemical breakdown or absorption.  

These examples highlight why exponential functions are so central to environmental science: they provide a first-order approximation of how systems evolve when the rate of change depends directly on the current state. Later, more complex models (e.g., logistic growth or coupled differential equations) can capture limits and interactions, but exponential functions form the foundation.  


---

### Key Insight

The exponential function grows faster and faster, but always in proportion to how large it already is. This self-replicating property makes it powerful—and dangerous—in real-world systems.

::: activity

**Discussion Prompt**

Can you think of natural exponential processes that are of concern>

:::

---

### Examples

1. \( \frac{d}{dx} e^x = e^x \)

2. \( \frac{d}{dx} 5e^x = 5e^x \) ⟶ constant multiple rule still applies

3. \( \frac{d}{dx} e^{2x} = 2e^{2x} \) ⟶ chain rule! (covered next)

---

### Summary

- The derivative of \( e^x \) is **itself**.
- Exponential models describe **growth at a rate proportional to value**.
- This rule forms the backbone of many environmental models involving **runaway feedback**, like **CO₂ buildup** or **species invasion**.

### The Natural Logarithm Rule

The natural logarithm function \( \ln(x) \) is the **inverse** of the exponential function \( e^x \). While the exponential function grows at a rate proportional to its current size, the logarithm grows more and more slowly as \( x \) increases.  

This slowing growth makes the logarithm especially useful in environmental models where **diminishing returns** or **slowing growth** are observed—for example, nutrient absorption in plants, the effect of adding fertilizer to soil, or population responses as they approach a carrying capacity.  

#### Inverse Functions

An **inverse function** basically *undoes* what the original function does.  
If a function takes you from input → output, the inverse takes you from that output → back to the input.  

Think of it like this:  

- Multiplying by 2 and then dividing by 2 get you back where you started.  
- Exponentials and logarithms work the same way—\(e^x\) and \(\ln(x)\) undo each other.  

So, an inverse function is just a mathematical “reverse button.”  

---

### Key Insight

- The logarithm is **only defined for positive values of \(x\)**.  
- Its rate of change is steep when \(x\) is close to zero, but gradually flattens out as \(x\) gets larger.  
- The derivative captures this behavior:

\[
\frac{d}{dx} \ln(x) = \frac{1}{x}
\]

This formula tells us that as \(x\) grows, the slope decreases. For very large \(x\), the slope approaches zero—indicating almost no change.

---

### Examples

1. **Basic Rule**  
   \[
   \frac{d}{dx} \ln(x) = \frac{1}{x}
   \]  
   At \(x=1\), the slope is \(1\). At \(x=10\), the slope is only \(0.1\).  

2. **Log of a scaled input**  
   \[
   \frac{d}{dx} \ln(3x) = \frac{1}{3x} \cdot 3 = \frac{1}{x}
   \]  
   Notice that the factor of 3 cancels out in this case, showing that \(\ln(3x)\) grows at the same *rate* as \(\ln(x)\).  

3. **Log of a power**  
   \[
   \frac{d}{dx} \ln(x^2) = \frac{1}{x^2} \cdot 2x = \frac{2}{x}
   \]  
   This highlights how the **chain rule** applies inside logarithms: the derivative of the “inside” function modifies the result.

---

### Environmental Interpretation

The logarithmic rule is useful for modeling processes where growth slows over time or with increasing input:  

- **Resource depletion:** As resources (like groundwater or fisheries) are consumed, the additional gain from each new unit of effort declines.  
- **Nutrient absorption:** Plants absorb nutrients quickly when soil is depleted, but the uptake rate slows as the soil becomes saturated.  
- **Population growth toward carrying capacity:** While early growth may look exponential, the long-term response often involves slowing growth that logarithms help describe.  

In each case, the logarithm captures the idea that the “marginal returns” diminish as the system grows larger.  

---

### Summary

- The derivative of \( \ln(x) \) is \( \tfrac{1}{x} \).  
- The slope decreases as \(x\) increases, showing **diminishing change**.  
- This makes logarithmic models ideal for studying **slowing growth**, **saturation effects**, and **resource limits** in environmental systems.  



## Product, Quotient, and Chain Rule

So far, we've explored basic derivative rules for individual terms—constants, powers, exponentials, and logarithms. But what happens when functions are **combined**?

In real-world environmental systems, variables are rarely isolated. Instead, they interact:

- A **pollution level** might depend on **population size × emission rate**
- A **species population** might grow based on **a changing habitat and food supply**
- A **rate of nutrient absorption** might be modeled by a **composite function** like \( \ln(e^{kt}) \)

In these situations, we need a new set of rules to differentiate **combined functions**:

- The **Product Rule**: for multiplying two functions
- The **Quotient Rule**: for dividing two functions
- The **Chain Rule**: for nested (composite) functions

These rules help us break down complex expressions into manageable parts, and they are essential for modeling dynamic, interconnected systems in environmental science.

---

### Summary Table: Rules for Combined Functions

| **Rule**        | **Expression**                                     | **Derivative**                                              |
|-----------------|-----------------------------------------------------|--------------------------------------------------------------|
| Product Rule    | \( f(x) \cdot g(x) \)                               | \( f'(x) \cdot g(x) + f(x) \cdot g'(x) \)                   |
| Quotient Rule   | \( \frac{f(x)}{g(x)} \)                             | \( \frac{f'(x) \cdot g(x) - f(x) \cdot g'(x)}{[g(x)]^2} \)  |
| Chain Rule      | \( f(g(x)) \)                                       | \( f'(g(x)) \cdot g'(x) \)                                  |

---

We'll walk through each of these rules with examples and visuals that connect directly to environmental applications.


### Real-World Applications of Product, Quotient, and Chain Rules

#### Product Rule: Two Changing Quantities Multiplied

**Scenario**: Modeling **carbon emissions** from a city.

Let:
- \( P(t) \) = population size (increasing over time)
- \( E(t) \) = emissions per person

Then **total emissions** is:

\[
C(t) = P(t) \cdot E(t)
\]

To find how fast emissions are growing, we use the **product rule**:

\[
C'(t) = P'(t) \cdot E(t) + P(t) \cdot E'(t)
\]

This lets us understand whether total emissions are rising due to more people, more emissions per person, or both.

---

#### Quotient Rule: One Changing Quantity Divided by Another

**Scenario**: Modeling **pollution concentration** in a lake.

Let:
- \( A(t) \) = total pollutant amount (changing due to runoff or decay)
- \( V(t) \) = lake volume (changing due to rainfall or evaporation)

Pollution concentration is:

\[
C(t) = \frac{A(t)}{V(t)}
\]

To find how the **concentration** is changing, we use the **quotient rule**:

\[
C'(t) = \frac{A'(t)V(t) - A(t)V'(t)}{[V(t)]^2}
\]

This shows that increasing volume (dilution) can decrease concentration, even if pollutant levels stay the same.

---

#### Chain Rule: One Function Inside Another

**Scenario**: Modeling **temperature-dependent species growth**.

Let:

- \( T(t) \) = temperature over time (changes daily or seasonally)
- \( G(T) \) = growth rate as a function of temperature

Then species growth rate over time is:

\[
G(t) = G(T(t))
\]

To find the rate of change of growth over time:

\[
G'(t) = G'(T) \cdot T'(t)
\]

This lets us account for how growth indirectly changes with time due to temperature variation.


---

These examples show how derivative rules help us uncover the drivers of change in complex, real-world systems.

---

## The Product Rule

In calculus, the **Product Rule** is used to take the derivative of a product of two functions. If a quantity depends on **two changing functions multiplied together**, we can’t simply take the derivative of each part independently. We need a rule that accounts for how both parts change together.

The rule is written as:
\[
\frac{d}{dx}\big[f(x)\,g(x)\big] \;=\; f(x)\,g'(x) \;+\; f'(x)\,g(x).
\]

You may also see this written using **\(u\)** and **\(v\)** notation:
\[
\frac{d}{dx}(uv) \;=\; u\,\frac{dv}{dx} \;+\; \frac{du}{dx}\,v
\quad\text{or}\quad
\frac{d}{dx}(uv) \;=\; u\,v' + u'\,v.
\]
Here, \(u\) and \(v\) are **stand-ins** for the two functions being multiplied. They’re just labels to organize your work.

---

### Why Does the Rule Work? (Intuition)

Think of a **rectangle** whose sides change with \(x\): width \(= f(x)\), height \(= g(x)\).  
The **area** is \(A(x) = f(x)\,g(x)\).

When \(x\) increases a tiny amount, both sides change a tiny bit. The area increases due to:
1) the width changing while the height is held (momentarily) fixed, **plus**
2) the height changing while the width is held (momentarily) fixed.

Those two contributions add up:
\[
\Delta A \approx f(x)\,\Delta g \;+\; g(x)\,\Delta f
\quad\Rightarrow\quad
A'(x) = f(x)g'(x) + f'(x)g(x).
\]

> **Mnemonic:** “**first × (derivative of second) + (derivative of first) × second**.”
> Only **one** part gets differentiated in each term.


::: activity
**Quick Activity: Visualizing the Product Rule with Area**

Imagine a square whose side length depends on \(x\). At some moment, its sides have length \(f(x)\) and \(g(x)\). The *area* is:

\[
A(x) = f(x) \cdot g(x)
\]

Now suppose \(x\) increases by a tiny amount. Both sides of the square grow a little bit, which means new *thin strips of area* get added around the edges.

**Instructions:**
1. *Draw a rectangle* on your paper. Label the horizontal side \(f(x)\) and the vertical side \(g(x)\).  
2. *Increase both sides slightly*: imagine the new side lengths are \(f(x) + \Delta f\) and \(g(x) + \Delta g\).  
3. *Shade in the new area* that gets added:

   - A *horizontal strip* of area: \(f(x)\cdot \Delta g\)  
   - A *vertical strip* of area: \(g(x)\cdot \Delta f\)  
   - (Ignore the tiny corner piece \(\Delta f \cdot \Delta g\)—it’s very small compared to the strips.)  
4. *Identify each piece*: which strip corresponds to \(f(x)\,g'(x)\), and which to \(f'(x)\,g(x)\)?  
5. *Connect back*: notice that the *total new area* is just the sum of those strips, which is exactly the *Product Rule*.

**Discussion Prompt**  
Why do we get two separate terms instead of one? What would happen if we only counted one strip?  
:::

---

### Step-by-Step Example (using \(u\) and \(v\) notation)

Differentiate \( f(x) = (2x^2)(\ln x) \).

---

1. **Label parts clearly**  
   Let’s break the product into two pieces:  
   - \(u = 2x^2\) (a polynomial)  
   - \(v = \ln x\) (a logarithm)  

   > **Tip:** Identifying \(u\) and \(v\) first prevents mistakes later when plugging into the Product Rule. It’s like naming the players before describing the play.

---

2. **Differentiate each part**  
   - \(u' = \frac{d}{dx}(2x^2) = 4x\)  
   - \(v' = \frac{d}{dx}(\ln x) = \tfrac{1}{x}\)  

   > Notice how the rules we’ve already learned come together here: the **Power Rule** for \(2x^2\), and the **Logarithm Rule** for \(\ln x\).

---

3. **Apply the Product Rule**  
   Recall: \((uv)' = u\,v' + u'\,v\).  
   Substitute in each part:  

   \[
   f'(x) = u\,v' + u'\,v
   = (2x^2)\left(\frac{1}{x}\right) + (4x)(\ln x).
   \]

   > The first term represents the change when \( \ln x \) changes while \( 2x^2 \) stays fixed.  
   > The second term represents the change when \( 2x^2 \) changes while \( \ln x \) stays fixed.  
   > Together, they capture the *total* change.

---

4. **Simplify the expression**  
   \[
   f'(x) = 2x + 4x\ln x.
   \]

   > Always check if terms can be simplified. Here, \( \tfrac{2x^2}{x} = 2x \).  
   > A simpler final answer makes patterns easier to see and avoids unnecessary complexity.

---

**What we learned here:** 

- The Product Rule ensures both changing parts of the product are accounted for.  
- Each derivative rule (Power, Logarithm, Constant Multiple) can appear inside a Product Rule problem.  
- Simplification at the end often reveals a cleaner structure—in this case, notice how both terms share a factor of \(x\):  
  \[
  f'(x) = x(2 + 4\ln x).
  \]

Factoring can make the expression easier to interpret, especially if you’ll use it later in optimization or modeling.


---

### Step-by-Step Example (using \(f\) and \(g\) notation)

Differentiate  
\[
h(x) = (x^3)(e^x).
\]

---

1. **Identify each function**  
   - Let \( f(x) = x^3 \)  
   - Let \( g(x) = e^x \)  

   Then \( h(x) = f(x) \cdot g(x) \).

---

2. **Differentiate each part**  
   - \( f'(x) = \frac{d}{dx}(x^3) = 3x^2 \)  
   - \( g'(x) = \frac{d}{dx}(e^x) = e^x \)  

   > Notice how we’re reusing familiar rules: the **Power Rule** and the **Exponential Rule**.

---

3. **Apply the Product Rule**  
   Recall:  
   \[
   (fg)' = f \cdot g' + f' \cdot g
   \]

   Substitute \(f(x), f'(x), g(x), g'(x)\):  
   \[
   h'(x) = f(x)g'(x) + f'(x)g(x)
   \]
   \[
   h'(x) = (x^3)(e^x) + (3x^2)(e^x).
   \]

---

4. **Simplify the expression**  
   Factor out the common term \(e^x\):  
   \[
   h'(x) = e^x(x^3 + 3x^2).
   \]

   > Factoring here makes the structure more visible: the exponential \(e^x\) “comes along for the ride,” while the polynomial part handles the changing powers of \(x\).

---

**Takeaway:**  
When using \(f\) and \(g\) notation, the Product Rule formula is straightforward:  

- Multiply the **original \(f\)** by the **derivative of \(g\)**,  
- Add the **derivative of \(f\)** times the **original \(g\)**.  

---

### Environmental Example

Suppose total forest **biomass** \(B(t)\) is the product of:
- **Tree density** \(D(t) = 50e^{0.02t}\)  
  (number of trees per hectare, growing approximately exponentially),  
- **Average tree mass** \(M(t) = 100 + 5\sin t\)  
  (average mass of an individual tree, fluctuating slightly with seasonal cycles).  

Then the total biomass is:
\[
B(t) = D(t)\,M(t).
\]

---

#### Step-by-Step Differentiation

1. **Label each part clearly**  
   - \(u = 50e^{0.02t}\) (density of trees, growing steadily)  
   - \(v = 100 + 5\sin t\) (mass per tree, oscillating over time)

   > Splitting the model into density and mass mirrors how ecologists break down population and individual traits.

---

2. **Differentiate each part**  
   - \(u' = \frac{d}{dt}(50e^{0.02t}) = 50 \cdot 0.02e^{0.02t} = e^{0.02t}\)  
     (density is increasing at a constant 2% growth rate)  
   - \(v' = \frac{d}{dt}(100 + 5\sin t) = 5\cos t\)  
     (tree mass changes cyclically, speeding up when \(\cos t\) is large)

   > Notice how exponential growth (for density) and oscillatory change (for mass) combine in this model.

---

3. **Apply the Product Rule**  
   \[
   B'(t) = u\,v' + u'\,v
   \]
   Substituting:  
   \[
   B'(t) = 50e^{0.02t}\cdot 5\cos t \;+\; e^{0.02t}(100 + 5\sin t).
   \]

   > The first term represents biomass change driven by **seasonal shifts in mass** while density stays fixed.  
   > The second term represents biomass change driven by **density growth** while tree mass stays fixed.

---

4. **Interpretation of the Result**  

- **First term:** \(50e^{0.02t}\cdot 5\cos t\)  
  captures the effect of **seasonal mass fluctuations** across all trees. When \(\cos t > 0\), trees are *gaining* mass on average, boosting biomass. When \(\cos t < 0\), they are *losing* mass, reducing biomass.  

- **Second term:** \(e^{0.02t}(100 + 5\sin t)\)  
  captures the effect of **increasing density**. Even if individual trees aren’t changing much, adding more trees steadily increases total biomass.  

Together, these terms highlight how **two biological processes interact**: population expansion and seasonal changes in individual growth. The Product Rule ensures both are accounted for simultaneously.  

---

**Big Idea:**  
The Product Rule isn’t just an abstract formula—it matches how environmental systems behave. Complex outcomes often come from **two simpler factors interacting**, and calculus gives us a precise way to measure those interactions.

---

### When to Use (and Not Use)

- **Use Product Rule** when the expression is a **product of two changing functions** (e.g., \(x^2 e^x\), \(\ln x \cdot \sin x\), \(D(t)M(t)\)).
- **Don’t** use Product Rule for a **constant multiple**: \(7x^3\) → use the Constant Multiple Rule.
- If you see a **composition** (e.g., \(e^{2x}\), \(\ln(x^2+1)\)), that’s the **Chain Rule** (sometimes combined with a Product Rule if there’s also a product).

---

### Common Pitfalls & Quick Checks

- **Wrong pattern:** \((uv)' \neq u'v'\) — you must **add** the two terms \(u\,v' + u'\,v\).
- **Units check:** If \(u\) has units A and \(v\) has units B, then each term in \(u\,v' + u'\,v\) has units A·B per unit \(x.\) Mismatched units can reveal algebra slips.
- **Symmetry check:** Swapping \(u\) and \(v\) leaves \(u\,v' + u'\,v\) unchanged—good sanity check.

---

::: activity
**Quick Practice**

1) Differentiate \(f(x) = x^3\ln x\).  
2) Differentiate \(g(x) = (1+x^2)\,e^x\) and factor the result.  
3) Environmental: If river discharge \(Q(t) = A(t)\,v(t)\) with cross-sectional area \(A\) and velocity \(v\), write \(Q'(t)\) and explain each term in words.

:::

---

### Summary

> The **Product Rule** handles derivatives of products:  
> \[
> (fg)' = f\,g' + f'\,g.
> \]
> Intuition: change from the first factor **plus** change from the second factor.  
> Differentiate **one factor at a time**, then add.


## The Quotient Rule

In calculus, the **Quotient Rule** is used to take the derivative of a **quotient of two functions**. If a quantity depends on one function **divided by** another, we need a rule that tells us how to account for both the numerator and the denominator changing at the same time.

The rule is written as:

\[
\frac{d}{dx} \left[ \frac{f(x)}{g(x)} \right] = \frac{g(x) \cdot f'(x) - f(x) \cdot g'(x)}{[g(x)]^2}
\]

You may also see this written using **\( u \)** and **\( v \)** notation:

\[
\frac{d}{dx} \left( \frac{u}{v} \right) = \frac{v \cdot u' - u \cdot v'}{v^2}
\]

Here, \( u \) and \( v \) are **dummy variables**—stand-ins for the numerator and denominator functions. They're useful for understanding the structure of the rule, but in real problems, you’ll work with named expressions like \( f(x) \), \( C(t) \), or \( B(t) \).

---

### Why Does the Rule Work?

When dividing two functions, both the top and bottom are changing—and that affects the overall rate of change. The **Quotient Rule** carefully balances these changes:

- It starts with the derivative of the **top** times the **bottom**
- Then subtracts the **top** times the derivative of the **bottom**
- And finally divides everything by the **square of the bottom function**

---

### Step-by-Step Process (Quotient Rule)

Differentiate:
\[
f(x) = \frac{2x^2 + 3}{x + 1}.
\]

---

1. **Label each part**  
   - \(u = 2x^2 + 3\) (the numerator)  
   - \(v = x + 1\) (the denominator)  

   > Just like in the Product Rule, we split the expression into two parts, but now we have a *fraction*. It helps to keep “top = \(u\)” and “bottom = \(v\)” in mind.

---

2. **Differentiate each part**  
   - \(u' = \frac{d}{dx}(2x^2 + 3) = 4x\)  
   - \(v' = \frac{d}{dx}(x+1) = 1\)  

   > Nothing tricky here: numerator is a polynomial, denominator is linear.

---

3. **Apply the Quotient Rule**  
   Recall the formula:  
   \[
   \frac{d}{dx}\left(\frac{u}{v}\right) = \frac{v\cdot u' - u\cdot v'}{v^2}.
   \]  

   Substituting our parts:  
   \[
   f'(x) = \frac{(x+1)(4x) - (2x^2+3)(1)}{(x+1)^2}.
   \]

   > Notice the subtraction in the numerator. This is what distinguishes the Quotient Rule from the Product Rule.  
   > A useful mnemonic: **“low d-high minus high d-low, over low-low.”**  
   > (low = denominator, high = numerator)

---

4. **Simplify the expression**  
   Expand and combine terms in the numerator:  
   \[
   f'(x) = \frac{4x^2 + 4x - 2x^2 - 3}{(x+1)^2}.
   \]  

   Combine like terms:  
   \[
   f'(x) = \frac{2x^2 + 4x - 3}{(x+1)^2}.
   \]

   > Simplifying makes the structure clearer and prepares the result for later use (like plugging into a model or checking limits).  

---

### Key Idea  
The Quotient Rule accounts for the fact that **both numerator and denominator are changing**. The subtraction ensures we capture the effect of the denominator’s change correctly. Forgetting the subtraction is one of the most common student mistakes.  


### Example 2: Environmental Application

Suppose the nutrient concentration in a lake is modeled by a ratio:

- Total nutrient mass: \( N(t) = 100 + 20t \)
- Water volume: \( V(t) = 500 + 10\cos(t) \)

Then:

\[
C(t) = \frac{N(t)}{V(t)} = \frac{100 + 20t}{500 + 10\cos(t)}
\]

Let’s find the rate of change \( C'(t) \), step by step.

---

#### Step-by-Step Process

1. **Label each part:**
   - Let \( u = 100 + 20t \)
   - Let \( v = 500 + 10\cos(t) \)

2. **Differentiate each part:**
   - \( u' = 20 \)
   - \( v' = -10\sin(t) \)

3. **Apply the quotient rule:**
   \[
   C'(t) = \frac{v \cdot u' - u \cdot v'}{v^2}
   \]
   \[
   C'(t) = \frac{(500 + 10\cos(t))(20) - (100 + 20t)(-10\sin(t))}{(500 + 10\cos(t))^2}
   \]

4. **Simplify the expression:**
   \[
   C'(t) = \frac{10000 + 200\cos(t) + 1000\sin(t) + 200t\sin(t)}{(500 + 10\cos(t))^2}
   \]

This expression tells you how quickly the concentration is changing over time—factoring in both the growing nutrient load and seasonal fluctuations in water volume.

---

### Summary

> The **Quotient Rule** helps you differentiate when one function is divided by another.
>
> Remember:
> - Differentiate the **top and bottom** separately
> - Apply the rule: \( \frac{v \cdot u' - u \cdot v'}{v^2} \)
> - Keep track of signs—subtraction matters!
> - This rule tells you **exactly what to do** in a structured, step-by-step way.


## The Chain Rule

In calculus, the **Chain Rule** is used when you're taking the derivative of a **composite function**—that is, one function **nested** inside another. This happens all the time in environmental modeling, where one variable depends on another that itself is changing.

The Chain Rule is written as:

\[
\frac{d}{dx} \left[ f(g(x)) \right] = f'(g(x)) \cdot g'(x)
\]

Or in **\( u \)-notation**:

\[
\frac{dy}{dx} = \frac{dy}{du} \cdot \frac{du}{dx}
\]

Here, \( y = f(u) \) and \( u = g(x) \), so you’re taking the derivative of the **outer function**, evaluated at the inner function, and then multiplying by the derivative of the **inner function**.

---

### Composite Functions (Plain English)

Before you apply the Chain Rule, it’s helpful to get a clear sense of what a **composite function** is in everyday terms.  

- A composite function is when **one function feeds into another**.  
- Think of it like a pipeline: the output of the first function becomes the input of the second.  

Examples in words:  

- “Take a number, square it, then take the sine.” That’s \(\sin(x^2)\).  
- “Take a number, double it, then plug it into an exponential.” That’s \(e^{2x}\).  

So instead of two functions side by side (product) or one divided by another (quotient), a composite is one function **wrapped inside** another.  

**Plain English rule:** When you differentiate a composite, you have to respect the layers. Start with the **outside layer**, then peel back and multiply by the derivative of the **inside layer**. It’s like peeling an onion, one layer at a time.


### Why Does the Rule Work?

::: activity
**Chain Rule Hike**

Let’s explore the Chain Rule through a real-world analogy.
 
- Your *trail distance* itself depends on how much *time* you’ve been walking.  

So the final question is: how fast is your *distance from camp changing with time*?

**Step 2: Break down the layers**  
Let’s introduce some symbols:  

- Let \(C\) = your distance from camp (in miles).  
- Let \(s\) = your distance along the trail (in miles).  
- Let \(t\) = time (in hours).  

Try this:
- What is C a function of? Write the expression
- What is s a function of? Write the expression
- Can you combine these two functions as a *composite* function?

**Step 3: Think about the rates separately**  
Ask yourself two smaller questions:  

1. How does \(C\) change when \(s\) changes?  
2. How does \(s\) change when \(t\) changes?  

**Step 4: Combine the rates**  
Now connect them. To find how \(C\) changes with respect to \(t\):  

\[
\frac{dC}{dt} = \frac{dC}{ds} \cdot \frac{ds}{dt}.
\]

This is the *Chain Rule* in action: the overall rate is the product of the two linked rates.

**Step 5: Reflect**

- Why do we multiply instead of add?  
- What would happen if the trail were straight versus winding?  
- How does this analogy help you understand why the Chain Rule is needed when functions are “nested”?

:::


>The Chain Rule is just formalizing the idea of a “rate of a rate.” If you know **how fast one thing changes with another**, and **how that second thing changes with time**, you can multiply the two to find the total rate of change.



---

### Step-by-Step Process (Chain Rule in Action)

Differentiate:
\[
f(x) = \sqrt{3x^2 + 1}.
\]

---

1. **Identify inner and outer functions**
The function has a *nesting*:  
- The **inner function** is what’s inside the square root:  
  \[
  u = 3x^2 + 1.
  \]  
- The **outer function** is the square root itself:  
  \[
  f(u) = \sqrt{u} = u^{1/2}.
  \]

> Tip: Always peel back the layers like an onion. First ask, “What’s wrapped around the \(x\)?” Here, the outer layer is the square root, wrapped around \(3x^2 + 1\).

---

2. **Differentiate each part**
- Differentiate the outer function with respect to its input \(u\):  
  \[
  f'(u) = \frac{d}{du}(u^{1/2}) = \tfrac{1}{2}u^{-1/2}.
  \]
  
> This says: the slope of the square root gets smaller as \(u\) gets bigger.  

- Differentiate the inner function with respect to \(x\):  
  \[
  u' = \frac{d}{dx}(3x^2+1) = 6x.
  \]  

> The inner part is just a polynomial; its slope grows linearly with \(x\).

---

3. **Apply the Chain Rule**
The Chain Rule says:
\[
\frac{df}{dx} = f'(u)\cdot u'.
\]

Substitute our results:
\[
f'(x) = \Big(\tfrac{1}{2}(3x^2+1)^{-1/2}\Big)\cdot (6x).
\]

> Interpretation: take the derivative of the square root **as if it were just \(u\)**, then multiply by how fast the inside (\(3x^2+1\)) is changing.

---

4. **Simplify the expression**
- Multiply constants:
\[
f'(x) = \frac{6x}{2}(3x^2+1)^{-1/2}.
\]

- Reduce the fraction:
\[
f'(x) = 3x(3x^2+1)^{-1/2}.
\]

- Rewrite the negative exponent as a square root:
\[
f'(x) = \frac{3x}{\sqrt{3x^2+1}}.
\]

---


**Takeaway:**  
The Chain Rule is about peeling off the outer layer, differentiating it, and then multiplying by how fast the inner layer changes. Without the inner derivative, the answer will always be incomplete.

---

### Example 2: Environmental Application

Suppose the water temperature in a stream increases due to sunlight exposure over time, and the rate of fish metabolism \( M \) depends on that temperature:

- Temperature: \( T(t) = 10 + 5\sin(t) \)  
  (base temperature of 10°C plus daily/seasonal fluctuations of ±5°C modeled by a sine wave)  
- Metabolism: \( M(T) = e^{0.2T} \)  
  (exponential relationship between metabolism and temperature, as often seen in biological processes)

So the composite function is:
\[
M(t) = e^{0.2(10 + 5\sin(t))}.
\]

We want to differentiate \(M(t)\) with respect to time \(t\) to see **how quickly fish metabolism is changing over time**.

---

#### Step-by-Step Process

1. **Label inner and outer functions**
   - Inner function:  
     \[
     u = 10 + 5\sin(t)
     \]  
     This is the temperature as a function of time.  
   - Outer function:  
     \[
     M(u) = e^{0.2u}
     \]  
     This is the metabolism as a function of temperature.

> Always start by peeling back the layers: first the sine-based temperature change (inner), then the exponential response (outer).

---

2. **Differentiate each part**
   - Outer derivative:  
     \[
     \frac{dM}{du} = 0.2e^{0.2u}
     \]  
     
> This says that metabolism increases at a rate proportional to its current level—classic exponential growth.  

   - Inner derivative:  
     \[
     \frac{du}{dt} = 5\cos(t)
     \]  

> The rate of temperature change depends on the cosine curve: steepest at midpoints of the sine wave (sunrise/sunset) and flat at peaks/troughs (midday/night).

---

3. **Apply the Chain Rule**
   Combine the pieces:  
   \[
   M'(t) = \frac{dM}{du} \cdot \frac{du}{dt}
   = 0.2e^{0.2(10 + 5\sin(t))} \cdot 5\cos(t).
   \]

> This formula captures two processes at once: the exponential sensitivity of metabolism to temperature **and** the sinusoidal fluctuation of temperature with time.

---

4. **Simplify the expression**
   Multiply constants:  
   \[
   M'(t) = e^{0.2(10 + 5\sin(t))} \cdot \cos(t).\\
   M'(t) = e^{2 + \sin(t)} \cdot \cos(t).\\
   \]

> Notice how cleanly the constants combine—the factor of \(0.2\) and the \(5\) become a neat multiplier of \(1\).  
> The result shows metabolism change is driven by both the **direction of temperature change** (sign of \(\cos t\)) and the **magnitude of metabolism** (the exponential).

---

### Interpretation

- When \(\cos(t) > 0\): temperature is rising, so metabolism is increasing.  
- When \(\cos(t) < 0\): temperature is falling, so metabolism is decreasing.  
- When \(\cos(t) = 0\): temperature is at its peak or trough, so metabolism is momentarily changing at rate zero (flat slope).  

This matches ecological intuition: metabolism accelerates most when temperature is changing quickly, and stabilizes when temperature is near extremes.  

---

**Takeaway:**  
This example shows why the **Chain Rule is essential in environmental modeling**, variables rarely depend on time directly. Instead, they depend on other variables (like temperature), which themselves depend on time or other drivers. The Chain Rule is how we connect all the links in the chain.


---

### Summary

> The **Chain Rule** is used when one function is **inside another**—a composite function.  
>
> - Step 1: Identify the **outer function** and the **inner function**.  
> - Step 2: Differentiate each one.  
> - Step 3: Multiply them together: (outer derivative, evaluated at the inner) × (inner derivative).  
>
> In plain terms: **differentiate layer by layer.**  
> This happens constantly in real-world systems, where one variable depends on another that is also changing.


## Practice: Basic Derivative Rules

### Skill Drills — Build Comfort Through Repetition

When you first learned multiplication tables, you didn’t stop after doing one or two — you practiced until you could recall them without thinking.  
Derivatives work the same way. The goal here isn’t to memorize answers, but to **train your brain to recognize patterns and apply the right rule automatically**.  

The faster and more confidently you can work with the **basic derivative rules** — constant, power, constant multiple, sum/difference, exponentials, logarithms, and simple trigonometric derivatives — the more mental bandwidth you’ll have later for tackling bigger problems involving multiple rules, chain rule, or real-world applications.  

These problems are your **reps in the derivative gym**. Work through them steadily, write each step, and check your answers. Over time, your speed and accuracy will improve without you even noticing.

---

**Instructions:**  
Differentiate each function using only the **basic derivative rules** (no product, quotient, or chain rules needed).

1. \( f(x) = 7 \)  
2. \( g(x) = x^5 \)  
3. \( h(x) = -3x^{-2} \)  
4. \( p(t) = \sqrt{t} \)  
5. \( q(s) = \dfrac{1}{s^3} \)  
6. \( r(x) = 4x^7 - 5x + 2 \)  
7. \( m(y) = -\dfrac{1}{2}y^{4} + 7y^{2} \)  
8. \( n(z) = 12e^{z} \)  
9. \( k(x) = e^{-5x} \)  
10. \( P(t) = \ln t \)  
11. \( Q(r) = 5\ln r - 2r \)  
12. \( R(\theta) = \sin\theta \)  
13. \( S(\theta) = \cos\theta \)  
14. \( T(x) = -\tan x \)  
15. \( U(s) = 9s^{1/3} - 4s^{-1/2} \)  
16. \( V(t) = 6e^{3t} - 10 \)  
17. \( W(x) = 0.1x^{4} + 0.25x^{-3} \)  
18. \( X(y) = 2\sin y - 3\cos y \)  
19. \( Y(x) = 4e^{x} + \ln x \)  
20. \( Z(\theta) = 5\theta^{2} - 7\sin\theta + e^{\theta} \)  
21. \( f(x) = -11 \)  
22. \( g(x) = 15x \)  
23. \( h(t) = t^{10} \)  
24. \( p(s) = -4s^{-4} \)  
25. \( q(y) = \sqrt[3]{y} \)  
26. \( r(x) = \dfrac{7}{x^5} \)  
27. \( m(z) = z^{1/2} - z^{-1/2} \)  
28. \( n(t) = 0.5t^{8} - 3t^{3} + 2t \)  
29. \( k(x) = -\tfrac{1}{4}x^{-6} + 9x^{2} \)  
30. \( P(t) = 100e^{t} \)  
31. \( Q(r) = e^{-0.2r} \)  
32. \( R(\theta) = \ln \theta \)  
33. \( S(x) = -6\ln x + 8x \)  
34. \( T(t) = \sin t \)  
35. \( U(\theta) = -\cos\theta \)  
36. \( V(x) = 12\tan x \)  
37. \( W(s) = s^{3/4} + 4s^{-2/3} \)  
38. \( X(t) = -0.3t^{5} + 2.5t^{-4} \)  
39. \( Y(y) = 9e^{7y} - 12 \)  
40. \( Z(x) = e^{-4x} + e^{5x} \)  
41. \( A(u) = 2\sin u + 3\cos u \)  
42. \( B(\theta) = -\sin\theta - \cos\theta \)  
43. \( C(x) = 7x^{9} + \tfrac{1}{3}x^{-9} \)  
44. \( D(t) = -8e^{2t} + 5e^{-t} \)  
45. \( E(s) = \ln s - 4e^{s} \)  
46. \( F(y) = 3y^{0.5} - 4y^{2.5} \)  
47. \( G(x) = x^{7/3} - 5x^{-7/3} \)  
48. \( H(z) = 0.01z^{5} + 12z^{-5} \)  
49. \( J(t) = 2\sin t - 0.5\tan t \)  
50. \( K(x) = e^{x} + \ln x - x^{4} \)


## Practice: Product & Quotient Rules 

Up until now, we’ve only differentiated expressions built from **single terms** or **simple sums**.  
But in real problems, functions are often **multiplied** or **divided** — and the basic power, exponential, or trig rules aren’t enough.  

That’s where the **product rule** and **quotient rule** come in:  

\[
\frac{d}{dx} \big[ f(x)g(x) \big] = f'(x)g(x) + f(x)g'(x)
\]  

\[
\frac{d}{dx} \left[ \frac{f(x)}{g(x)} \right] = \frac{f'(x)g(x) - f(x)g'(x)}{\big[g(x)\big]^2}
\]  

**Tips for success:**  
1. Identify which rule applies by looking at the *top level* of the function: product or quotient.  
2. Differentiate **one part at a time**, keeping the other part unchanged.  
3. Use parentheses generously — they protect you from sign and order mistakes.  

---

**Instructions:**  
Differentiate each function. Apply the product or quotient rule as needed. Use other basic rules inside as appropriate.

1. \( f(x) = (x^3+2)(x^2-5) \)  
2. \( g(t) = t^4 e^{t} \)  
3. \( h(x) = \sin x \cdot \ln x \)  
4. \( p(z) = (3z^{-2} - 5)(z^4 + 1) \)  
5. \( q(t) = \dfrac{t^3 + 4}{t^2 - 1} \)  
6. \( r(x) = \dfrac{\ln x}{x^3} \)  
7. \( s(\theta) = \dfrac{\sin\theta}{e^{\theta}} \)  
8. \( u(y) = (y^2+1)(e^{-2y}) \)  
9. \( v(x) = \dfrac{e^{x} + x}{\sqrt{x}} \)  
10. \( w(t) = \dfrac{\cos t}{1+t^2} \)  
11. \( m(x) = (5x^2 - 4x + 1)(x^{-3}) \)  
12. \( n(t) = \dfrac{t e^{t}}{t^2+1} \)  
13. \( k(y) = (y^3+2y)\sin y \)  
14. \( P(r) = \dfrac{1 + \ln r}{r} \)  
15. \( Q(\theta) = \dfrac{\cos(2\theta)}{\theta^4} \)  
16. \( R(z) = (z^5 - z)\,e^{-z} \)  
17. \( S(x) = \dfrac{x^2+1}{e^{3x}} \)  
18. \( T(t) = (4t^{-1} + t^2)(\ln t) \)  
19. \( U(y) = \dfrac{\sin y + y^2}{\cos y} \)  
20. \( V(x) = \dfrac{x e^{x}}{x-5} \)


## Practice: Chain Rule Practice

Sometimes a function isn’t just one clean layer — it’s a function **inside** another function.  
That’s when you use the **chain rule**. It tells us:  

\[
\frac{d}{dx} f(g(x)) = f'(g(x)) \cdot g'(x)
\]  

Think of it as peeling an onion:  
1. Differentiate the **outer function** first (leave the inside unchanged).  
2. Multiply by the derivative of the **inner function**.  

**Tips for success:**  
- Always **identify the inner and outer function** before you start.  
- Parentheses are your friend — they keep track of what’s “inside.”  
- Don’t forget the inner derivative — that’s the most common mistake!  

---

**Instructions:**  
Differentiate each function using the chain rule. Apply other basic rules inside as needed.

1. \( f(x) = (3x+2)^{5} \)  
2. \( g(t) = \sqrt{5t^2 - 4} \)  
3. \( h(z) = \ln(4z^3 - z) \)  
4. \( p(y) = e^{2y - y^2} \)  
5. \( q(\theta) = \sin(7\theta - 3) \)  
6. \( r(x) = \cos(x^4 + 1) \)  
7. \( s(t) = \tan(5t) \)  
8. \( u(x) = \sqrt[3]{x^2 + 4x - 1} \)  
9. \( v(z) = e^{\sin z} \)  
10. \( w(r) = \ln\!\big(\sqrt{r^2+9}\big) \)

## Practice: Mixed Rules — Putting It All Together

So far, you’ve practiced each rule on its own:  
- **Basic rules** for powers, exponentials, logs, and trig.  
- **Product and quotient rules** for multiplication and division.  
- **Chain rule** for nested functions.  

But real functions — especially in environmental science models — rarely come neatly packaged for a single rule. You might have to use the **product rule** on the outside and the **chain rule** inside, or start with the **quotient rule** and then use a basic rule on each term.  

This section helps you get comfortable **choosing and combining** rules. The goal isn’t speed — it’s knowing how to break the problem down into manageable steps.

---

**Instructions:**  
Differentiate each function. Identify **which rules** you are using at each step.

1. \( f(x) = (x^2+1) \cdot e^{3x^2} \)  
2. \( g(t) = \dfrac{\ln(1+t^2)}{t^3} \)  
3. \( h(z) = e^{-2z} \cdot \sin(z^2) \)  
4. \( p(y) = \dfrac{e^{5y}}{\sqrt{y^2+4}} \)  
5. \( q(\theta) = (\sin\theta)(\cos(2\theta)) \)  
6. \( r(x) = \dfrac{x^3+1}{e^{x^2}} \)  
7. \( s(t) = (t^2+3t+1)^{4} \cdot \ln t \)  
8. \( u(x) = \dfrac{\sin(3x)}{\ln(x^2+1)} \)  
9. \( v(r) = (r^2-4)^{-3} \cdot e^{-r} \)  
10. \( w(z) = \dfrac{e^{\sin z} + z^2}{\sqrt{\tan z}} \)


## Practice: Applications — Derivatives in Environmental Models

The functions you’ll encounter in environmental science are often built from multiple mathematical pieces:  
- Seasonal variation (**trig**)  
- Long-term trends (**polynomials, exponentials**)  
- Constraints (**rational functions**)  
- Response curves (**logistic models**)  

These rarely fit into a single derivative rule — instead, they combine product, quotient, and chain rules.  
In each of these problems, focus on **interpreting the derivative** in terms of the situation after you compute it.

---

**Instructions:**  
For each problem:  
(a) Define variables and units clearly.  
(b) Differentiate the given function.  
(c) Interpret \( f'(x) \) in context.

---

**1. Wetland Nutrient Uptake (logistic response)**  
\[
N(t) = \frac{80}{1 + 5e^{-0.4t}}
\]
- Find \(N'(t)\).  
- Explain what the sign of \(N'(t)\) indicates about nutrient removal over time.

---

**2. Atmospheric Pollutant Decline (trend × decay)**  
\[
M(t) = (200 - 3t)\,e^{-0.05t}
\]
- Find \(M'(t)\).  
- Interpret \(M'(50)\) (tonnes/day) after 50 days.

---

**3. Reservoir Water Level (quotient with root)**  
\[
H(t) = \frac{100 + 4t}{\sqrt{t+1}}
\]
- Find \(H'(t)\).  
- Interpret \(H'(10)\) for dam release planning.

---

**4. Tree Growth with Resource Limitation (saturating growth)**  
\[
h(t) = 25\left(1 - e^{-0.2t}\right)
\]
- Find \(h'(t)\).  
- Describe how \(h'(t)\) evolves as the tree matures.

---

**5. Seasonal River Inflow (single trig component)**  
\[
Q(t) = \big(120 + 18\sin\!\tfrac{2\pi t}{12}\big)\,e^{-0.03t}
\]
- Find \(Q'(t)\).  
- Explain how the seasonal term vs. the decay term influence \(Q'(t)\) in the short run and long run.
