# Applications of Derivatives: Understanding Change and Extremes in Environmental Systems

## Using Derivatives to Find Extremes

In environmental science, many of the most important questions revolve around finding **extremes** — the highest or lowest points of a system’s behavior. These points often represent **tipping points**, **thresholds**, or **limits** that define how natural systems respond to change.

For example:

- What is the **maximum population** an ecosystem can support before resources become limiting?  
- When is the **lowest point** of atmospheric CO₂ during a seasonal cycle?  
- At what point in time does a lake’s **temperature reach its annual peak**?  
- How much rainfall leads to the **maximum runoff** in a watershed?  
- When does a species’ **growth rate slow down** as competition increases?

Each of these questions asks about **extrema** — the points where something stops increasing and starts decreasing (a **maximum**) or stops decreasing and starts increasing (a **minimum**).

---

### Why Extremes Matter

In environmental systems, extrema are rarely just mathematical curiosities. They often mark **ecological thresholds** or **critical transitions** where a system shifts from one state to another.  

- A **maximum** might signal the carrying capacity of a population or the peak of a pollution event.  
- A **minimum** could correspond to the lowest nutrient level needed to sustain life or the point of greatest drought intensity.

Understanding these points helps scientists **predict**, **manage**, and **mitigate** changes in ecosystems.


![A cartoon split into two contrasting halves under the title “Why Extremes Matter.” The left side shows a lush green landscape with a large leafy tree, a flowing river, and three white rabbits sitting on grass. A wooden sign labeled “Maximum” stands near the tree. The right side depicts a barren, cracked landscape with a dead tree, a wilted plant, and dark storm clouds overhead. A wooden sign labeled “Minimum” stands near the dead tree. The center of the image forms a curved transition between the two environments, illustrating ecological extremes. Text below explains that maxima can indicate carrying capacity or peak pollution, while minima can represent nutrient scarcity or drought intensity, emphasizing the importance of understanding these points in ecosystems](images/extrema.png)

---

### The Mathematical Lens

From a mathematical perspective, these questions are all about the **shape of a function** — how it rises and falls. When a function reaches a peak or valley, its graph changes direction. At that turning point, the slope of the tangent line becomes **zero**.

That’s where **derivatives** come in.

A **derivative** measures how fast something is changing — the *slope* of the function at any given point.  

- When the derivative is **positive**, the function is increasing.  
- When it is **negative**, the function is decreasing.  
- When it equals **zero**, the function is neither rising nor falling — it’s *level* — and that’s where we look for potential extrema.

---

### The Big Idea

Derivatives act as a **mathematical microscope**: they let us zoom in on how a function behaves locally — whether it’s climbing, falling, or flattening out.  

By finding where the derivative equals zero (or is undefined), we identify **critical points** — the likely spots for maxima or minima.

From there, we use the **sign of the derivative** (and sometimes the second derivative) to determine whether each point is a **peak**, **valley**, or **inflection** in the system.

---

### Environmental Interpretation

In practice, this process might mean:

- Identifying the **time** when a glacier’s melt rate switches from accelerating to slowing.  
- Finding the **nutrient concentration** that maximizes algal growth before it declines due to self-shading.  
- Determining the **temperature** where enzyme activity in a soil process is optimal.  

In all these cases, calculus provides the same insight: the moment where **change stops changing direction** — a snapshot of balance in a dynamic world.

---

## Understanding Maximums and Minimums

Before we dive deeper into finding extrema, let’s clarify what we actually mean by **maximums** and **minimums** — and the difference between **local** and **global** extremes.

---

### What Is a Maximum?

A **maximum** is a point where the function reaches its **highest value** — either within a small region or across its entire domain.

- **Local (Relative) Maximum:**  
  A point where the function is higher than *nearby* points.  
  Think of it as a small hilltop on a larger landscape.  
  Mathematically, \( f(a) \) is a local maximum if:
  \[
  f(a) \ge f(x) \quad \text{for all } x \text{ near } a
  \]

- **Global (Absolute) Maximum:**  
  The single **highest point** on the entire graph.  
  Mathematically, \( f(a) \) is a global maximum if:
  \[
  f(a) \ge f(x) \quad \text{for every } x \text{ in the domain}
  \]

> **Example:**  
> In a year-long temperature record, the hottest day of the year is the **global maximum**,  
> while the warmest day in each month would be **local maxima**.

---

### What Is a Minimum?

A **minimum** is a point where the function reaches its **lowest value**.

- **Local (Relative) Minimum:**  
  A point where the function is lower than nearby points — a small valley in the landscape.  
  \[
  f(a) \le f(x) \quad \text{for all } x \text{ near } a
  \]

- **Global (Absolute) Minimum:**  
  The **lowest point** across the entire function.  
  \[
  f(a) \le f(x) \quad \text{for every } x \text{ in the domain}
  \]

> **Example:**  
> For a river’s flow rate, the **lowest discharge of the year** is a global minimum,  
> while each dip between storms is a local minimum.

---

### Global Extrema and Finite Domains

When a function is defined only over a **finite interval** — for example, \( a \le x \le b \) —  
the **global maximum or minimum** can occur **at the endpoints** as well as at interior critical points.  

That means you must **always check the function’s boundaries** when finding global extrema.  
Sometimes, the highest or lowest value of a system happens right at the edge of the range you’re studying.

> **Example:**  
> If we model daily solar radiation between sunrise and sunset,  
> the **minimum** radiation occurs **at sunrise and sunset** — the endpoints of the domain.

---

### Visualizing Local vs. Global Extrema

Imagine hiking through a mountain range:

- Each **hilltop** you reach is a **local maximum**.  
- The **tallest mountain** of all is the **global maximum**.  
- Each **valley floor** between peaks is a **local minimum**.  
- The **deepest valley** overall is the **global minimum**.  
- If your hike starts and ends in a valley, those **endpoints** might also mark global extrema — even if no peak lies between.

```{r four-extrema-types, echo=FALSE, fig.width=10, fig.height=8}
library(ggplot2)
library(patchwork)

# ==========================================================
# Function 1: Cubic with local max & min
# ==========================================================
f1 <- function(x) x^3 - 3*x^2 + 2
x1 <- seq(-1, 3.5, length.out = 400)
df1 <- data.frame(x = x1, y = f1(x1))

# Critical points
crit1 <- data.frame(x = c(0, 2), y = f1(c(0, 2)),
                    type = c("Local Max", "Local Min"))

p1 <- ggplot(df1, aes(x, y)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(data = crit1, aes(x, y, color = type), size = 3) +
  geom_text(data = crit1, aes(label = type, y = y + 1.2), color = "black") +
  labs(title = "A. Local Max and Min (Cubic Function)",
       subtitle = expression(f(x) == x^3 - 3*x^2 + 2),
       x = "x", y = "f(x)") +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = c("Local Max" = "red", "Local Min" = "blue"))

# ==========================================================
# Function 2: Quadratic with global extrema at endpoints
# ==========================================================
f2 <- function(x) -(x - 3)^2 + 5
x2 <- seq(0, 6, length.out = 400)
df2 <- data.frame(x = x2, y = f2(x2))

# Endpoints & vertex
crit2 <- data.frame(x = c(3, 0, 6),
                    y = f2(c(3, 0, 6)),
                    type = c("Global Max", "Endpoint Min", "Endpoint Min"))

p2 <- ggplot(df2, aes(x, y)) +
  geom_line(color = "darkgreen", linewidth = 1.2) +
  geom_point(data = crit2, aes(x, y, color = type), size = 3) +
  geom_text(data = crit2, aes(label = type, y = y + 0.6), color = "black") +
  labs(title = "B. Global Extrema at Endpoints (Finite Domain)",
       subtitle = expression(f(x) == -~(x-3)^2 + 5~~" over [0,6]"),
       x = "x", y = "f(x)") +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = c("Global Max" = "red", "Endpoint Min" = "darkgreen"))

# ==========================================================
# Function 3: Flat region (derivative = 0 but not extrema)
# ==========================================================
f3 <- function(x) x^3
x3 <- seq(-2, 2, length.out = 400)
df3 <- data.frame(x = x3, y = f3(x3))

crit3 <- data.frame(x = 0, y = f3(0), type = "Flat (No Extremum)")

p3 <- ggplot(df3, aes(x, y)) +
  geom_line(color = "orange", linewidth = 1.2) +
  geom_point(data = crit3, aes(x, y), color = "black", size = 3) +
  geom_text(data = crit3, aes(label = type, y = y + 1), color = "black") +
  labs(title = "C. Flat Region —  No Extrema",
       subtitle = expression(f(x) == x^3),
       x = "x", y = "f(x)") +
  theme_minimal(base_size = 12)

# ==========================================================
# Function 4: Periodic (repeating local maxima and minima)
# ==========================================================
f4 <- function(x) sin(x)
x4 <- seq(0, 2*pi, length.out = 400)
df4 <- data.frame(x = x4, y = f4(x4))

crit4 <- data.frame(x = c(pi/2, 3*pi/2),
                    y = f4(c(pi/2, 3*pi/2)),
                    type = c("Local Max", "Local Min"))

p4 <- ggplot(df4, aes(x, y)) +
  geom_line(color = "purple", linewidth = 1.2) +
  geom_point(data = crit4, aes(x, y, color = type), size = 3) +
  geom_text(data = crit4, aes(label = type, y = y + 0.3), color = "black") +
  labs(title = "D. Periodic Function with Repeating Extrema",
       subtitle = expression(f(x) == sin(x)~~" over [0, 2π]"),
       x = "x", y = "f(x)") +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = c("Local Max" = "red", "Local Min" = "blue"))

# ==========================================================
# Combine all four panels into a 2x2 grid
# ==========================================================
(p1 | p2) / (p3 | p4)

```

### What each panel illustrates:

| Panel | Function | Concept Highlighted |
|--------|-----------|---------------------|
| **A** | \( f(x) = x^3 - 3x^2 + 2 \) | Typical local **max and min** inside the domain |
| **B** | \( f(x) = -(x - 3)^2 + 5 \) over \([0,6]\) | **Global extrema at endpoints** |
| **C** | \( f(x) = x^3 \) | **Flat point** with \( f'(x)=0 \) but no extremum |
| **D** | \( f(x) = \sin(x) \) | **Repeating local maxima/minima** in periodic behavior |


---

### Why the Difference Matters

When analyzing environmental data — like temperature, population, or rainfall —  
it’s often the **local extrema** that show **short-term cycles** or **seasonal variation**,  
while **global extrema** reveal **long-term limits** or **absolute thresholds**.

For instance:

- The **local maximum** in CO₂ each year reflects seasonal plant activity.  
- The **global maximum** shows the all-time record high in atmospheric concentration.

---

> **In summary:**

> - **Local extrema** describe *regional or temporary* highs and lows.  
> - **Global extrema** describe the *overall* highest or lowest points in a dataset or model.  
> - On a **finite domain**, global extrema can occur **at the endpoints** as well as where \( f'(x) = 0 \).  
>  
> Both are found using the same calculus tools — derivatives — but interpreted in different ways depending on the system’s context.


## The Idea

Every smooth curve tells a story about change. When a function reaches a **peak** (a high point) or a **valley** (a low point), its direction of change reverses — from increasing to decreasing, or vice versa. These turning points mark moments where something important happens: the system pauses before shifting course.

Mathematically, this change in direction shows up in the **slope** of the function.  
The **derivative**, \( f'(x) \), measures that slope — how fast the function is increasing or decreasing at any given point.

- When \( f'(x) > 0 \), the slope is **positive** → the function is **rising**.  
- When \( f'(x) < 0 \), the slope is **negative** → the function is **falling**.  
- When \( f'(x) = 0 \), the slope is **flat** → the function is **level**.

At a **peak or valley**, the tangent line to the curve flattens out — it’s neither rising nor falling.  
That’s why these special points occur when:

\[
f'(x) = 0
\]

These are called **critical points**, and they are the *candidates* for extreme values — places where a maximum or minimum *might* occur.

---

### Visualizing the Idea

Imagine hiking through a mountain landscape:

- When you’re **climbing**, the slope beneath your feet is positive — you’re going up.  
- When you’re **descending**, the slope is negative — you’re going down.  
- At the **summit**, for just a moment, the ground feels level — your slope is zero.  
  That’s the mathematical equivalent of \( f'(x) = 0 \).  

Similarly, in environmental systems:

- The point when a **river’s discharge** stops rising after a storm marks a **maximum**.  
- The moment when **CO₂ concentration** stops falling in spring marks a **minimum**.  

These transitions — from increasing to decreasing or vice versa — are what we call **extrema**.

---

### Why “Critical Points” Aren’t Always Extremes

Finding where \( f'(x) = 0 \) is an important step, but it doesn’t automatically mean the function has a **maximum** or **minimum** there.  

A zero slope simply means the tangent line is flat — the function could be turning around **or** continuing in the same direction.

---

#### When Flat Doesn’t Mean Peak or Valley

Sometimes the derivative equals zero, but the function just keeps going up or down.  
Consider the function:

\[
f(x) = x^3
\]

Its derivative is:

\[
f'(x) = 3x^2
\]

Here, \( f'(0) = 0 \), so \( x = 0 \) is a **critical point**.  
But if you graph \( f(x) = x^3 \), you’ll see that the curve flattens briefly at the origin, then continues increasing.  
There’s **no maximum or minimum** — the function simply changes **concavity** (the direction it bends).


```{r x3-derivative-side-by-side, echo=FALSE, fig.width=8, fig.height=4}
library(ggplot2)
library(patchwork)  # for side-by-side plots

# Define the function and its derivative
f <- function(x) x^3
f_prime <- function(x) 3 * x^2

# Create a sequence of x values
x_vals <- seq(-3, 3, length.out = 200)
data <- data.frame(
  x = x_vals,
  fx = f(x_vals),
  fpx = f_prime(x_vals)
)

# Plot f(x)
p1 <- ggplot(data, aes(x, fx)) +
  geom_line(color = "forestgreen", linewidth = 1.2) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_vline(xintercept = 0, color = "gray60", linetype = "dashed") +
  labs(title = expression(f(x) == x^3),
       x = "x",
       y = "f(x)") +
  theme_minimal(base_size = 13)

# Plot f'(x)
p2 <- ggplot(data, aes(x, fpx)) +
  geom_line(color = "darkorange", linewidth = 1.2) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_vline(xintercept = 0, color = "gray60", linetype = "dashed") +
  labs(title = expression(f*minute(x) == 3*x^2),
       x = "x",
       y = "f'(x)") +
  theme_minimal(base_size = 13)

# Combine the two plots side by side
p1 + p2
```

---

#### What This Means

Setting \( f'(x) = 0 \) identifies **potential extrema**, not guaranteed ones.  
To determine what kind of point it is, we need more information about the function’s behavior near that location.

We can use:

- A **first derivative sign chart** to see if the slope changes from positive → negative (maximum) or negative → positive (minimum).  
- Or the **second derivative test**, which checks whether the curve bends upward or downward at that point.

---

#### Key Takeaways

- A **critical point** is where \( f'(x) = 0 \) or \( f'(x) \) is undefined.  
- A **true extremum** occurs only if the function **changes direction** there.  
- Use derivative **sign changes** or **concavity** to tell the difference.

> In short:  
> Solving \( f'(x) = 0 \) gives us the *candidates* for maxima and minima —  
> further analysis confirms which ones truly are.


### Environmental Example

In population ecology, imagine a species whose population \( P(t) \) grows rapidly at first but slows down as resources become limited.  

If we graph \( P(t) \) over time, the population might climb to a **maximum** before leveling off or declining.  
At that turning point, the **rate of growth** — the derivative \( P'(t) \) — equals **zero**.  

Finding that zero slope helps us identify the moment of **ecological balance**, where birth rates equal death rates and the system momentarily stabilizes.

In short, **extrema mark the balance points of change**, and the derivative is our tool for detecting them.  
By finding where the slope becomes zero, we uncover where a system pauses, pivots, or peaks —  
the fingerprints of dynamic processes across environmental systems.


## Step-by-Step: Finding Extrema

When we want to find where a function reaches its **highest** or **lowest** point, we follow a logical, repeatable process.

You can think of this as a “recipe” for finding turning points — it always works, no matter how complicated the function.

---

### The Four-Step Process

1. **Take the derivative** of the function.  
   → This gives you a formula for the **slope** (how fast the function is changing).

2. **Find critical points** by solving \( f'(x) = 0 \) (or where \( f'(x) \) is undefined).  
   → These are *possible* peaks or valleys.

3. **Use a sign chart or second derivative** to classify each critical point.  
   → This tells you whether each point is a **maximum**, **minimum**, or **neither**.

4. **Evaluate the function** at those points.  
   → This gives the actual *values* of the extremes (the “how high” or “how low”).

---

## Step 1: Take the Derivative

To begin, take the **derivative** of your function \( f(x) \).  
This tells you the **rate of change** — how fast \( f(x) \) is rising or falling at each point.

If you imagine the graph of your function as a landscape:

- The **slope** of the function tells you whether you’re climbing **uphill** (positive slope) or going **downhill** (negative slope).
- At a **peak or valley**, the slope **flattens out** — the hill levels before changing direction.
  
That’s why extrema occur when the derivative equals **zero**.

---

### How to Differentiate

Apply the **derivative rules** you already know:

| Rule Type | Example | Derivative |
|------------|----------|------------|
| Power Rule | \( x^n \) | \( n x^{n-1} \) |
| Sum Rule | \( f(x) + g(x) \) | \( f'(x) + g'(x) \) |
| Product Rule | \( f(x)g(x) \) | \( f'(x)g(x) + f(x)g'(x) \) |
| Quotient Rule | \( \frac{f(x)}{g(x)} \) | \( \frac{f'(x)g(x) - f(x)g'(x)}{g(x)^2} \) |
| Chain Rule | \( f(g(x)) \) | \( f'(g(x)) \cdot g'(x) \) |

> **Tip:** Always rewrite the function cleanly before differentiating — simplify fractions, expand products, and identify which rule(s) you’ll need.

---

### What You’re Looking For

You’re not differentiating just to practice — you’re looking for **turning points**, where the slope might change sign.  
These are the potential **peaks** or **valleys** of the system, the places where growth stops and reversal begins.

Once you have \( f'(x) \), the next task is to find where it equals **zero**.

---

## Step 2: Find the Critical Points

A **critical point** is any point on the function where:

1. \( f'(x) = 0 \), or  
2. \( f'(x) \) does not exist (but \( f(x) \) does).

These are the places where the function might reach a **maximum** or **minimum**.

---

### Why Set the Derivative to Zero?

At a peak or valley, the tangent line to the curve is **flat** — no uphill or downhill slope — meaning:
\[
f'(x) = 0
\]

That’s the condition we solve to find potential extrema.

---

### How to Solve \( f'(x) = 0 \)

Depending on your derivative, different algebraic tools work best:

- **Factoring:** The fastest approach if possible.  
  Example: \( f'(x) = 3x(x - 2) = 0 \Rightarrow x = 0, 2 \)
- **Quadratic Formula:**  
  Use if \( f'(x) \) is quadratic and doesn’t factor nicely:  
  \[
  x = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}
  \]
- **Graph or numerical methods:**  
  If the equation is more complicated (exponential, trigonometric, higher-order), use a graphing tool like **Desmos**, **R**, or a calculator to approximate roots.

> **Note:** Sometimes \( f'(x) \) might be undefined (like with absolute value functions).  
> Those points can also be critical points if \( f(x) \) itself is still defined there.

---

### Quick Example

If \( f(x) = x^3 - 6x^2 + 9x + 1 \), then  
\( f'(x) = 3x^2 - 12x + 9 \).  
Setting \( f'(x) = 0 \) gives \( 3(x^2 - 4x + 3) = 0 \)
Factoring give us \( (x - 1)(x - 3) = 0 \).  
So \( x = 1, 3 \) are our **critical points**.

Now we need to determine what kind of points they are.

---

## Step 3: Classify Each Critical Point Using a Sign Chart

Once you’ve found the critical points, the next question is:  

> “Is this point a peak, a valley, or just a flat spot?”

To answer that, we use a **sign chart** — a simple, visual way to test how the slope behaves around each critical point.

---

### How to Build a First Derivative Sign Chart

1. **Factor the derivative** (if possible).  
   This makes it easy to check the signs of each factor.
2. **Mark the critical points** on a number line.  
   They divide the number line into intervals.
3. **Choose a test value** in each interval.  
   Substitute it into \( f'(x) \) to check the sign (positive or negative).
4. **Record the results**.  
   - If \( f'(x) \) changes from **+ → –**, it’s a **local maximum**.  
   - If \( f'(x) \) changes from **– → +**, it’s a **local minimum**.  
   - If there’s **no sign change**, it’s not an extreme point (could be an inflection).

---

#### Example Table

| Interval | Test Value | Sign of \( f'(x) \) | Function Behavior | Result |
|:---------:|:-----------:|:------------------:|:-----------------:|:------:|
| Left of \( x=1 \) | 0 | \( + \) | Increasing |  |
| Between \( 1 \) and \( 3 \) | 2 | \( - \) | Decreasing | Max at \( x=1 \) |
| Right of \( x=3 \) | 4 | \( + \) | Increasing | Min at \( x=3 \) |

---

### Visualizing the Sign Chart

You can think of the sign chart as a slope “map”:

- \( + \) means the function rises (↗️)
- \( - \) means it falls (↘️)
- A **change in direction** marks a turning point

This method is powerful because it shows *why* maxima and minima occur.

---

## Step 4: Evaluate the Function at the Critical Points

Now that you know which points are maxima or minima, you can find their **actual heights or depths** by plugging those \( x \)-values back into the **original function** \( f(x) \).

For each critical point \( x_i \):
\[
y_i = f(x_i)
\]

These \( (x_i, y_i) \) pairs are your **extrema coordinates**.

---

### How to Interpret the Results

| Type | Behavior | Example Interpretation |
|------|------------|------------------------|
| Local Maximum | Function rises then falls | Peak temperature or maximum biomass |
| Local Minimum | Function falls then rises | Lowest CO₂ concentration or drought minimum |
| Neither | Function flattens but continues | Inflection in trend or temporary slowdown |

> **Tip:** Always check whether your extrema are *local* (within a neighborhood) or *global* (the absolute highest or lowest value on the domain).

---

### Putting It All Together

By following these steps, you can analyze any smooth function systematically:

| Step | What You Do | What It Tells You |
|:----:|:------------:|:----------------:|
| 1. Differentiate | Find \( f'(x) \) | The rate of change |
| 2. Set \( f'(x) = 0 \) | Solve for \( x \) | Potential turning points |
| 3. Sign chart or \( f''(x) \) test | Check slope behavior | Identify max/min |
| 4. Evaluate \( f(x) \) | Get actual values | How high or low the extremes are |

---

### Environmental Perspective

In environmental modeling, this process might look like:

- Finding when a **river flow rate** peaks after rainfall (maximum).
- Determining the **lowest temperature** in a yearly cycle (minimum).
- Locating when a **population growth curve** levels off at carrying capacity (slope = 0).

In all cases, derivatives let us move from *observation* (“the graph changes”) to *quantification* (“it changes **where** \( f'(x) = 0 \) and **how much** it changes around that point”).

---




## 🌱 Environmental Example: A Cubic Population Model

Let’s explore how derivatives help us find **maximum** and **minimum** points using a realistic environmental example — a population that grows, stabilizes, and then declines due to changing conditions.

---

### Step 0: The Model

We’ll use this **cubic population function**:

\[
P(t) = -t^3 + 6t^2 - 9t + 20
\]

Here:

- \( P(t) \) represents population size at time \( t \)
- The negative cubic term (\(-t^3\)) eventually causes decline — simulating limits like **competition, disease, or limited resources**
- The middle terms model an initial **growth phase** before the downturn

```{r, echo=FALSE}
P <- function(t) -t^3 + 6*t^2 - 9*t + 20
curve(P(x), from = 0, to = 5, 
      xlab = "Time (t)", 
      ylab = "Population P(t)", 
      main = expression(P(t) == -t^3 + 6*t^2 - 9*t + 20),
      col = "forestgreen", lwd = 2)
grid()
```

---

### Step 1: Find the First Derivative

To determine when the population reaches a **high point (maximum)** or **low point (minimum)**, we start by finding how the population is **changing over time** — its **rate of change**.  
This rate of change is given by the **first derivative** of the function \( P(t) \).

Our model is:

\[
P(t) = -t^3 + 6t^2 - 9t + 20
\]

Apply the **power rule** to each term:

| Term | Derivative |
|:----:|:-----------:|
| \( -t^3 \) | \( -3t^2 \) |
| \( 6t^2 \) | \( 12t \) |
| \( -9t \) | \( -9 \) |
| \( 20 \) | \( 0 \) (constants don’t change) |

Combine these results:

\[
P'(t) = -3t^2 + 12t - 9
\]

---

### What the Derivative Tells Us

- \( P'(t) > 0 \): the population is **increasing** (growth phase)  
- \( P'(t) < 0 \): the population is **decreasing** (decline phase)  
- \( P'(t) = 0 \): the growth rate is **zero** — the population is **neither rising nor falling**  

These “flat slope” points are where the population might reach a **maximum** or **minimum** — the key turning points in its behavior.

---

> **Next Step:**  
> Set \( P'(t) = 0 \) to find when the growth rate becomes zero.  
> These values of \( t \) are called **critical points**, and they’re the candidates for the population’s peaks and valleys.


### Step 2: Find the Critical Points

Now that we have the derivative, we can find where the population stops increasing or decreasing — these are called **critical points**.

Critical points occur where:

- The **slope is zero** (a flat or horizontal tangent line), or  
- The **derivative is undefined** (sharp corners or cusps — we’ll explore those later)

For this model, we set the derivative equal to zero:

\[
P'(t) = -3t^2 + 12t - 9 = 0
\]

This is a **quadratic equation**, so we can solve it by **factoring** or using the **quadratic formula**.

---

#### Step 2a: Solve by Factoring

First, simplify by dividing both sides by \(-3\):

\[
t^2 - 4t + 3 = 0
\]

Now factor:

\[
(t - 1)(t - 3) = 0
\]

So:

\[
t = 1 \quad \text{and} \quad t = 3
\]

These are our **critical points** — times when the slope is zero and the population might reach a **maximum** or **minimum**.

---

#### Step 2b: Solving with the Quadratic Formula (if needed)

If factoring doesn’t work, we can use the **quadratic formula**:

\[
t = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}
\]

For \( P'(t) = -3t^2 + 12t - 9 \):

\[
a = -3, \quad b = 12, \quad c = -9
\]

Substitute into the formula:

\[
t = \frac{-12 \pm \sqrt{(12)^2 - 4(-3)(-9)}}{2(-3)} = \frac{-12 \pm \sqrt{144 - 108}}{-6}
\]

\[
t = \frac{-12 \pm 6}{-6}
\]

\[
t = 1 \quad \text{and} \quad t = 3
\]

---

### Interpretation

The slope of \( P(t) \) becomes zero at **\( t = 1 \)** and **\( t = 3 \)** 

These are the **critical times** when the population stops changing direction.  
In the next step, we’ll determine **which point is a maximum** (peak population)  
and **which is a minimum** (lowest population) by examining the **sign of the derivative** around them.

> **Key Idea:**  
> Critical points are where “change pauses.”  
> The derivative equals zero, marking potential peaks or valleys in the system’s behavior.


### Step 3: Construct a Sign Chart

Now that we’ve found the **critical points** at \( t = 1 \) and \( t = 3 \), we can determine whether each one represents a **maximum** or a **minimum**.  
To do this, we analyze how the **sign of the derivative** changes around those points.

---

### Why Use a Sign Chart?

The **first derivative** tells us if the function is:

- **Positive** → the function is **increasing** (uphill)
- **Negative** → the function is **decreasing** (downhill)

At a **maximum**, the slope changes from positive to negative.  
At a **minimum**, the slope changes from negative to positive.

A **sign chart** helps us visualize these changes.

---

### The Derivative (Factored Form)

We’ll use the factored version of the derivative:

\[
P'(t) = -3(t - 1)(t - 3)
\]

Because it’s already factored, we can easily test the sign of each factor in different regions.

---

### Step 3a: Divide the Number Line

Mark the **critical points** on a number line. These divide the domain into three intervals:

- Interval A: \( (-\infty, 1) \)  
- Interval B: \( (1, 3) \)  
- Interval C: \( (3, \infty) \)


```{r, echo=FALSE}
library(ggplot2)

ggplot() +
  # Number line with arrowheads
  geom_segment(aes(x = 0, xend = 5, y = -0, yend = 0),
               arrow = arrow(type = "closed", ends = "both", length = unit(0.15, "inches")),
               linewidth = 1.2) +
  
  # Critical points
  geom_point(aes(x = 1, y = 0), size = 3) +
  geom_point(aes(x = 3, y = 0), size = 3) +
  
  # Labels just above the line
  annotate("text", x = 1, y = 0.07, label = "1 (critical point)", hjust = 0.5) +
  annotate("text", x = 3, y = 0.07, label = "3 (critical point)", hjust = 0.5) +
  
  # Interval labels just below the line
  annotate("text", x = 0.5, y = -0.07, label = "A", color = "blue") +
  annotate("text", x = 2, y = -0.07, label = "B", color = "forestgreen") +
  annotate("text", x = 4, y = -0.07, label = "C", color = "red") +
   coord_cartesian(ylim = c(-1, 1))+
  theme_void()+ ylim(-0.3,0.3)

```
                 
---

### Step 3b: Choose Test Points

Next, pick one **test value** in each interval to check whether the derivative is positive or negative in that region.  
Remember, we only need one test point per interval — if the derivative’s sign is positive or negative at one point in an interval, it will have the same sign everywhere in that interval.

| **Interval** | **Test Point** | **\( t - 1 \)** | **\( t - 3 \)** | **× \(-3\)** | **Sign of \( P'(t) \)** | **Behavior of \( P(t) \)** |
|:-------------:|:--------------:|:---------------:|:---------------:|:-------------:|:-------------------------:|:---------------------------:|
| \( t < 1 \)   | 0              | \(-\)           | \(-\)           | \(-\)         | **Negative**              | **Decreasing**              |
| \( 1 < t < 3 \) | 2             | \(+\)           | \(-\)           | \(+\)         | **Positive**              | **Increasing**              |
| \( t > 3 \)   | 4              | \(+\)           | \(+\)           | \(-\)         | **Negative**              | **Decreasing**              |


---

### Step 3c: Interpret the Signs

Let’s translate what the sign chart tells us about the population’s behavior:

- For \( t < 1 \):  \( P'(t) < 0 \) → population is **decreasing**
- For \( 1 < t < 3 \): \( P'(t) > 0 \) → population is **increasing**
- For \( t > 3 \): \( P'(t) < 0 \) → population is **decreasing** again

That means:

- At **\( t = 1 \)**, the slope changes from negative → positive → **local minimum**  
- At **\( t = 3 \)**, the slope changes from positive → negative → **local maximum**

---

### Step 3d: What the Signs Mean

| Change in \( P'(t) \) | What Happens to \( P(t) \) | Type of Point |
|:----------------------:|:---------------------------:|:--------------:|
| \( + \to - \) | Increasing → Decreasing | **Local Maximum** |
| \( - \to + \) | Decreasing → Increasing | **Local Minimum** |
| No sign change | Same direction | **Not an Extremum** |

---

### Step 3e: Key Takeaway

The **sign chart** gives a simple, visual way to connect the derivative’s sign to the shape of the function.

- **Positive slope** → the function is going up.  
- **Negative slope** → the function is going down.  
- **Sign changes** → that’s where turning points (maxima or minima) occur.

> **In summary:**  
> The population \( P(t) \) decreases, reaches a **minimum** at \( t = 1 \), increases to a **maximum** at \( t = 3 \), then begins to decrease again.  
>  
> The derivative’s sign tells the whole story of how the population grows and declines over time.


### Step 4: Evaluate the Function

Now that we’ve found the **critical points** at \( t = 1 \) and \( t = 3 \), let’s calculate the **population values** at those times by substituting them into the original function:

\[
P(t) = -t^3 + 6t^2 - 9t + 20
\]

---

#### Evaluate at \( t = 1 \)

\[
P(1) = -(1)^3 + 6(1)^2 - 9(1) + 20
\]
\[
P(1) = -1 + 6 - 9 + 20 = 16
\]

#### Evaluate at \( t = 3 \)

\[
P(3) = -(3)^3 + 6(3)^2 - 9(3) + 20
\]
\[
P(3) = -27 + 54 - 27 + 20 = 20
\]

---

### Step 4a: Interpret the Results

| \( t \) | \( P(t) \) | Type | Description |
|:--------:|:-----------:|:----:|:------------:|
| 1 | 16 | **Local Minimum** | Population reaches its lowest point before growing again |
| 3 | 20 | **Local Maximum** | Population peaks before starting to decline |

So we can write our extrema as:
\[
\text{Local Minimum: } (1, 16) \quad \text{and} \quad \text{Local Maximum: } (3, 20)
\]

---

### Summary

- **At \( t = 1 \):** Population = 16 → local **minimum**  
- **At \( t = 3 \):** Population = 20 → local **maximum**  

Between these points, the population decreases, then grows to a peak, and finally begins to decline — a pattern that mirrors how populations often respond to shifting environmental conditions.



```{r cubic-extrema-plot, echo=FALSE, fig.height=4}
P <- function(t) -t^3 + 6*t^2 - 9*t + 20

curve(P(x), from = 0, to = 5, ylim = c(0, 25), col = "forestgreen", lwd = 2,
      xlab = "Time (t)", ylab = "Population P(t)",
      main = "Cubic Model with Extrema")

# Critical point coordinates
t1 <- 1
t2 <- 3
P1 <- P(t1)
P2 <- P(t2)

# Mark critical points
points(t1, P1, pch = 19, col = "blue")
points(t2, P2, pch = 19, col = "red")

# Dashed horizontal lines at population values
abline(h = c(P1, P2), lty = 2, col = "gray60")

# Dashed vertical lines at t-values
abline(v = c(t1, t2), lty = 2, col = "gray70")

# Labels
text(t1, P1 + 1.5, labels = "Local Min\n(1, 16)", col = "blue", pos = 3)
text(t2, P2 , labels = "Local Max\n(3, 20)", col = "red", pos = 3)

```

### Final Takeaways

- **Work with the factored form** of the derivative — it makes sign analysis quicker and more transparent.  
- **Test just one value per interval** — if the derivative’s sign is positive or negative at one point, it’s the same throughout that entire region.  
- **Remember the leading coefficient** — it can flip all the signs in your chart if it’s negative.  
- The **sign of \( P'(t) \)** tells you whether the function \( P(t) \) is **increasing** (positive slope) or **decreasing** (negative slope).  

Using a **first derivative sign chart** allows you to clearly determine whether each **critical point** represents a **maximum**, **minimum**, or neither — turning algebraic work into a visual map of how the system behaves.


---

## Second Derivative Sign Charts: Understanding Concavity

The **first derivative** tells us whether a function is increasing or decreasing.  
The **second derivative** goes one level deeper — it tells us how that rate of change itself is changing.  
In other words, it describes the **curvature** or **shape** of the function’s graph, also known as **concavity**.

```{r concavity-visualization-smooth-2x2, message=FALSE, echo=FALSE, warning=FALSE, fig.width=9, fig.height=8}
library(ggplot2)
library(patchwork)
library(dplyr)

# Helper function for smooth concavity visualization
plot_concavity <- function(f, f2, range = c(-2, 2), inf_pts = NULL, title, subtitle) {
  x_vals <- seq(range[1], range[2], length.out = 400)
  df <- data.frame(x = x_vals, y = f(x_vals), f2 = f2(x_vals))
  inf_df <- if (!is.null(inf_pts)) data.frame(x = inf_pts, y = f(inf_pts)) else NULL
  
  p <- ggplot(df, aes(x, y)) +
    geom_line(color = "black", linewidth = 1.3) +
    theme_minimal(base_size = 12) +
    labs(title = title, subtitle = subtitle, x = "x", y = "f(x)") +
    theme(legend.position = "none")
  
  # Add background shading for concavity regions
  if (!is.null(inf_pts)) {
    # Determine region bounds for shading
    region_bounds <- c(range[1], inf_pts, range[2])
    for (i in seq_along(region_bounds[-1])) {
      region_fill <- if (mean(df$f2[df$x >= region_bounds[i] & df$x <= region_bounds[i + 1]]) > 0) "blue" else "red"
      p <- p +
        annotate("rect",
                 xmin = region_bounds[i], xmax = region_bounds[i + 1],
                 ymin = -Inf, ymax = Inf,
                 alpha = 0.08,
                 fill = region_fill)
    }
    # Add inflection points
    p <- p +
      geom_vline(xintercept = inf_pts, linetype = "dashed", color = "gray50") +
      geom_point(data = inf_df, aes(x, y), color = "purple", size = 3) +
      geom_text(data = inf_df, aes(x, y + 0.4, label = "Inflection"),
                color = "purple", vjust = -0.3, size = 3.5)
  } else {
    # For always up/down curves, just shade entire background
    fill_color <- if (mean(df$f2) > 0) "blue" else "red"
    p <- p +
      annotate("rect", xmin = range[1], xmax = range[2],
               ymin = -Inf, ymax = Inf, alpha = 0.08, fill = fill_color)
  }
  
  return(p)
}

# ----------------------------------------------------------
# A. Standard concavity change (x³)
f1  <- function(x) x^3
f2_1 <- function(x) 6*x
p1 <- plot_concavity(f1, f2_1, c(-2, 2), 0,
                     "A. Standard Concavity Change",
                     expression(f(x) == x^3))

# ----------------------------------------------------------
# B. Always concave up (x²)
f2  <- function(x) x^2
f2_2 <- function(x) 2
p2 <- plot_concavity(f2, f2_2, c(-2, 2), NULL,
                     "B. Always Concave Up",
                     expression(f(x) == x^2))

# ----------------------------------------------------------
# C. Always concave down (−x²)
f3  <- function(x) -x^2
f2_3 <- function(x) -2
p3 <- plot_concavity(f3, f2_3, c(-2, 2), NULL,
                     "C. Always Concave Down",
                     expression(f(x) == -x^2))

# ----------------------------------------------------------
# D. Multiple inflection points (0.2x⁵ − x³)
f4  <- function(x) 0.2*x^5 - x^3
f2_4 <- function(x) 4*x^3 - 6*x
inf_pts <- c(-sqrt(1.5), 0, sqrt(1.5))
p4 <- plot_concavity(f4, f2_4, c(-2, 2), inf_pts,
                     "D. Multiple Changes in Concavity",
                     expression(f(x) == 0.2*x^5 - x^3))

# Combine into 2×2 panel
(p1 | p2) / (p3 | p4)



```

#### Understanding Concavity Through Visual Examples

This 2×2 panel shows how the **second derivative** helps us interpret the **shape** or **curvature** of a function.  
While the **first derivative** tells us whether a function is increasing or decreasing,  
the **second derivative** tells us *how that rate of change itself is changing* — in other words, the function’s **concavity**.

---

**Panel A – Standard Concavity Change (\( f(x) = x^3 \))** 

- The curve bends downward (red) for \( x < 0 \) and upward (blue) for \( x > 0 \).  
- At \( x = 0 \), the concavity changes sign — this is a **classic inflection point**.  
- The slope passes smoothly through zero, marking a transition from “frowning” to “smiling” curvature.

---

**Panel B – Always Concave Up (\( f(x) = x^2 \))**  

- The curve opens upward everywhere.  
- The second derivative is always positive (\( f''(x) = 2 > 0 \)).  
- There are **no inflection points**, because the curvature never changes direction.  
- This kind of shape models stable, bowl-like growth (e.g., potential wells, cost functions).

---

**Panel C – Always Concave Down (\( f(x) = -x^2 \))**  

- The curve bends downward everywhere, like an inverted bowl.  
- The second derivative is always negative (\( f''(x) = -2 < 0 \)).  
- Again, no inflection points — the function is entirely “frowning.”  
- This is the opposite of Panel B and appears in models of diminishing returns or peaked processes.

---

**Panel D – Multiple Concavity Changes (\( f(x) = 0.2x^5 - x^3 \))**

- The curve alternates between concave down (red) and concave up (blue).  
- The **three inflection points** mark where the function’s curvature flips direction.  
- Such patterns often arise in complex systems — e.g., ecosystems or climate variables —  
  where growth, decline, and recovery phases alternate.

---

**In summary:**  

- **Concave up (blue)** means the slope is increasing — the function curves upward.  
- **Concave down (red)** means the slope is decreasing — the function curves downward.  
- **Inflection points** occur where \( f''(x) = 0 \) *and* the sign of \( f''(x) \) changes.  
- Visualizing concavity helps us understand **how processes accelerate or decelerate**,  
  revealing turning behaviors and transition phases in environmental systems.


---

### What Does the Second Derivative Tell Us?

Think of the graph of a function like a landscape of hills and valleys:

- **Concave Up** → shaped like a **valley** (opens upward)  
  → the slope is **increasing**  
  → the second derivative is **positive**

- **Concave Down** → shaped like a **hill** (opens downward)  
  → the slope is **decreasing**  
  → the second derivative is **negative**

So while the first derivative tracks *how fast* the function is changing,  
the second derivative tracks *how the rate of change itself is changing*.

---

### Step-by-Step: Building a Second Derivative Sign Chart

The process is almost identical to the one you used for the first derivative:

1. **Take the second derivative** of the function \( f''(x) \).  
2. **Set \( f''(x) = 0 \)** (or find where it’s undefined) — these are potential **inflection points**.  
3. **Plot those values** on a number line.  
4. **Pick test points** in each interval and plug them into \( f''(x) \) to determine its sign.  
5. **Interpret the signs**:
   - \( f''(x) > 0 \) → function is **concave up** (valley shape)  
   - \( f''(x) < 0 \) → function is **concave down** (hill shape)

> **Tip:** The difference between the first and second derivative charts lies in *what they describe*.  
> The first derivative shows when the function is increasing or decreasing,  
> while the second derivative shows how the *slope itself* is bending.

---

### What Are Inflection Points?

An **inflection point** is where the graph changes concavity — from **concave up** to **concave down**, or vice versa.  
It’s a “bend change” in the curve.

These occur where the **second derivative changes sign**.  

```{r inflection-types-panel-updated, echo=FALSE, warning=FALSE, fig.width=9, fig.height=8}
library(ggplot2)
library(patchwork)
library(dplyr)

# Helper function for consistent plotting
plot_inflection <- function(f, f2, range, inf_pts, title, subtitle) {
  x_vals <- seq(range[1], range[2], length.out = 400)
  df <- data.frame(x = x_vals, y = f(x_vals), f2 = f2(x_vals))
  df <- df %>%
    mutate(concavity = ifelse(f2 > 0, "Concave Up (f'' > 0)", "Concave Down (f'' < 0)"))
  inf_df <- data.frame(x = inf_pts, y = f(inf_pts))
  
  ggplot(df, aes(x, y)) +
    geom_line(aes(color = concavity), linewidth = 1.2) +
    geom_vline(xintercept = inf_pts, linetype = "dashed", color = "gray50") +
    geom_point(data = inf_df, aes(x, y), color = "purple", size = 3) +
    geom_text(data = inf_df, aes(x, y + 0.5, label = "Inflection"), color = "purple", vjust = -0.3) +
    scale_color_manual(values = c("Concave Down (f'' < 0)" = "red",
                                  "Concave Up (f'' > 0)" = "blue")) +
    labs(title = title, subtitle = subtitle, x = "x", y = "f(x)", color = "Concavity") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none")
}

# ----------------------------------------------------------
# A. Standard smooth S-shape  f(x) = x^3
f1  <- function(x) x^3
f2_1 <- function(x) 6*x
p1 <- plot_inflection(f1, f2_1, c(-3,3), 0,
                      "A. Standard Inflection",
                      expression(f(x) == x^3))

# ----------------------------------------------------------
# B. Flat inflection  f(x) = x^5
f2  <- function(x) x^5
f2_2 <- function(x) 20*x^3
p2 <- plot_inflection(f2, f2_2, c(-2,2), 0,
                      "B. Flat Inflection (Slope = 0 at Inflection)",
                      expression(f(x) == x^5))

# ----------------------------------------------------------
# C. Steep / nonzero-slope inflection  f(x) = (x-2)^3 + (x-2)
f3  <- function(x) (x-2)^3 + (x-2)
f2_3 <- function(x) 6*(x-2)
p3 <- plot_inflection(f3, f2_3, c(0,4), 2,
                      "C. Inflection with Nonzero Slope",
                      expression(f(x) == (x-2)^3 + (x-2)))

# ----------------------------------------------------------
# E. Steep inflection (tanh)
f5  <- function(x) tanh(x)
f2_5 <- function(x) -2 * tanh(x) / (cosh(x)^2)
p4 <- plot_inflection(f5, f2_5, c(-4,4), 0,
                      "D. Steep Inflection — Rapid Concavity Change",
                      expression(f(x) == tanh(x)))

# Combine into 2×2 grid
(p1 | p2) / (p3 | p4)
```

---

#### Summary: Understanding Different Types of Inflection Points

Inflection points mark where a function’s **concavity changes** — from curving upward to curving downward or vice versa.  

They reveal how the **rate of change itself is changing**, and they can take several distinct forms depending on the shape of the function.

---

#### **A. Standard Inflection — Smooth S-Shape (\( f(x) = x^3 \))**  {-}

- The classic “S-curve.”  
- The slope passes smoothly through zero.  
- The concavity flips cleanly from down to up.  
- Common in population or logistic-type models during a mid-growth phase.  

**Key idea:** The curvature change is balanced and gradual.

---

#### **B. Flat Inflection — Gentle Transition (\( f(x) = x^5 \))**  {-}

- The graph flattens around the inflection point — almost straight.  
- Both \( f'(x) \) and \( f''(x) \) are close to zero for a range of values.  
- The curvature changes slowly rather than sharply.

**Key idea:** Concavity changes, but very gradually — typical in smoothed or plateauing growth.

---

#### **C. Tilted Inflection — Nonzero Slope (\( f(x) = (x - 2)^3 + (x - 2) \))**  {-}

- The function is slanted, not symmetric.  
- The inflection occurs while the function is still increasing or decreasing steadily.  
- The slope is **not** zero at the inflection.  

**Key idea:** A concavity change can happen even when the function is rising or falling — the slope doesn’t need to flatten.

---

#### **D. Steep Inflection — Rapid Curvature Flip (\( f(x) = \tanh(x) \))**  {-}

- The curvature switches direction very quickly.  
- The slope becomes extremely steep near the inflection.  
- Found in **sigmoid** or **logistic-type** curves — classic in ecology, chemistry, and systems modeling.

**Key idea:** The inflection marks a **rapid acceleration–deceleration transition**, where the system shifts behavior quickly.

---

### Why This Matters in Environmental Modeling

Inflection points are where **processes change momentum** —  
they tell us *when growth slows down, acceleration reverses, or a new regime begins.*

Examples:

- When a population transitions from rapid growth to saturation.  
- When CO₂ concentration stops accelerating and starts to stabilize.  
- When temperature trends bend toward a new climate regime.

---

### Key Takeaways

- Inflection points occur where **\( f''(x) = 0 \)** *and* the **sign of \( f''(x) \)** changes.  
- They describe **how the rate of change itself changes** — not just whether the function increases or decreases.  
- Not all inflections look the same — they can be **flat, slanted, or steep**, depending on the function’s shape.  
- Recognizing these patterns helps interpret **tipping points** and **transitions** in environmental systems.

---

### Summary Table

| Interval | \( f''(x) \) Sign | Concavity | Interpretation |
|-----------|------------------|------------|----------------|
| \( f''(x) > 0 \) | Positive | Concave Up | Slope is increasing (valley) |
| \( f''(x) = 0 \) | Zero | Possible Inflection | Check for sign change |
| \( f''(x) < 0 \) | Negative | Concave Down | Slope is decreasing (hill) |

---

### Key Takeaway

- The **first derivative** tells us where a function rises or falls.  
- The **second derivative** tells us how that rising or falling **bends** — whether it’s speeding up or slowing down.  
- Together, they give a **complete picture** of a system’s behavior: when it’s growing, when it’s stabilizing, and when it’s beginning to turn.


## Checking Asymptotes: Interpreting Limits and Long-Term Behavior

Derivatives tell us how a function **changes** — where it rises, falls, or bends.  
But to understand the **boundaries** of a system — what happens far into the future, or near values where behavior “blows up” — we turn to **asymptotes** and **limits**.

Asymptotes describe how a system behaves at its **edges**: under extreme conditions or near critical thresholds where the function becomes undefined.

---

### Why Asymptotes Matter in Environmental Systems

In environmental modeling, asymptotes often represent **natural limits** or **steady states** — the values that processes approach but never quite reach.  
They can also signal **thresholds** where a process diverges or becomes undefined.

Examples:

- **Logistic population models** approach a **horizontal asymptote** at the **carrying capacity** — growth slows but never exceeds it.  
- **Pollutant concentration** decays toward zero, forming a horizontal asymptote.  
- **Runoff models** or rational functions can have **vertical asymptotes** where the model becomes undefined — such as a saturation or overflow threshold.

In short: asymptotes describe the **long-term limits** and **critical boundaries** of change.

---

### The Two Types of Asymptotes

| Type | Description | How to Check |
|------|--------------|--------------|
| **Vertical Asymptote** | Function value grows toward \( \pm\infty \) near some \( x = a \) | Examine the denominator → does \( f(x) \to \infty \) or \( -\infty \) as \( x \to a \)? |
| **Horizontal Asymptote** | Function approaches a constant value as \( x \to \pm\infty \) | Take the **limit** of \( f(x) \) as \( x \to \infty \) and \( x \to -\infty \) |

---

### Step-by-Step: How to Check for Asymptotes

#### **1. Vertical Asymptotes** {-}

Look for where the function is **undefined** (usually where the denominator = 0).  
Then check the **limit** from both sides.

\[
\text{If } \lim_{x \to a^+} f(x) = \pm\infty \text{ or } \lim_{x \to a^-} f(x) = \pm\infty,
\text{ then } x = a \text{ is a vertical asymptote.}
\]

> **Example:**  
> \( f(x) = \dfrac{1}{x - 2} \)  
> Denominator → 0 when \( x = 2 \).  
> As \( x \to 2^- \), \( f(x) \to -\infty \); as \( x \to 2^+ \), \( f(x) \to +\infty \).  
> So \( x = 2 \) is a **vertical asymptote**.

---

#### **2. Horizontal Asymptotes** {-}

These describe the **end behavior** — what happens as \( x \to \infty \) or \( x \to -\infty \).

\[
\text{If } \lim_{x \to \infty} f(x) = L \text{ or } \lim_{x \to -\infty} f(x) = L,
\text{ then } y = L \text{ is a horizontal asymptote.}
\]

> **Example:**  
> \( f(x) = \dfrac{3x^2 + 5}{x^2 + 1} \)  
> As \( x \to \infty \), leading terms dominate → \( \dfrac{3x^2}{x^2} = 3 \).  
> So \( y = 3 \) is a **horizontal asymptote**.

In environmental terms, this might describe a pollutant concentration that rises and then stabilizes at a long-term equilibrium.

---

### Using L’Hôpital’s Rule to Evaluate Asymptotic Limits

Sometimes, when finding asymptotes, direct substitution gives an **indeterminate form** such as  
\[
\frac{0}{0} \quad \text{or} \quad \frac{\infty}{\infty}.
\]  
In these cases, **L’Hôpital’s Rule** allows us to use derivatives to evaluate the limit and confirm the asymptotic behavior.

---

#### The Rule {-}

If both the numerator and denominator approach **0** or **∞**,  
then:

\[
\lim_{x \to a} \frac{f(x)}{g(x)} =
\lim_{x \to a} \frac{f'(x)}{g'(x)},
\]
provided \( f \) and \( g \) are differentiable near \( a \) and the new limit exists.

---

#### Step-by-Step {-}

1. **Check the form.** Confirm it’s \( 0/0 \) or \( ∞/∞ \).  
2. **Differentiate numerator and denominator separately.**  
3. **Re-evaluate the limit.** If it’s still indeterminate, apply the rule again.

---

#### Example 1 — Exponential Decay {-}

\[
\lim_{x \to \infty} \frac{x}{e^x}.
\]

Both → \( \infty \). Apply L’Hôpital’s Rule:

\[
\lim_{x \to \infty} \frac{1}{e^x} = 0.
\]

So \( y = 0 \) is a **horizontal asymptote**.  
In context, this could describe a population or pollutant diminishing toward zero over time.

---

#### Example 2 — Sine Ratio {-}

\[
\lim_{x \to 0} \frac{\sin x}{x}.
\]

Both → 0. Differentiate:

\[
\lim_{x \to 0} \frac{\cos x}{1} = 1.
\]

The ratio stabilizes to 1 — showing smooth near-zero behavior, not a blow-up.

---

#### Example 3 — Environmental Ratio {-}

\[
\lim_{t \to \infty} \frac{e^{-0.1t}}{e^{-0.3t}}.
\]

Both → 0. Apply the rule:

\[
\lim_{t \to \infty} \frac{-0.1e^{-0.1t}}{-0.3e^{-0.3t}}
= \frac{0.1}{0.3} e^{0.2t} \to \infty.
\]

No horizontal asymptote — uptake dominates decay as time grows.

---

### Key Takeaways

- **Vertical asymptotes** show where functions become undefined or diverge — often thresholds or physical limits.  
- **Horizontal asymptotes** show long-term limits — steady states or equilibrium conditions.  
- **L’Hôpital’s Rule** helps confirm these limits when algebra gives indeterminate forms.

---

### Quick Practice — Asymptotes & Limits

Find horizontal and vertical asymptotes (if any):

1. \( f(x) = \dfrac{2x^2 - 1}{x^2 + 1} \)  
2. \( f(x) = \dfrac{1}{x - 4} \)  
3. \( f(x) = \dfrac{x}{e^x} \)  
4. \( f(x) = \dfrac{e^x - 1}{x} \)

<details>
<summary><strong>Show Example Answers</strong></summary>

| Function | Asymptote | Type | How Found |
|-----------|------------|------|-----------|
| \( \dfrac{2x^2 - 1}{x^2 + 1} \) | \( y = 2 \) | Horizontal | Leading term ratio |
| \( \dfrac{1}{x - 4} \) | \( x = 4 \) | Vertical | Denominator → 0 |
| \( \dfrac{x}{e^x} \) | \( y = 0 \) | Horizontal | L’Hôpital’s Rule |
| \( \dfrac{e^x - 1}{x} \) | none | – | Limit \( \to \infty \) as \( x \to \infty \) |

</details>

---

### Reflection

1. How does L’Hôpital’s Rule connect **derivatives** to **asymptotic limits**?  
2. Why might a **horizontal asymptote** represent equilibrium, while a **vertical asymptote** represents system failure or a boundary?  
3. How can evaluating these limits guide environmental interpretations — like steady-state pollution, population plateaus, or saturation thresholds?

> **Big Idea:**  
> Derivatives describe **how** systems change.  
> Asymptotes and limits reveal **where** those systems are headed.



## Practice: Applying Derivatives to Real-World Change

These exercises will help you reinforce the key ideas from this chapter — how to identify where functions rise and fall, locate local and global extrema, and interpret concavity and inflection points in context.

Work symbolically where possible, but always connect your mathematical findings back to what they *mean* in an environmental system.

---

### Practice A — Local vs. Global Extrema and Increasing/Decreasing Behavior {-}

Answer each question briefly. These are concept checks — you don’t need full algebraic work unless noted.

---

#### 1. Concept Review {-}
a) In your own words, what is the difference between a **local maximum** and a **global maximum**?  
b) Can a function have more than one local maximum?  
c) Can it have more than one global maximum? Explain why or why not.  
d) Where might global extrema occur if the function is only defined over a **finite interval**?  
e) Give one environmental example where a **local maximum** matters more than the global one.

<details>
<summary><strong>Show Example Answers</strong></summary>

**a)**  
A **local maximum** is the highest point in a small neighborhood — higher than nearby values.  
A **global maximum** is the highest point across the entire function or dataset.

---

**b)**  
Yes — a function can have **multiple local maxima** if it rises and falls several times,  
like temperature peaks over several days.

---

**c)**  
Usually, there can be **only one global maximum**, since it’s the single highest value overall.  
However, two points can share the same **global maximum value** if they reach exactly the same height.

---

**d)**  
When a function is defined over a **finite interval**, global extrema can occur  
either at **critical points inside the interval** or at the **endpoints** of the domain.

---

**e)**  
An example is a **daily air temperature cycle** —  
the **local maximum** (afternoon high) matters more for understanding daily heat stress  
than the **global maximum** (the hottest day of the year).

</details>




---

#### 2. Identify Extrema from Description {-}
For each situation below, decide whether the described value is a **local** or **global** maximum/minimum — and explain why.

| Scenario | Your Answer | Why? |
|-----------|--------------|------|
| A lake’s surface temperature is warmest in July but colder in January. |  |  |
| The single hottest day recorded in 2024. |  |  |
| A species’ population peaks every spring but overall declines over decades. |  |  |
| A drought reaches its most severe week within a summer. |  |  |
| The coldest temperature ever recorded in the city’s history. |  |  |

<details>
<summary><strong>Show Example Answers</strong></summary>

| Scenario | Your Answer | Why? |
|-----------|:------------:|------|
| A lake’s surface temperature is warmest in July but colder in January. | **Local Maximum** | It’s the warmest point within a repeating yearly cycle, not the highest ever recorded. |
| The single hottest day recorded in 2024. | **Global Maximum** | It’s the highest temperature across the entire year’s data. |
| A species’ population peaks every spring but overall declines over decades. | **Local Maximum** | Each spring peak is local — it’s high for that season but not the highest overall. |
| A drought reaches its most severe week within a summer. | **Local Minimum** | It’s the lowest rainfall (or highest dryness) point within that season. |
| The coldest temperature ever recorded in the city’s history. | **Global Minimum** | It’s the lowest value across all data collected — the absolute record low. |

</details>

---

#### 3. Sketch and Label (Quick Graph Practice) {-}
Without detailed computation, sketch rough graphs showing examples of each:

1. A function with **one local max** and **one local min**.  
2. A function with **global extrema at endpoints**.  
3. A function that is **increasing everywhere** but has no maximum.  
4. A **periodic** function with repeating local maxima and minima.

Label your axes and mark each point as “local” or “global.”

<details>
<summary><strong>Show Example Sketches and Explanations</strong></summary>

**1. One Local Max and One Local Min**  
A simple **cubic** shape — rises, peaks, dips, and rises again.  
Example: \( f(x) = x^3 - 3x \)  
- Local max around \( x = -1 \)  
- Local min around \( x = 1 \)

---

**2. Global Extrema at Endpoints**  
A **downward-opening parabola** restricted to a finite domain.  
Example: \( f(x) = -x^2 + 4x \) on \( [0, 5] \)  
- Global max at \( x = 2 \) (inside domain)  
- Global min at \( x = 0 \) or \( x = 5 \) (endpoints)

---

**3. Increasing Everywhere — No Maximum**  
An **exponential growth** curve, always rising.  
Example: \( f(x) = e^x \)  
- No maximum; the function keeps increasing.  
- Minimum occurs at the smallest \( x \) in the domain.

---

**4. Periodic Function with Repeating Maxima and Minima**  
A **sine wave** pattern.  
Example: \( f(x) = \sin(x) \)  
- Local max at \( x = \pi/2, \, 5\pi/2, \ldots \)  
- Local min at \( x = 3\pi/2, \, 7\pi/2, \ldots \)

---

**Sketching Notes:**  
Label your axes clearly.  
Mark approximate peaks and valleys with small dots and note whether they’re local or global.  
You don’t need perfect scaling — focus on capturing shape and turning points.

</details>

---

#### 4. Visual Interpretation: Reading Extrema from a Graph {-}

Look at the following function graphs (a smooth curve representing \( f(x) \)):  

```{r extrema-2x2-panel, echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=8}
library(ggplot2)
library(patchwork)

# 1. Function with local extrema only (cubic)
f1 <- function(x) x^3 - 3*x^2 + 2
df1 <- data.frame(x = seq(-1, 4, length.out = 300), y = f1(seq(-1, 4, length.out = 300)))
p1 <- ggplot(df1, aes(x, y)) +
  geom_line(color = "steelblue", linewidth = 1.1) +
  labs(title = expression(f(x) == x^3 - 3*x^2 + 2),
       x = "x", y = "f(x)") +
  theme_minimal(base_size = 13)

# 2. Quadratic with defined interval — endpoints matter
f2 <- function(x) -x^2 + 4*x
df2 <- data.frame(x = seq(0, 5, length.out = 300), y = f2(seq(0, 5, length.out = 300)))
p2 <- ggplot(df2, aes(x, y)) +
  geom_line(color = "darkorange", linewidth = 1.1) +
  geom_vline(xintercept = c(0, 5), linetype = "dashed", color = "gray50") +
  labs(title = expression(f(x) == -x^2 + 4*x ~ ",~ 0 <= x <= 5"),
       x = "x", y = "f(x)") +
  theme_minimal(base_size = 13)

# 3. Sinusoidal — repeated local extrema
f3 <- function(x) sin(x)
df3 <- data.frame(x = seq(0, 2*pi, length.out = 300), y = f3(seq(0, 2*pi, length.out = 300)))
p3 <- ggplot(df3, aes(x, y)) +
  geom_line(color = "forestgreen", linewidth = 1.1) +
  labs(title = expression(f(x) == sin(x) ~ ",~ 0 <= x <= 2*pi"),
       x = "x", y = "f(x)") +
  theme_minimal(base_size = 13)

# 4. Quartic with both local and global extrema
f4 <- function(x) x^4 - 4*x^2 + 2
df4 <- data.frame(x = seq(-2, 2, length.out = 300), y = f4(seq(-2, 2, length.out = 300)))
p4 <- ggplot(df4, aes(x, y)) +
  geom_line(color = "purple", linewidth = 1.1) +
  labs(title = expression(f(x) == x^4 - 4*x^2 + 2),
       x = "x", y = "f(x)") +
  theme_minimal(base_size = 13)

# Combine into 2x2 panel
(p1 + p2) / (p3 + p4)


```

Use the graphs to answer the questions below.  
No calculations — just interpret visually.

---

**a)** Where is the function **increasing**?  
Where is it **decreasing**?

**b)** Identify all **local maxima** and **local minima** by estimating their \( x \)-locations.  

**c)** Which point appears to be the **global maximum**?  
Which point is the **global minimum**?  

**d)** Does the graph suggest the global extrema occur **within** the domain or **at the endpoints**?

**e)** In words, describe what this pattern could represent in an environmental system —  
for example, a population rising, stabilizing, and then declining.

---

> **Reflection:**  
> When you read a graph like this, what clues tell you where the slope changes from positive to negative (and vice versa)?  
> How does that connect to the idea of the **derivative being zero** at turning points?

<details>
<summary><strong>Show Solution</strong></summary>

**Function 1: \( f(x) = x^3 - 3x^2 + 2 \)**  
- **Derivative:** \( f'(x) = 3x^2 - 6x = 3x(x - 2) \)  
- **Critical Points:** \( x = 0, 2 \)  
- \( f(0) = 2 \) (local **max**), \( f(2) = -2 \) (local **min**)  
- **No global extrema** (function extends to \( \pm \infty \))  

---

**Function 2: \( f(x) = -x^2 + 4x, \, 0 \le x \le 5 \)**  
- **Derivative:** \( f'(x) = -2x + 4 \) → zero at \( x = 2 \)  
- \( f(2) = 4 \) (local **and global max**)  
- Endpoints: \( f(0) = 0, f(5) = -5 \) → global **min** at \( x = 5 \)

---

**Function 3: \( f(x) = \sin(x), \, 0 \le x \le 2\pi \)**  
- **Derivative:** \( f'(x) = \cos(x) \) → zero at \( x = \pi/2, 3\pi/2 \)  
- \( f(\pi/2) = 1 \) (local **max**)  
- \( f(3\pi/2) = -1 \) (local **min**)  
- Both also serve as **global extrema** within this domain.

---

**Function 4: \( f(x) = x^4 - 4x^2 + 2 \)**  
- **Derivative:** \( f'(x) = 4x^3 - 8x = 4x(x^2 - 2) \)  
- **Critical Points:** \( x = 0, \pm \sqrt{2} \)  
- \( f(0) = 2 \) (local **max**)  
- \( f(\pm \sqrt{2}) = -2 \) (local and global **min**)  

---

**Summary:**

| Function | Local Max | Local Min | Global Max | Global Min |
|-----------|------------|------------|--------------|--------------|
| \( x^3 - 3x^2 + 2 \) | (0, 2) | (2, –2) | — | — |
| \( -x^2 + 4x, [0,5] \) | (2, 4) | — | (2, 4) | (5, –5) |
| \( \sin(x), [0,2\pi] \) | (π/2, 1) | (3π/2, –1) | (π/2, 1) | (3π/2, –1) |
| \( x^4 - 4x^2 + 2 \) | (0, 2) | (±√2, –2) | — | (±√2, –2) |

</details>


#### 5. Environmental Reflection {-}

For each model below, explain what a **maximum** and **minimum** represent physically:

- A **population growth curve**  
- A **pollutant concentration** through time  
- A **temperature cycle** across a year  

---

**Discussion Prompt:**  
How might a **local maximum** signal a short-term event,  
while a **global maximum** represents a long-term limit?

<details>
<summary><strong>Show Example Answers</strong></summary>

**Population Growth Curve**  
- A **maximum** represents the largest population the environment can support — often called the **carrying capacity**.  
- A **minimum** could represent a population crash or recovery low point after a disturbance.

---

**Pollutant Concentration Through Time**  
- A **maximum** indicates the **peak pollution level**, possibly following an industrial discharge or storm runoff.  
- A **minimum** shows when pollutant levels are **lowest**, perhaps after cleanup or dilution events.

---

**Temperature Cycle Across a Year**  
- The **maximum** represents the **hottest part of the year**, typically midsummer.  
- The **minimum** represents the **coldest period**, usually midwinter.

---

**Discussion:**  
- A **local maximum** reflects a short-term spike or seasonal high — a temporary event such as a heatwave, bloom, or discharge.  
- A **global maximum** represents the absolute upper limit across the full record — for example, the **warmest year on record** or the **highest CO₂ concentration** in history.

</details>


### Practice B — Algebraic Skill Building {-}

Find all \(x\)-values where \(f'(x)=0\). Keep answers exact. If a factor is repeated, note its multiplicity.

---

**1) Polynomials **  
a) \( f'(x)=4x(x-2)(x+3) \)  
b) \( f'(x)=-2(x-1)^2(x+4) \)  
c) \( f'(x)=3x^3-12x^2+9x \)  *(hint: factor out the 3 first)*  
d) \( f'(x)=(x^2-1)(x+2) \)

---

**2) Trig & Exponential Factors **  
e) \( f'(x)=e^{x}\big(x^2-9\big) \)  
f) \( f'(x)=\sin x\,(2\cos x-1) \)  
g) \( f'(x)=e^{2x}(x-5)^2(3-x) \)  
h) \( f'(x)=\cos x\,(1-2\sin x)(\sin x+1) \)

> *Notes:*  
> • \(e^{x}>0\) for all \(x\), so zeros come from the **other** factors.  
> • For trig factors, solve each factor’s equation on \(\mathbb{R}\) or the interval specified (if any).

---

**3) Mixed Product/Chain Structure**  
i) \( f'(x)=x^2(1-x)^3(2x+1) \)  
j) \( f'(x)=(x^2+1)(x-2)^4(\,x+ \sqrt{3}\,) \)  
k) \( f'(x)=\big(3x-1\big)^2\big(x^2-4\big)\big(x+1\big) \)

---

**4) Quotients — Remember:**  
If \( f'(x)=\dfrac{N(x)}{D(x)} \), then \( f'(x)=0 \iff N(x)=0 \) **and** \(D(x)\ne 0\). Also exclude points where \(D(x)=0\) (not in the derivative’s domain).

l) \( f'(x)=\dfrac{(x-3)(x+2)}{(x-1)^2} \)  
m) \( f'(x)=\dfrac{(2\sin x-1)\,(\cos x+1)}{x^2+1} \)  
n) \( f'(x)=\dfrac{(x-4)^3(x+5)}{(x^2-9)} \)  
o) \( f'(x)=\dfrac{e^{x}(x-1)}{(x+2)^4} \)

---

<details>
<summary><strong>Show Solutions </strong></summary>

**1) Polynomials**  
a) \(x=0,\,2,\,-3\) (all mult. 1)  
b) \(x=1\) (mult. 2), \(x=-4\) (mult. 1)  
c) \(3x^3-12x^2+9x=3x(x-3)(x-1)\Rightarrow x=0,1,3\) (all mult. 1)  
d) \((x^2-1)(x+2)=(x-1)(x+1)(x+2)\Rightarrow x=-2,-1,1\)

**2) Trig & Exponential**  
e) \(e^x\neq 0\Rightarrow x^2-9=0 \Rightarrow x=\pm 3\)  
f) \(\sin x=0\Rightarrow x=k\pi;\; 2\cos x-1=0\Rightarrow \cos x=\tfrac12 \Rightarrow x=\pm \tfrac{\pi}{3}+2k\pi\)  
g) \(e^{2x}\neq 0\Rightarrow (x-5)^2(3-x)=0 \Rightarrow x=5\) (mult. 2), \(x=3\)  
h) \(\cos x=0\Rightarrow x=\tfrac{\pi}{2}+k\pi;\; 1-2\sin x=0\Rightarrow \sin x=\tfrac12 \Rightarrow x=\tfrac{\pi}{6}+2k\pi,\; \tfrac{5\pi}{6}+2k\pi;\; \sin x+1=0\Rightarrow x=\tfrac{3\pi}{2}+2k\pi\)

**3) Mixed Product/Chain**  
i) \(x=0\) (mult. 2), \(x=1\) (mult. 3), \(x=-\tfrac12\)  
j) \(x=2\) (mult. 4), \(x=-\sqrt{3}\)  *(note: \(x^2+1\neq 0\))*  
k) \(x=\tfrac13\) (mult. 2), \(x=\pm 2\), \(x=-1\)

**4) Quotients** *(zeros from numerator, exclude denominator zeros)*  
l) Zeros: \(x=3,-2\); exclude \(x=1\) (denominator)  
m) Zeros: \(2\sin x-1=0\Rightarrow \sin x=\tfrac12\) and/or \(\cos x+1=0\Rightarrow x=\pi+2k\pi\); denominator \(x^2+1>0\) always  
n) Zeros: \(x=4\) (mult. 3), \(x=-5\); exclude \(x=\pm 3\) (denominator)  
o) Zeros: \(x=1\); exclude \(x=-2\) (denominator)

</details>

---

### Practice C — First Derivative Sign Chart Construction {-}

For each derivative,  
1. Factor if possible.  
2. Mark critical points on a number line.  
3. Use test points to build a **sign chart** for \( f'(x) \).  
4. Identify where \( f(x) \) is **increasing** or **decreasing**, and locate **local maxima/minima**.

| Function | Derivative |
|-----------|-------------|
| \( f(x) = -x^3 + 6x^2 - 9x + 20 \) | \( f'(x) = -3x^2 + 12x - 9 \) |
| \( f(x) = \sin(x) + \frac{x}{2} \) | \( f'(x) = \cos(x) + \frac{1}{2} \) |
| \( f(x) = x^4 - 4x^2 \) | \( f'(x) = 4x^3 - 8x \) |

> **Challenge:**  
> Sketch a rough graph of each function based only on your sign chart.  
> Where are the turning points, and how do you know?


<details>
<summary><strong>Show Solutions</strong></summary>

---

**1) \( f(x) = -x^3 + 6x^2 - 9x + 20 \)  **
\( f'(x) = -3x^2 + 12x - 9 = -3(x-1)(x-3) \)

**Critical points:** \(x=1,\,3\).

**Sign chart for \(f'(x)\):**  
Intervals: \((-\infty,1)\), \((1,3)\), \((3,\infty)\)  
- Test \(x=0\): \(f'(x)<0\) → **decreasing**  
- Test \(x=2\): \(f'(x)>0\) → **increasing**  
- Test \(x=4\): \(f'(x)<0\) → **decreasing**

**Conclusions:**  
- Increasing on \((1,3)\)  
- Decreasing on \((-\infty,1)\) and \((3,\infty)\)  
- \(x=1\): \(- \to +\) ⇒ **local minimum**  
- \(x=3\): \(+ \to -\) ⇒ **local maximum**

---

**2) \( f(x) = \sin x + \tfrac{x}{2} \)  **
\( f'(x) = \cos x + \tfrac{1}{2} \)

**Critical points (general):** solve \(\cos x = -\tfrac{1}{2}\)  
\[
x = \frac{2\pi}{3} + 2k\pi,\quad x = \frac{4\pi}{3} + 2k\pi,\quad k\in\mathbb{Z}.
\]

**On \([0,2\pi]\) for concreteness:** critical points at \(x=\tfrac{2\pi}{3},\tfrac{4\pi}{3}\).

**Sign chart on \([0,2\pi]\):**  
Intervals: \((0,\tfrac{2\pi}{3})\), \((\tfrac{2\pi}{3},\tfrac{4\pi}{3})\), \((\tfrac{4\pi}{3},2\pi)\)  
- Test \(x=0\): \(f'(0)=1+\tfrac12>0\) → **increasing**  
- Test \(x=\pi\): \(-1+\tfrac12<0\) → **decreasing**  
- Test \(x=\tfrac{3\pi}{2}\): \(0+\tfrac12>0\) → **increasing**

**Conclusions (repeat every \(2\pi\))**  
- Increasing on \(\big(2k\pi,\,\tfrac{2\pi}{3}+2k\pi\big)\) and \(\big(\tfrac{4\pi}{3}+2k\pi,\,2\pi+2k\pi\big)\)  
- Decreasing on \(\big(\tfrac{2\pi}{3}+2k\pi,\,\tfrac{4\pi}{3}+2k\pi\big)\)  
- \(x=\tfrac{2\pi}{3}+2k\pi\): \(+ \to -\) ⇒ **local maximum**  
- \(x=\tfrac{4\pi}{3}+2k\pi\): \(- \to +\) ⇒ **local minimum**

---

**3) \( f(x) = x^4 - 4x^2 \)**
\( f'(x) = 4x^3 - 8x = 4x(x^2-2) = 4x(x-\sqrt2)(x+\sqrt2) \)

**Critical points:** \(x=-\sqrt2,\,0,\,\sqrt2\).

**Sign chart for \(f'(x)\):**  
Intervals: \((-\infty,-\sqrt2)\), \((-\sqrt2,0)\), \((0,\sqrt2)\), \((\sqrt2,\infty)\)  
- Test \(x=-2\): \(\,f'(x)<0\) → **decreasing**  
- Test \(x=-\tfrac12\): \(f'(x)>0\) → **increasing**  
- Test \(x=\tfrac12\): \(f'(x)<0\) → **decreasing**  
- Test \(x=2\): \(f'(x)>0\) → **increasing**

**Conclusions:**  
- Increasing on \((-\sqrt2,0)\) and \((\sqrt2,\infty)\)  
- Decreasing on \((-\infty,-\sqrt2)\) and \((0,\sqrt2)\)  
- \(x=-\sqrt2\): \(- \to +\) ⇒ **local minimum**  
- \(x=0\): \(+ \to -\) ⇒ **local maximum**  
- \(x=\sqrt2\): \(- \to +\) ⇒ **local minimum**  
- (Values: \(f(0)=0\); \(f(\pm\sqrt2)=-4\) → those are the **global minima**.)

---

**Challenge (sketching from sign charts):**  
- Mark critical points on the \(x\)-axis.  
- Use the intervals of “increasing/decreasing” to draw the qualitative shape.  
- Place peaks where \(f'\) goes \(+\to-\) and valleys where \(f'\) goes \(-\to+\).  
- You may add concavity hints using \(f''\) if desired, but the sign chart alone is enough to locate turning points.

</details>


### Practice D — Second Derivative and Concavity{-}
For each function,  
a) Compute \( f''(x) \).  
b) Find where \( f''(x) = 0 \).  
c) Determine the intervals of **concave up** and **concave down**.  
d) Identify any **inflection points**.

| Function | Context |
|-----------|----------|
| \( f(x) = x^3 - 6x^2 + 9x \) | Population growth with competition |
| \( f(x) = \tanh(x) \) | Logistic response to environmental stress |
| \( f(x) = x^4 - 2x^2 \) | Potential energy curve of a bistable system |

> **Think about shape:**  
> How does a concave-up region differ in meaning from a concave-down region in these examples?


<details>
<summary><strong>Show Solutions</strong></summary>

---

**1) \( f(x) = x^3 - 6x^2 + 9x \) **  *(Population growth with competition)* 

**a) \( f''(x) \):**  
\( f'(x) = 3x^2 - 12x + 9 \) → \( f''(x) = 6x - 12 \)

**b) Solve \( f''(x)=0 \):**  
\( 6x - 12 = 0 \Rightarrow x = 2 \)

**c) Concavity:**  
- \( x < 2 \): \( f''(x) < 0 \) → **concave down**  
- \( x > 2 \): \( f''(x) > 0 \) → **concave up**

**d) Inflection point:**  
At \( x=2 \), \( f(2) = 8 - 24 + 18 = 2 \).  
**Inflection:** \( (2,\,2) \)

---

**2) \( f(x) = \tanh(x) \)**  *(Logistic response to environmental stress)*

**a) \( f''(x) \):**  
\( f'(x) = \operatorname{sech}^2(x) \)  
\( f''(x) = -2\,\operatorname{sech}^2(x)\,\tanh(x) \)

**b) Solve \( f''(x)=0 \):**  
\(\operatorname{sech}^2(x) > 0\) for all \(x\), so zero occurs when \(\tanh(x)=0 \Rightarrow x=0\).

**c) Concavity:**  
- \( x < 0 \): \(\tanh(x) < 0 \Rightarrow f''(x) > 0\) → **concave up**  
- \( x > 0 \): \(\tanh(x) > 0 \Rightarrow f''(x) < 0\) → **concave down**

**d) Inflection point:**  
\( f(0) = 0 \).  
**Inflection:** \( (0,\,0) \)

---

**3) \( f(x) = x^4 - 2x^2 \)**  

**a) \( f''(x) \):**  
\( f'(x) = 4x^3 - 8x \) → \( f''(x) = 12x^2 - 4 = 4(3x^2 - 1) \)

**b) Solve \( f''(x)=0 \):**  
\( 3x^2 - 1 = 0 \Rightarrow x = \pm \frac{1}{\sqrt{3}} \)

**c) Concavity:**  
- \( |x| < \frac{1}{\sqrt{3}} \): \( f''(x) < 0 \) → **concave down**  
- \( |x| > \frac{1}{\sqrt{3}} \): \( f''(x) > 0 \) → **concave up**

**d) Inflection points:**  
\( f\!\left(\pm \tfrac{1}{\sqrt{3}}\right) = \left(\tfrac{1}{\sqrt{3}}\right)^4 - 2\left(\tfrac{1}{\sqrt{3}}\right)^2
= \tfrac{1}{9} - \tfrac{2}{3} = -\tfrac{5}{9} \).  
**Inflections:** \( \left(\pm \tfrac{1}{\sqrt{3}},\, -\tfrac{5}{9}\right) \)

---

**Interpreting “concave up” vs “concave down” in context**

- **Concave up (\( f''>0 \))**: the rate of change is **increasing** (acceleration).  
  *Population example:* recovery phase where growth is speeding up.  
- **Concave down (\( f''<0 \))**: the rate of change is **decreasing** (deceleration).  
  *Population example:* crowding/competition slows growth as carrying capacity is approached.

</details>

---

### Practice E — Second Derivative Sign Charts and Inflection Behavior{-}

For each function,  
1. Find \( f''(x) \).  
2. Solve \( f''(x) = 0 \).  
3. Use test points to fill in a **second derivative sign chart**.  
4. Identify **concavity regions** and **inflection points**.

| Function | 
|:---------:|
| \( f(x) = x^3 - 3x \) |
| \( f(x) = \frac{x^3}{3} - x \) | 
| \( f(x) = 0.2x^5 - x^3 \) | 

**Interpretation:**  

- Where does each function “switch curvature”?  
- How might that correspond to a transition or tipping point in an environmental process (e.g., when growth stops accelerating)?

---

### Reflection

1. How does analyzing **derivatives** deepen your understanding of how systems *change* rather than simply *behave*?  
2. What does a “flat slope” or “bend change” mean in a real environmental context?  
3. Why do scientists care about *when* a rate of change stops increasing — even before it becomes negative?

> *Hint:* These are often the earliest signals of approaching thresholds or turning points —  
> moments when proactive management can still alter the course of change.


<details>
<summary><strong>Show Solutions</strong></summary>

---

**1) \( f(x) = x^3 - 3x \)**

**Second derivative:** \( f''(x) = 6x \)

**Solve \( f''(x)=0 \):** \( x=0 \)

**Sign chart for \( f''(x) \):**

| Interval | Test \(x\) | \(f''(x)\) sign | Concavity     |
|:-------:|:-----------:|:---------------:|---------------|
| \((-\infty,0)\) | \(-1\) | \(-\) | Concave **down** |
| \((0,\infty)\)  | \(1\)  | \(+\) | Concave **up**   |

**Inflection point:** \(x=0\), \(f(0)=0\) → \((0,0)\).

**Interpretation:** Curvature flips at 0 (classic S-shape); marks a transition from decelerating to accelerating growth.

---

**2) \( f(x) = \dfrac{x^3}{3} - x \)**

**Second derivative:** \( f''(x) = 2x \)

**Solve \( f''(x)=0 \):** \( x=0 \)

**Sign chart for \( f''(x) \):**

| Interval | Test \(x\) | \(f''(x)\) sign | Concavity     |
|:-------:|:-----------:|:---------------:|---------------|
| \((-\infty,0)\) | \(-1\) | \(-\) | Concave **down** |
| \((0,\infty)\)  | \(1\)  | \(+\) | Concave **up**   |

**Inflection point:** \(x=0\), \(f(0)=0\) → \((0,0)\).

**Interpretation:** Same qualitative behavior as #1; the scale changes but the curvature switch is identical.

---

**3) \( f(x) = 0.2x^5 - x^3 \)**

**Second derivative:**  
\( f'(x) = x^4 - 3x^2 \) so \( f''(x) = 4x^3 - 6x = 2x(2x^2 - 3) \)

**Solve \( f''(x)=0 \):**  
\( x=0 \) or \( 2x^2-3=0 \Rightarrow x=\pm\sqrt{\tfrac{3}{2}} \approx \pm 1.225 \)

**Sign chart for \( f''(x) \):**

| Interval | Test \(x\) | \(f''(x)\) sign        | Concavity       |
|:----------------------:|:-----------:|:-----------------------:|-----------------|
| \((-\infty,-\sqrt{3/2})\) | \(-2\) | \(-\) | Concave **down** |
| \((-\sqrt{3/2},0)\)       | \(-1\) | \(+\) | Concave **up**   |
| \((0,\sqrt{3/2})\)        | \(0.5\) | \(-\) | Concave **down** |
| \((\sqrt{3/2},\infty)\)   | \(2\) | \(+\) | Concave **up**   |

**Inflection points (sign changes at each zero of \(f''\)):**  
- \( x=-\sqrt{\tfrac{3}{2}} \), \( f(x) \approx +1.286 \)  
- \( x=0 \), \( f(0)=0 \)  
- \( x=+\sqrt{\tfrac{3}{2}} \), \( f(x) \approx -1.286 \)

**Interpretation:** Multiple curvature flips (down→up→down→up) — a richer transition pattern, typical of systems cycling through phases of acceleration and deceleration.

---

**Big-Picture Interpretation (all three)**

- A **curvature switch** (sign change in \( f'' \)) indicates a transition in *how* the system is changing (acceleration ↔ deceleration).  
- In environmental terms, these are early indicators of **tipping points** (e.g., growth moving from speeding up to slowing down before peaking).


**Reflection**

1. Derivatives let you study **change of change** (acceleration), revealing transitions before peaks or troughs appear.  
2. A **flat slope** (\( f'(x)=0 \)) marks a turning point; a **bend change** (sign change in \( f'' \)) marks a shift in acceleration — both are diagnostically different.  
3. Scientists watch for where the rate of change **stops increasing** ( \( f''\) changes sign) because it often precedes critical thresholds — a chance for **early intervention**.

</details>

---





