# Derivative Rules - Workbook

## Learning Objectives — Derivative Rules

By the end of this chapter, you should be able to:

1. **Read mathematical expressions as instructions** and decompose a function into sums, constants, powers, products, quotients, and compositions.  
2. **Apply basic rules** — constant, power, constant multiple, sum/difference, \( e^x \), \( \ln x \), \( \sin x \), \( \cos x \) — to differentiate common terms without reverting to limits.  
3. **Use the product, quotient, and chain rules** to differentiate functions built from multiple changing parts, showing clear intermediate steps and correct notation.  
4. **Select the right rule(s)** for a given function by identifying its top-level structure and any nested “inside–outside” relationships.  
5. **Interpret derivatives in context**, explaining what a computed rate of change means for an environmental quantity (e.g., pollutant concentration, biomass, streamflow).  
6. **Check and simplify results**, factoring, combining like terms, and performing quick reasonableness checks (units, signs, limiting behavior).  
7. **Communicate with multiple notations** — prime, Leibniz, and operator — and translate between them accurately.  
8. **Diagnose common errors**, such as forgetting the inner derivative or misusing product vs. chain rule, and correct them using structured reasoning or unit checks.  

---

**Success looks like:**  
Given a new environmental model, you can:

- Identify which derivative rules apply,  
- Compute \( f'(x) \) cleanly, and  
- Explain in words what that derivative tells you about how the system is changing.

## Activity: Reading Math as Instructions — The Derivative of a Constant

Not all functions *change*. A constant function is perfectly flat — no growth, no decline, no change over time or space.  

The derivative gives us a way to *prove* that this lack of change equals a slope of zero.

---

### Step 1: Read the Expression Aloud

::: promptbox
**Prompt:**  
Read this equation aloud and explain what each part is telling you to do:

\[
\frac{d}{dx}(c) = 0
\]

| Symbol / Term | What It Tells You To Do |
|:---------------|:------------------------|
| \( \frac{d}{dx} \) | |
| \( c \) | |
| \( = 0 \) | |

**Notes:**  
<br><br><br><br><br><br>
:::

---

### Step 2: Translate to a Step-by-Step Recipe

::: promptbox
**Activity:**  
In your own words, describe what the rule for a constant tells you to do.

1. Recognize that the function’s value **never changes** — it’s constant for all \( x \).  
2. The “rate of change” or “slope” is therefore **zero everywhere**.  
3. When you apply the derivative operator \( \frac{d}{dx} \), the output is \( 0 \).

Now apply this to three examples:

| Function \( f(x) \) | What It Represents | Derivative \( f'(x) \) |
|:--------------------:|:------------------:|:-----------------------:|
| \( f(x) = 7 \) | Constant value | |
| \( f(x) = -12 \) | Constant offset | |
| \( f(x) = 3.14 \) | Physical constant | |

**Notes:**  
<br><br><br><br><br><br><br>
:::

---

### Step 3: Deriving the Rule from the Limit Definition

We can confirm this result directly from the **limit definition of a derivative**:

\[
f'(x) = \lim_{h \to 0} \frac{f(x+h) - f(x)}{h}
\]

If \( f(x) = c \) (a constant), then:

\[
f'(x) = \lim_{h \to 0} \frac{c - c}{h} = \lim_{h \to 0} \frac{0}{h} = 0
\]

Because the numerator is always zero, the slope — the rate of change — is zero everywhere.

> In words: “When nothing changes, the rate of change is zero.”

---

### Step 4: Apply to Environmental Contexts

::: promptbox
**Practice:**  
For each situation, identify whether the derivative (rate of change) would be zero or nonzero.

| Scenario | Description | Is the Derivative Zero? | Why? |
|:----------|:-------------|:-----------------------:|:-----|
| A lake’s surface area remains 10 km² for 10 years | | | |
| Atmospheric CO₂ remains steady at 420 ppm for one month | | | |
| Soil temperature remains constant throughout the day | | | |

Now write one more *realistic* example where a quantity stays constant in your field (e.g., hydrology, forestry, marine science).

**Notes:**  
<br><br><br><br><br><br><br>
:::

---

### Step 5: Reflect — Flatness as Zero Change

::: promptbox
**Reflection Prompt:**  
Complete this thought:

> “When I see \( \frac{d}{dx}(c) = 0 \), I think of it as a statement that ________.”

**Notes:**  
<br><br><br><br><br><br>
:::



## Activity: Reading Math as Instructions — The Power Rule

Mathematics is a language of instructions.  
When you see a formula like the **Power Rule**, you’re not just seeing symbols — you’re seeing a *procedure* that tells you *what to do* and *how each step transforms the function*.

---

### Step 1: Read the Expression Aloud

::: promptbox
**Prompt:**  
Read this equation aloud and explain what each part is asking you to do:

\[
\frac{d}{dx}\big(x^n\big) = n x^{n-1}
\]

| Symbol / Term | What It Tells You To Do |
|:---------------|:------------------------|
| \( \frac{d}{dx} \) | |
| \( x^n \) | |
| \( n \) | |
| \( x^{n-1} \) | |

**Notes:**  
<br><br><br><br><br><br>
:::

---

### Step 2: Translate to a Step-by-Step Recipe

::: promptbox
**Activity (Think–Pair–Share):**  
In your own words, write the **recipe** the Power Rule gives you for differentiating \( x^n \):

1. Look at the exponent \( n \).  
2. Multiply the whole expression by that exponent.  
3. Subtract one from the exponent.  
4. Keep the base \( x \) the same.  

Now apply this recipe to three quick examples:

| Function \( f(x) \) | Your Steps | Derivative \( f'(x) \) |
|:--------------------:|:-----------|:-----------------------:|
| \( x^2 \) | | |
| \( x^5 \) | | |
| \( x^{-3} \) | | |

**Notes:**  
<br><br><br><br><br><br><br>
:::

---

### Step 3: Deriving the Power Rule from the Limit Definition

Let’s see how the Power Rule comes directly from the **limit definition of a derivative**.

We start with:

\[
f'(x) = \lim_{h \to 0} \frac{(x+h)^n - x^n}{h}
\]

We expand \((x+h)^n\) using the **Binomial Theorem**:

\[
(x + h)^n = x^n + n x^{n-1}h + \frac{n(n-1)}{2}x^{n-2}h^2 + \cdots + h^n
\]

Substitute this expansion into the limit expression:

\[
f'(x) = \lim_{h \to 0} 
\frac{\big(x^n + n x^{n-1}h + \frac{n(n-1)}{2}x^{n-2}h^2 + \cdots + h^n\big) - x^n}{h}
\]

Simplify the numerator — the \(x^n\) terms cancel:

\[
f'(x) = \lim_{h \to 0} 
\frac{n x^{n-1}h + \frac{n(n-1)}{2}x^{n-2}h^2 + \cdots + h^n}{h}
\]

Now factor out \(h\):

\[
f'(x) = \lim_{h \to 0} 
\Big( n x^{n-1} + \frac{n(n-1)}{2}x^{n-2}h + \cdots + h^{n-1} \Big)
\]

As \(h \to 0\), all terms containing \(h\) vanish, leaving only:

\[
\boxed{f'(x) = n x^{n-1}}
\]

That’s the **Power Rule** — it’s the simplified result of the limit definition.

---

### Step 4: Apply the Instructions to a Real Example

::: promptbox
**Practice:**  
Let \( f(x) = x^4 \).

1. Identify the base and exponent.  
2. Apply the Power Rule as a set of instructions.  
3. Simplify your result.  
4. Interpret your answer: what does \( f'(x) \) *mean* if \(f(x)\) represents biomass growth rate over time?

**Notes:**  
<br><br><br><br><br><br><br><br><br><br>
:::

---

### Step 5: Reflect — Reading vs. Memorizing

::: promptbox
**Reflection Prompt:**  
Complete this thought:

> “When I see \( \frac{d}{dx}(x^n) = n x^{n-1} \), I now think of it as a set of instructions to ________.”

**Notes:**  
<br><br><br><br><br><br>
:::

## Activity: Reading Math as Instructions — The Constant Multiplier Rule

The **Constant Multiplier Rule** tells us how constants behave when we take derivatives.  
It’s the mathematical version of saying: “Scaling a process doesn’t change *how fast* it changes — only *by how much*.”

---

### Step 1: Read the Expression Aloud

::: promptbox
**Prompt:**  
Read this equation aloud and explain what each part is telling you to do:

\[
\frac{d}{dx}\big[c \cdot f(x)\big] = c \cdot f'(x) = c \cdot \frac{d}{dx}\big[f(x)\big]
\]

| Symbol / Term | What It Tells You To Do |
|:---------------|:------------------------|
| \( c \) | |
| \( f(x) \) | |
| \( f'(x) \) | |
| \( c \cdot f'(x) \) | |

**Notes:**  
<br><br><br><br><br><br>
:::

---

### Step 2: Translate to a Step-by-Step Recipe

::: promptbox
**Activity (Think–Pair–Share):**  
Describe the **procedure** this rule gives you in plain English.

1. Identify the constant \( c \).  
2. Differentiate the variable part \( f(x) \) as usual.  
3. Keep the constant \( c \) in front — it “comes along for the ride.”  
4. Multiply \( c \) by the derivative \( f'(x) \).

Now apply this recipe to three examples:

| Function \( f(x) \) | Constant \( c \) | Derivative \( f'(x) \) |
|:--------------------:|:----------------:|:-----------------------:|
| \( 5x^3 \) |  | |
| \( -2x^4 \) |  | |
| \( 7 \) |  | |

**Notes:**  
<br><br><br><br><br><br><br>
:::

---

### Step 3: Deriving the Rule from the Limit Definition

We can prove the rule using the **limit definition of a derivative**:

\[
f'(x) = \lim_{h \to 0} \frac{f(x+h) - f(x)}{h}
\]

For a constant multiple \( g(x) = c \cdot f(x) \):

\[
g'(x) = \lim_{h \to 0} \frac{c \cdot f(x+h) - c \cdot f(x)}{h}
\]

Factor out the constant \( c \):

\[
g'(x) = c \cdot \lim_{h \to 0} \frac{f(x+h) - f(x)}{h}
\]

Recognize that the remaining limit is just \( f'(x) \):

\[
\boxed{g'(x) = c \cdot f'(x)}
\]

That’s the **Constant Multiplier Rule** — the constant stays the same, while the variable part changes according to its own derivative.

---

### Step 4: Apply the Instructions to a Real Example

::: promptbox
**Practice:**  
Let \( f(x) = x^2 \), and define \( g(x) = 3f(x)\).

1. Identify the constant multiplier \( c \).  
2. Use the Power Rule to find \( f'(x) \).  
3. Apply the Constant Multiplier Rule.  
4. Simplify the result.



**Interpretation:**  
If \( f(x) \) describes tree height over time, multiplying by 3 simply scales the entire process — the *rate of growth* is three times faster, but the shape of the change is identical.

**Notes:**  
<br><br><br><br><br><br><br>
:::

---

### Step 5: Reflect — Scaling and Change

::: promptbox
**Reflection Prompt:**  
Complete this thought:

> “When I see \( \frac{d}{dx}\big[c \cdot f(x)\big] = c \cdot f'(x) \), I think of it as saying ________.”

**Notes:**  
<br><br><br><br><br><br>
:::


## Activity: Reading Math as Instructions — The Sum and Difference Rules

When we add or subtract functions, each one still changes according to its own rule.  
The **Sum and Difference Rules** tell us that differentiation distributes over addition and subtraction —  
you can take the derivative of each part *separately* and then combine the results.

---

### Step 1: Read the Expression Aloud

::: promptbox
**Prompt:**  
Read the following equations aloud and describe what they’re instructing you to do:

\[
\frac{d}{dx}\big[f(x) + g(x)\big] = f'(x) + g'(x)
\]

\[
\frac{d}{dx}\big[f(x) - g(x)\big] = f'(x) - g'(x)
\]

| Symbol / Term | What It Tells You To Do |
|:---------------|:------------------------|
| \( f(x) + g(x) \) | |
| \( f'(x) + g'(x) \) | |
| \( f(x) - g(x) \) | |
| \( f'(x) - g'(x) \) | |

**Notes:**  
<br><br><br><br><br><br>
:::

---

### Step 2: Translate to a Step-by-Step Recipe

::: promptbox
**Activity (Think–Pair–Share):**  
Write the *recipe* the Sum and Difference Rules describe in your own words.

1. Break the function into parts being added or subtracted.  
2. Take the derivative of each part *individually*.  
3. Keep the same operation (addition or subtraction) between them.  
4. Simplify the result.

Now apply this recipe to three quick examples:

| Function \( f(x) \) | Steps You’d Take | Derivative \( f'(x) \) |
|:--------------------:|:----------------|:-----------------------:|
| \( x^2 + 3x \) | | |
| \( 4x^3 - 5x^2 \) | | |
| \( 2x^4 - 7x + 9 \) | | |

**Notes:**  
<br><br><br><br><br><br><br>
:::

---

### Step 3: Deriving the Rule from the Limit Definition

Let’s see how this rule arises from the **limit definition of the derivative**.

We start with:
\[
h'(x) = \lim_{h \to 0} \frac{h(x+h) - h(x)}{h}
\]
and let \( h(x) = f(x) + g(x) \).

Then:
\[
h'(x) = \lim_{h \to 0} \frac{[f(x+h) + g(x+h)] - [f(x) + g(x)]}{h}
\]

Simplify:
\[
h'(x) = \lim_{h \to 0} \frac{f(x+h) - f(x)}{h} + \lim_{h \to 0} \frac{g(x+h) - g(x)}{h}
\]

Recognize each term as a derivative:
\[
\boxed{h'(x) = f'(x) + g'(x)}
\]

The same reasoning works for subtraction:
\[
\boxed{\frac{d}{dx}[f(x) - g(x)] = f'(x) - g'(x)}
\]

In short — **differentiation distributes over addition and subtraction.**

---

### Step 4: Apply the Instructions to a Real Example

::: promptbox
**Practice:**  
Let \( f(x) = 3x^3 + 4x^2 - 10x \).

1. Identify how many terms there are.  
2. Apply the derivative to each term using the Power Rule.  
3. Combine the results.  
4. Simplify your answer.



**Notes:**  
<br><br><br><br><br><br><br>
:::

---

### Step 5: Visualize What’s Happening

::: promptbox
**Visualization Prompt:**  
Examine the graphs of \( f(x) = x^2 \) and \( g(x) = x \).  
Their sum \( f(x) + g(x) \) combines both shapes —  
the derivative \( f'(x) + g'(x) \) describes how *both rates of change* add together at each point.

```{r, echo=FALSE,fig.width=7, fig.height=10}
# Clean, proportional plots for f(x) = x^2 + x and f'(x) = 2x + 1
# x-axis from -5 to 5; square-looking panels; clear component comparison

# ----- Functions -----
f1 <- function(x) x^2
g  <- function(x) x
f  <- function(x) f1(x) + g(x)

# Derivatives
f1p <- function(x) 2 * x
gp  <- function(x) rep(1, length(x))
fp  <- function(x) f1p(x) + gp(x)

# ----- Domain -----
x <- seq(-5, 5, length.out = 800)
y_f1  <- f1(x)
y_g   <- g(x)
y_f   <- f(x)
y_f1p <- f1p(x)
y_gp  <- gp(x)
y_fp  <- fp(x)

# ----- Plotting -----
op <- par(mfrow = c(2,1), mar = c(4,4,2,1), cex = 1.2)

# (1) Function view
plot(x, y_f, type = "l", lwd = 3, col = "steelblue",
     xlab = "x", ylab = "y",
     xlim = c(-5, 5), ylim = c(-10, 30),
     main = "Functions: f(x) = x^2 + x and Its Components")
grid()
lines(x, y_f1, lwd = 2, col = "darkorange")
lines(x, y_g,  lwd = 2, col = "seagreen")
abline(h = 0, v = 0, col = "gray70", lty = 2)

legend("topleft",
       legend = c("f(x) = x^2 + x", "x^2", "x"),
       col = c("steelblue", "darkorange", "seagreen"),
       lwd = c(3,2,2), bty = "n")

# (2) Derivative view
plot(x, y_fp, type = "l", lwd = 3, col = "steelblue",
     xlab = "x", ylab = "f'(x)",
     xlim = c(-5, 5), ylim = c(-10, 10),
     main = "Derivatives: f'(x) = 2x + 1 and Its Components")
grid()
lines(x, y_f1p, lwd = 2, col = "darkorange")
lines(x, y_gp,  lwd = 2, col = "seagreen")
abline(h = 0, v = 0, col = "gray70", lty = 2)

legend("topleft",
       legend = c("f'(x) = 2x + 1", "(x^2)' = 2x", "(x)' = 1"),
       col = c("steelblue", "darkorange", "seagreen"),
       lwd = c(3,2,2), bty = "n")

par(op)


```

What do you notice about how the slopes of the two functions interact?

**Notes:**  
<br><br><br><br><br><br><br>
:::

---

### Step 6: Reflect — Combining Changes

::: promptbox
**Reflection Prompt:**  
Complete this thought:

> “When I see \( \frac{d}{dx}[f(x) \pm g(x)] = f'(x) \pm g'(x) \),  
> I think of it as telling me that ________.”

**Notes:**  
<br><br><br><br><br><br>
:::





## Practice: Identifying and Applying Derivative Rules

Each of the following problems involves one or more of the **basic derivative rules** you’ve learned so far:

- **Constant Rule**: \( \frac{d}{dx}(c) = 0 \)  
- **Constant Multiplier Rule**: \( \frac{d}{dx}[c \cdot f(x)] = c \cdot f'(x) \)  
- **Power Rule**: \( \frac{d}{dx}(x^n) = n x^{n-1} \)  
- **Sum/Difference Rule**: \( \frac{d}{dx}[f(x) \pm g(x)] = f'(x) \pm g'(x) \)

For each function below:

1. Identify **which rule(s)** apply.  
2. Write each step of differentiation clearly.  
3. Simplify your final answer.  
4. When possible, interpret what the derivative tells you about how the quantity is changing.

---

### Problem 1
\[
f(x) = 7
\]

**Tasks:**  
- Identify which rule applies.  
- Explain why the derivative is zero.

**Notes:**  
<br><br><br><br><br><br>

---

### Problem 2
\[
f(x) = 5x^3
\]

**Tasks:**  
- Identify which rule(s) apply.  
- Compute \( f'(x) \).  
- Interpret what happens to the rate of change when you multiply the function by 5.

**Notes:**  
<br><br><br><br><br><br>

---

### Problem 3
\[
f(x) = -2x^4 + 9x
\]

**Tasks:**  
- Identify all rules used.  
- Compute \( f'(x) \).  
- Check whether the derivative is positive, negative, or zero when \( x = 1 \).

**Notes:**  
<br><br><br><br><br><br>

---

### Problem 4
\[
f(x) = 6x^2 - 8
\]

**Tasks:**  
- Identify the constant, power, and constant multiplier rules as they appear.  
- Compute \( f'(x) \).  
- Which term disappears? Why?

**Notes:**  
<br><br><br><br><br><br>

---

### Problem 5
\[
f(x) = 3x^5 + 2x^2 - 10
\]

**Tasks:**  
- Identify each rule used.  
- Write out the derivative term by term.  
- Simplify the final expression.

**Notes:**  
<br><br><br><br><br><br>

---

### Problem 6
\[
f(x) = -7x^{-3}
\]

**Tasks:**  
- Identify which rule(s) apply.  
- Compute \( f'(x) \) using the Power Rule for negative exponents.  
- How does the sign of the derivative compare to the original function?

**Notes:**  
<br><br><br><br><br><br>

---

### Problem 7
\[
f(x) = \frac{1}{4}x^4 - 9x^2 + 12
\]

**Tasks:**  
- Identify the rules used for each term.  
- Compute the derivative step-by-step.  
- Simplify and factor the result if possible.

**Notes:**  
<br><br><br><br><br><br>

---

### Problem 8
\[
f(x) = 2(5x^3 - 4x)
\]

**Tasks:**  
- Identify which rule applies first (constant multiplier or power?).  
- Apply the constant multiplier rule correctly.  
- Expand and simplify \( f'(x) \).

**Notes:**  
<br><br><br><br><br><br>

---

### Problem 9
\[
f(x) = 4x^2 + 3x^{-1} - 7
\]

**Tasks:**  
- Identify the power and constant rules used.  
- Write each differentiation step.  
- Discuss what happens to the derivative as \( x \to 0 \).

**Notes:**  
<br><br><br><br><br><br>

---

### Problem 10
\[
f(x) = -6x^5 + 3x^3 - 2x + 8
\]

**Tasks:**  
- Identify all rules you’re applying.  
- Compute \( f'(x) \).  
- Evaluate \( f'(2) \) and interpret it as a rate of change (in context: imagine \( f(x) \) represents total carbon stored in a forest).

**Notes:**  
<br><br><br><br><br><br>

---

### Challenge 1
\[
f(x) = \sqrt{x}
\]


**Notes:**  
<br><br><br><br><br><br>

---

### Challenge 2
\[
f(x) = \frac{1}{x}
\]

**Notes:**  
<br><br><br><br><br><br>

---

### Challenge 3
\[
f(x) = x^{3/2}
\]


**Notes:**  
<br><br><br><br><br><br>

---

## Workbook: Trigonometric Derivatives

**Learning goals**

- Recognize and apply the derivative rules for sine, cosine, and tangent.
- Visually connect a function’s graph to the shape of its derivative.
- Practice taking derivatives that require the chain rule with trig functions.

---

### 1) Visual Warm-Up: Sketch the derivatives\( \sin x \), \( \cos x \), \( \tan x \)

::: promptbox

**Prompt:**  

- For each top plot, sketch the derivative on the corresponding blank axes below.  
- Where is the slope positive/negative/zero? Mark a few key points (e.g., where slope is zero or maximally positive/negative).

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=8, fig.height=8}
library(tidyverse)
library(patchwork)

# Domain
x <- seq(-2*pi, 2*pi, length.out = 1500)

# Data for sin and cos
df_sin <- tibble(x = x, y = sin(x))
df_cos <- tibble(x = x, y = cos(x))

# Data for tan (clip extreme values to avoid vertical asymptote spikes)
df_tan <- tibble(x = x, y = tan(x)) |>
  mutate(y = ifelse(abs(y) > 3, NA_real_, y))  # hide |tan| > 3 for readability

# Common theming
base_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# Function plots
p_sin <- ggplot(df_sin, aes(x, y)) +
  geom_line() +
  labs(title = "f(x) = sin(x)", y = NULL, x = NULL) +
  base_theme

p_cos <- ggplot(df_cos, aes(x, y)) +
  geom_line() +
  labs(title = "f(x) = cos(x)", y = NULL, x = NULL) +
  base_theme

p_tan <- ggplot(df_tan, aes(x, y)) +
  geom_line(na.rm = TRUE) +
  coord_cartesian(ylim = c(-3, 3)) +
  labs(title = "f(x) = tan(x)", y = NULL, x = NULL) +
  base_theme

# Blank "derivative sketch" axes (matching x & y ranges)
blank_sin <- ggplot() +
  coord_cartesian(xlim = c(-2*pi, 2*pi), ylim = c(-1.5, 1.5)) +
  labs(title = "Sketch f′(x) here (for sin)", x = "x", y = NULL) +
  base_theme

blank_cos <- ggplot() +
  coord_cartesian(xlim = c(-2*pi, 2*pi), ylim = c(-1.5, 1.5)) +
  labs(title = "Sketch f′(x) here (for cos)", x = "x", y = NULL) +
  base_theme

blank_tan <- ggplot() +
  coord_cartesian(xlim = c(-2*pi, 2*pi), ylim = c(-3, 3)) +
  labs(title = "Sketch f′(x) here (for tan)", x = "x", y = NULL) +
  base_theme

# Arrange as a 3x2 grid: function on top, blank axis below
(p_sin / blank_sin) | (p_cos / blank_cos) | (p_tan / blank_tan)

```

:::

---

### 2) Trig Derivative Rules (for reference)

| Function | Derivative |
|:--------:|:----------:|
| \(\sin x\) | \(\cos x\) |
| \(\cos x\) | \(-\sin x\) |
| \(\tan x\) | \(\sec^2 x\) |
| \(\sec x\) | \(\sec x \tan x\) |
| \(\csc x\) | \(-\csc x \cot x\) |
| \(\cot x\) | \(-\csc^2 x\) |

And a reminder of some of the trig functions

| Function | Definition (in terms of sine and cosine) |
|:----------|:----------------------------------------|
| \( \tan(x) \) | \( \dfrac{\sin(x)}{\cos(x)} \) |
| \( \sec(x) \) | \( \dfrac{1}{\cos(x)} \) |
| \( \cot(x) \) | \( \dfrac{\cos(x)}{\sin(x)} \) |
| \( \csc(x) \) | \( \dfrac{1}{\sin(x)} \) |


*Notes:*  

- These are **base** identities. The **ONLY** when the trigonometric operator is **x**. In a week we'll learn how to take the derivative of something that takes the form of \(sin(f(x))\)

---

### Practice Problems: Trigonometric Derivatives

Use the derivative rules below to find each derivative.

::: promptbox
Find the derivative of  
\[
f(x) = 10\sin(x)
\]

**Notes:**  
<br><br><br><br><br><br>


<details>
<summary><strong>Solution</strong></summary>

\[
f'(x) = 10\cos(x)
\]
</details>
:::
---

::: promptbox
Find the derivative of  
\[
f(x) = -7\cos(x)
\]


**Notes:**  
<br><br><br><br><br><br>

<details>
<summary><strong>Solution</strong></summary>

\[
f'(x) = -7[-\sin(x)] = 7\sin(x)
\]
</details>

:::

---

::: promptbox
Find the derivative of  
\[
f(x) = 4\tan(x) - 2\csc(x)
\]


**Notes:**  
<br><br><br><br><br><br>

<details>
<summary><strong>Solution</strong></summary>

\[
f'(x) = 4\sec^2(x) - 2[-\csc(x)\cot(x)] = 4\sec^2(x) + 2\csc(x)\cot(x)
\]
</details>

:::


## Workbook: Exponential and Logarithmic Derivatives

**Learning goals**

- Recognize and apply the derivative rules for exponential and logarithmic functions.
- Visually connect the function and its derivative through growth and decay patterns.
- Practice applying constant multiple and sum rules with exponential and logarithmic functions.

---

### 1) Visual Warm-Up: Sketch the derivatives of \( e^x \), \( a^x \), and \( \ln(x) \)

::: promptbox

**Prompt:**  

- For each top plot, sketch the derivative on the corresponding blank axes below.  
- How does the derivative compare to the original function?  
- Where is the rate of change the greatest?

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=8, fig.height=8}
library(tidyverse)
library(patchwork)

# Domain
x <- seq(-2, 4, length.out = 1000)

# Example base functions
df_exp <- tibble(x = x, y = exp(x))
df_2x <- tibble(x = x, y = 2^x)
df_ln <- tibble(x = seq(0.05, 4, length.out = 1000), y = log(x))

# Common theme
base_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# Function plots
p_exp <- ggplot(df_exp, aes(x, y)) +
  geom_line() +
  labs(title = "f(x) = e^x", y = NULL, x = NULL) +
  base_theme

p_2x <- ggplot(df_2x, aes(x, y)) +
  geom_line() +
  labs(title = "f(x) = 2^x", y = NULL, x = NULL) +
  base_theme

p_ln <- ggplot(df_ln, aes(x, y)) +
  geom_line() +
  labs(title = "f(x) = ln(x)", y = NULL, x = NULL) +
  base_theme

# Blank axes
blank_exp <- ggplot() +
  coord_cartesian(xlim = c(-2, 4), ylim = c(0, 55)) +
  labs(title = "Sketch f′(x) here (for e^x)", x = "x", y = NULL) +
  base_theme

blank_2x <- ggplot() +
  coord_cartesian(xlim = c(-2, 4), ylim = c(0, 55)) +
  labs(title = "Sketch f′(x) here (for 2^x)", x = "x", y = NULL) +
  base_theme

blank_ln <- ggplot() +
  coord_cartesian(xlim = c(0, 4), ylim = c(-2, 2)) +
  labs(title = "Sketch f′(x) here (for ln(x))", x = "x", y = NULL) +
  base_theme

# Combine plots
(p_exp / blank_exp) | (p_2x / blank_2x) | (p_ln / blank_ln)

```


**Exponential & Logarithmic Derivatives**

| Function | Derivative |
|:--------:|:----------:|
| \(a^x\) | \(a^x \ln(a)\) |
| \(e^x\) | \(e^x\) |
| \(\log_a(x)\) | \(\dfrac{1}{x \ln(a)}\) |
| \(\ln(x)\) | \(\dfrac{1}{x}\) |


*Notes:*  

- These are **base identities**, valid only when the exponent or argument is **x** (not another function).  

### Practice Problems: Exponential and Logarithmic Derivatives

Use the derivative rules above to find each derivative.

---

::: promptbox
Find the derivative of  
\[
f(x) = 6e^x
\]

**Notes:**  
<br><br><br><br><br><br>

<details>
<summary><strong>Solution</strong></summary>

\[
f'(x) = 6e^x
\]
</details>
:::

---

::: promptbox
Find the derivative of  
\[
f(x) = 5 \cdot 3^x
\]

**Notes:**  
<br><br><br><br><br><br>

<details>
<summary><strong>Solution</strong></summary>

\[
f'(x) = 5 \cdot 3^x \ln(3)
\]
</details>
:::

---

::: promptbox
Find the derivative of  
\[
f(x) = 4\ln(x) - 2\log_{10}(x)
\]

**Notes:**  
<br><br><br><br><br><br>

<details>
<summary><strong>Solution</strong></summary>

\[
f'(x) = 4\left(\frac{1}{x}\right) - 2\left(\frac{1}{x \ln(10)}\right)
\]
</details>
:::

## Reading Math Self Check

**Learning goal:**  
Strengthen your ability to *read mathematical notation* and apply derivative rules that involve inner functions \( g(x) \).  

---

### Exponential and Logarithmic Rules (General Form)

| Function | Derivative |
|:--------:|:----------:|
| \(a^{g(x)}\) | \(g'(x)\, a^{g(x)} \ln(a)\) |
| \(e^{g(x)}\) | \(g'(x)\, e^{g(x)}\) |
| \(\log_a(g(x))\) | \(\dfrac{g'(x)}{g(x)\ln(a)}\) |
| \(\ln(g(x))\) | \(\dfrac{g'(x)}{g(x)}\) |

---

### How to Use This Section

This section is meant to help you check your understanding — **without step-by-step guidance**.  

Read each rule carefully, then apply it to the corresponding problem.  


---

::: promptbox
#### Problem 1 — Using \(a^{g(x)}\)

**Rule reminder:**  
\[
\frac{d}{dx}\big[a^{g(x)}\big] = g'(x)\,a^{g(x)}\ln(a)
\]

Find the derivative of  
\[
f(x) = 3^{2x}
\]

**Notes:**  
<br><br><br><br><br><br>

<details>
<summary><strong>Solution</strong></summary>

Here \(a = 3\) and \(g(x) = 2x\), so \(g'(x) = 2\).  
\[
f'(x) = 2 \cdot 3^{2x} \ln(3)
\]
</details>
:::

---

::: promptbox
#### Problem 2 — Using \(e^{g(x)}\)

**Rule reminder:**  
\[
\frac{d}{dx}\big[e^{g(x)}\big] = g'(x)\,e^{g(x)}
\]

Find the derivative of  
\[
f(x) = e^{5x - 1}
\]

**Notes:**  
<br><br><br><br><br><br>



What happens if you use this rule to find the derivative of 
\[
f(x) = e^{x}
\]
<details>
<summary><strong>Solution</strong></summary>

Here \(g(x) = 5x - 1\), so \(g'(x) = 5\).  
\[
f'(x) = 5e^{5x - 1}
\]
</details>
:::

---

::: promptbox
#### Problem 3 — Using \(\log_a(g(x))\)

**Rule reminder:**  
\[
\frac{d}{dx}\big[\log_a(g(x))\big] = \frac{g'(x)}{g(x)\ln(a)}
\]

Find the derivative of  
\[
f(x) = \log_2(x^3 + 1)
\]

**Notes:**  
<br><br><br><br><br><br>

<details>
<summary><strong>Solution</strong></summary>

Here \(a = 2\), \(g(x) = x^3 + 1\), and \(g'(x) = 3x^2\).  
\[
f'(x) = \frac{3x^2}{(x^3 + 1)\ln(2)}
\]
</details>
:::

---

::: promptbox
#### Problem 4 — Using \(\ln(g(x))\)

**Rule reminder:**  
\[
\frac{d}{dx}\big[\ln(g(x))\big] = \frac{g'(x)}{g(x)}
\]

Find the derivative of  
\[
f(x) = \ln(4x^2 + 5)
\]

**Notes:**  
<br><br><br><br><br><br>


What happens if you use this rule to find the derivative of 
\[
f(x) = \ln(x)
\]
<details>
<summary><strong>Solution</strong></summary>

Here \(g(x) = 4x^2 + 5\), so \(g'(x) = 8x\).  
\[
f'(x) = \frac{8x}{4x^2 + 5}
\]
</details>
:::

---

### Reflection

- Which parts of the rule notation did you initially overlook?  
- What helped you identify \(g(x)\) and \(g'(x)\) correctly?  
- How does this process differ from using the simpler base forms like \(e^x\) or \(\ln(x)\)?

*Key takeaway:*  
Being able to “read” and interpret symbolic rules like these is a core mathematical skill — it’s not just about memorization, but about recognizing structure and patterns across different functions.
