# Applications of Derivatives - Workbook

## Introduction — Reading Behavior Through Derivatives

In this workbook, we’ll use **derivatives** to uncover what’s happening within a function — not just its values, but how it *behaves*.  
By studying derivatives, we can see where a system is **growing**, **slowing**, or **reversing direction** — and what those shifts mean in an environmental context.

- We’ll use **first derivatives** to identify where the function is **increasing or decreasing** and to locate **extrema** — points where growth stops and reverses.  
- We’ll use **second derivatives** to understand how that rate of change itself is evolving — whether growth is **accelerating**, **slowing**, or approaching a **tipping point**.

Together, these tools allow us to interpret a system’s story: when it’s changing most rapidly, when it levels off, and when it begins to turn — revealing the dynamic patterns that shape natural processes.


## Guided Activity — Using Derivatives to Find Extremes

This activity will help you explore how derivatives identify **extrema** — the highest and lowest points of a system’s behavior.

---

### Thinking About Extremes

::: promptbox
Before jumping into equations, think about these situations:

| Scenario | What variable reaches a high or low point? | What does that point mean ecologically? |
|--------------------|-------------------------------------|----------------------------------|
| A glacier’s melt rate throughout the summer | | |
| Algal population growth as nutrient levels rise | | |
| Temperature of a lake over a 24-hour cycle | | |
| Rainfall-runoff relationship during a storm | | |

**Prompt:**  
In each case, what would the *slope* of the curve (the derivative) tell you?  
When might that slope be **zero**?

:::

---

### Visualizing Local and Global Extrema

In this section we will interpret these functions and points of interest.

```{r four-extrema-types-wb, echo=FALSE, fig.width=10, fig.height=8}
library(ggplot2)
library(patchwork)

# A. Local Max and Min
f1 <- function(x) x^3 - 3*x^2 + 2
x1 <- seq(-1, 3.5, length.out = 400)
df1 <- data.frame(x = x1, y = f1(x1))
p1 <- ggplot(df1, aes(x, y)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  labs(title = "A. Local Max and Min (Cubic Function)",
       subtitle = expression(f(x) == x^3 - 3*x^2 + 2),
       x = "x", y = "f(x)") +
  theme_minimal(base_size = 12)

# B. Global Extrema at Endpoints
f2 <- function(x) -(x - 3)^2 + 5
x2 <- seq(0, 6, length.out = 400)
df2 <- data.frame(x = x2, y = f2(x2))
p2 <- ggplot(df2, aes(x, y)) +
  geom_line(color = "darkgreen", linewidth = 1.2) +
  labs(title = "B. Global Extrema at Endpoints (Finite Domain)",
       subtitle = expression(f(x) == -~(x-3)^2 + 5~~" over [0,6]"),
       x = "x", y = "f(x)") +
  theme_minimal(base_size = 12)

# C. Flat Region (No Extremum)
f3 <- function(x) x^3
x3 <- seq(-2, 2, length.out = 400)
df3 <- data.frame(x = x3, y = f3(x3))
p3 <- ggplot(df3, aes(x, y)) +
  geom_line(color = "orange", linewidth = 1.2) +
  labs(title = "C. Flat Region — No Extrema",
       subtitle = expression(f(x) == x^3),
       x = "x", y = "f(x)") +
  theme_minimal(base_size = 12)

# D. Periodic Function
f4 <- function(x) sin(x)
x4 <- seq(0, 2*pi, length.out = 400)
df4 <- data.frame(x = x4, y = f4(x4))
p4 <- ggplot(df4, aes(x, y)) +
  geom_line(color = "purple", linewidth = 1.2) +
  labs(title = "D. Periodic Function with Repeating Extrema",
       subtitle = expression(f(x) == sin(x)~~" over [0,2π]"),
       x = "x", y = "f(x)") +
  theme_minimal(base_size = 12)

(p1 | p2) / (p3 | p4)

```

::: promptbox

#### Visual Observations {-}

| Panel | What kind of extrema do you see? | Where does the curve rise/fall? | Any flat regions? |
|--------|----------------------------------|---------------------------------|-------------------|
| A | | | |
| B | | | |
| C | | | |
| D | | | |


:::

---

### Local vs. Global Thinking

::: promptbox
Using what you’ve seen:

1. How is a **local extremum** different from a **global extremum**?  
2. Why can **endpoints** matter when the domain is limited (like daily radiation between sunrise and sunset)?  

:::

---



### Connecting to Derivatives


::: promptbox

#### We Are “Difference Machines” {-}

Our brains are wired to notice **change**, not constancy.

If a light stays on, we stop seeing it.  
If a sound is steady, it fades into the background.  
But when something shifts — brightness flickers, volume drops, temperature rises — our attention snaps to it instantly.

Your mind is not recording the world frame by frame — it’s continuously comparing **what just happened** to **what’s happening now**.

We don’t process absolute information; we process **differences**.

---

#### Why We Measure Change {-}

Environmental scientists do the same thing.

We rarely ask, “What is the exact value right now?”  
Instead, we ask:
- *How fast is the glacier melting?*  
- *Is the population growing or shrinking?*  
- *How quickly is CO₂ increasing?*  
- *When does the trend reverse?*

These are all **questions about change**, not state.

Just as our brains focus on shifting inputs, our mathematical tools — **derivatives** — are designed to measure **how one quantity changes relative to another**.

---

#### The Calculus Connection{-}

- The **first derivative** \( f'(x) \) measures *how fast* something changes — our mathematical version of noticing movement, drift, or growth.  
- The **second derivative** \( f''(x) \) measures *how the rate of change itself evolves* — like sensing acceleration or sudden reversals.

Both capture what your brain already prioritizes: **difference** and **change over time**.

:::

---

At each peak or valley, the slope of the tangent line is **zero**.

| Observation | Symbolic Meaning | Real-World Analogy |
|:-------------:|:----------------:|:------------------:|
| Curve is increasing | \( f'(x) > 0 \) | Lake temperature rising during morning |
| Curve is decreasing | \( f'(x) < 0 \) | River discharge falling after storm peak |
| Curve flattens | \( f'(x) = 0 \) | Population growth stops accelerating |

::: promptbox

#### Finding Critical Points {-}

How can we use \( f'(x) = 0 \) to predict where peaks and valleys occur before plotting the function?

:::

---

### Classify Behavior from the Derivative

For each derivative, decide where the function is **increasing**, **decreasing**, or **flat** and over which intervals?

Note we don't have the function here - just its derivative.

| Derivative Expression | Critical points of \( f'(x) \) | Behavior of \( f(x) \) |
|------------------------|--------------------|-------------------------|
| \( f'(x) = 3x^2 - 6x \) | | |
| \( f'(x) = -2(x-3) \) | | |
| \( f'(x) = 4x^3 - 10x \) | | |

There will be times where factoring isn't simple. **IF** the function is quadratic you can always use the **quadratic formula** which gives the solutions (or roots) of a quadratic equation  
\[
ax^2 + bx + c = 0
\]
and is written as:

\[
x = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}
\]

::: promptbox

#### How to build a sign chart {-}
 \( f'(x) = 3x^2 - 6x \)
<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

\( f'(x) = -2(x-3) \)
<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

\( f'(x) = 4x^3 - 10x \)
<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

:::


---



### Apply to a New Example

Let’s model population dynamics of bears:
\[
P(t) = -t^3 + 6t^2 - 9t + 20
\]

```{r, echo=FALSE}
P <- function(t) -t^3 + 6*t^2 - 9*t + 20
curve(P(x), from = 0, to = 5,
col = "forestgreen", lwd = 2,
xlab = "t", ylab = "P(t)",
main = "Population Model P(t)")
```



::: promptbox
#### Prompts {-}

- Between which time intervals does the population increase or decrease?  
- At what points might \( P'(t) = 0 \)?  
- Which point seems to be a **local maximum** and which a **local minimum**?  
- How could this correspond to resource limitation or recovery in an ecosystem?  
- What might happen if external factors (like drought or nutrient input) shifted these extrema?

:::

---

### Investigating the Derivative 

Now that you’ve visualized the population curve, let’s explore how its **rate of change** behaves.  

We’ll look at the derivative \( P'(t) \), which tells us whether the population is growing or shrinking at any moment.

```{r, echo=FALSE}
P_prime <- function(t) -3*t^2 + 12*t - 9
curve(P_prime(x), from = 0, to = 5, col = "steelblue", lwd = 2,
      ylab = "P'(t)", xlab = "t",
      main = "Derivative of the Population Model")
abline(h = 0, lty = 2, col = "gray50")
```

::: promptbox
#### Prompts {-}

- Where does \( P'(t) = 0 \)?  
- Between which values of \(t\) is \(P'(t)\) positive or negative?  
- How does the sign of \( P'(t) \) relate to whether \( P(t) \) is increasing or decreasing?  
- How do these intervals correspond to growth and decline phases in a population?  
- What might cause these transitions in a real system (e.g., changes in food, space, or temperature)?  

:::

---

### Linking Derivatives to System Behavior

Use your graph to fill in the table below.

| Observation | Derivative Sign | Population Behavior | Ecological Meaning |
|--------------|----------------|----------------------|--------------------|
| \( P'(t) > 0 \) | Positive | \( P(t) \) increasing | Growth phase — population expanding |
| \( P'(t) = 0 \) | Zero | Slope of \( P(t) \) flat | Turning point — possible max or min |
| \( P'(t) < 0 \) | Negative | \( P(t) \) decreasing | Decline — competition or limitation |

---

### Adding the Second Derivative

Next, explore the **curvature** of population change — whether the growth rate itself is speeding up or slowing down.  

We’ll look at the second derivative \( P''(t) \).

```{r, echo=FALSE}
P_doubleprime <- function(t) -6*t + 12
curve(P_doubleprime(x), from = 0, to = 5, col = "darkorange", lwd = 2,
      ylab = "P''(t)", xlab = "t",
      main = "Second Derivative — Curvature of Population Change")
abline(h = 0, lty = 2, col = "gray50")
```


::: promptbox

#### Prompts{-}

- Where does \( P''(t) = 0 \)?  
- What happens to \( P'(t) \) and \( P(t) \) near this point?  
- How does the curvature (concavity) change before and after \( P''(t) = 0 \)?  
- What does a **positive** \( P''(t) \) tell you about the rate of change?  
- What does a **negative** \( P''(t) \) indicate about slowing or reversing growth?  
- In ecological terms, how might this point represent a **tipping point** or **carrying capacity**?  

:::

---

Now consider the **second derivative**, which describes how the rate of change itself is changing.  
Use this table to connect the **sign of \( P''(t) \)** with system behavior.

| Observation | Second Derivative Sign | Curvature / Behavior of \( P(t) \) | Ecological Meaning |
|--------------|------------------------|------------------------------------|--------------------|
| \( P''(t) > 0 \) | Positive | Concave up — slope of \( P(t) \) increasing | Growth is **accelerating**; system recovering or resources increasing |
| \( P''(t) = 0 \) | Zero | Point of inflection — curvature changes sign | **Tipping point** where growth shifts from accelerating to slowing |
| \( P''(t) < 0 \) | Negative | Concave down — slope of \( P(t) \) decreasing | Growth is **slowing**; system approaching **carrying capacity** or limitation |


::: promptbox

#### Discussion {-}

1. When \( P'(t) = 0 \), what does this mean physically in the system?  
2. When \( P''(t) = 0 \), what does this indicate about how the rate of change itself is behaving?  
3. How could identifying **inflection points** help forecast environmental transitions before they occur?  
4. Think of a real example — such as CO₂ uptake by vegetation, or lake temperature over a year.  
   - Where might you expect **maxima**, **minima**, and **inflection points** to occur?  
   - What could each tell you about system resilience or change?  

> **Big Takeaway:**   
> - \( f'(x) \) shows **how fast** something changes.  
> - \( f''(x) \) shows **how that rate** itself changes.  
> - Together, they reveal whether a system is **accelerating, stabilizing, or reversing** — the mathematical fingerprints of environmental balance.

:::


\newpage

## Sign Chart Practice Worksheet  
**Goal:** Practice creating **first** and **second derivative sign charts** for both original functions and their derivatives — learning to infer how a system behaves from either starting point.

Each part contains **two examples**: 

- One where you start from the **original function \( f(x) \)**.  
- One where you start directly from the **derivative \( f'(x) \)**.

---

\newpage

### Part 1 — Polynomial Patterns

#### A. Start from \( f(x) \){-}
Given \( f(x) = x^3 - 3x^2 \):  
1. Compute \( f'(x) \) and \( f''(x) \).  
2. Construct **first** and **second derivative sign charts**.  
3. Identify intervals of increase/decrease and concave up/down.  
4. Sketch \( f(x) \) using your charts — label local extrema and inflection points.

\newpage

#### B. Start from \( f'(x) \){-}
Given \( f'(x) = 2x(3-x) \):  
1. Identify critical points where \( f'(x) = 0 \).  
2. Construct **first** and **second derivative sign charts**.  
3. Identify intervals of increase/decrease and concave up/down.  
4. Sketch \( f(x) \) using your charts — label local extrema and inflection points.

**Reflection:**  
How do the two examples connect?  
If you only know \( f'(x) \), what key features of \( f(x) \) can you still determine?

---

\newpage

### Part 2 — Factored Derivatives and Flat Points

#### A. Start from \( f(x) \) {-}
Given \( f(x) = -x^3 + 3x^2 + 4x \):  
1. Compute \( f'(x) \) and factor it.  
2. Build a **first derivative sign chart** and find where slope changes sign.  
3. Compute \( f''(x) \) to locate concavity changes.  
4. Identify local maxima, minima, and inflection points.

\newpage

#### B. Start from \( f'(x) \) {-}
Given \( f'(x) = -3(x - 1)(x + 2)^2 \):  
1. Create a **sign chart** for \( f'(x) \).  
2. Build a **first derivative sign chart** and find where slope changes sign.  
3. Compute \( f''(x) \) to locate concavity changes.  
4. Identify local maxima, minima, and inflection points.

**Think about it:**  
How does a **repeated factor** like \((x+2)^2\) affect whether \( f(x) \) actually rises or falls around that point?

---

\newpage

### Part 3 — Exponential Behavior

#### A. Start from \( f(x) \){-}
Given \( f(x) = e^x(x - 2) \):  
1. Compute \( f'(x) \) and \( f''(x) \).  
2. Use **sign charts** to find where the function is increasing, decreasing, concave up, or concave down.  
3. Sketch the overall shape of \( f(x) \) and describe its long-term trend.


---

\newpage

### Part 4 — Undefined Derivatives and Asymptotes

#### A. Start from \( f(x) \){-}
Given \( f(x) = \frac{x}{x - 1} \):  
1. Compute \( f'(x) \) and \( f''(x) \).  
2. Identify points where the derivative is **undefined** or changes sign.  
3. Construct **sign charts** for both derivatives.  
4. Sketch \( f(x) \), marking asymptotes and concavity regions.

\newpage

#### B. Start from \( f'(x) \){-}
Given \( f'(x) = \frac{e^x(x^2 - 4)}{x - 2} \):  
1. Find critical and undefined points.  
2. Build a **first derivative sign chart** showing slope behavior.  
3. Sketch a possible \( f(x) \) consistent with that derivative’s pattern.

**Questions to guide you:**  

- How does an **undefined derivative** signal a boundary or asymptote?  
- Can \( f(x) \) still have a local extremum near such a point?
    
---

\newpage

### Reflection — From Slopes to Shapes

1. How does starting from \( f(x) \) vs. \( f'(x) \) change your reasoning process?  
2. Which features can you reliably infer from \( f'(x) \) alone — and which require \( f''(x) \)?  
3. In environmental contexts, what might local maxima, minima, and inflection points represent ?  


---

